### 2016-12-19 pasp/IOPLaTeXGuidelines.tex
dont steal plox

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%    INSTITUTE OF PHYSICS PUBLISHING                                   %
%                                                                      %
%   `Preparing an article for publication in an Institute of Physics   %
%    Publishing journal using LaTeX'                                   %
%                                                                      %
%    LaTeX source code `ioplau2e.tex' used to generate `author         %
%    guidelines', the documentation explaining and demonstrating use   %
%    of the Institute of Physics Publishing LaTeX preprint files       %
%    `iopart.cls, iopart12.clo and iopart10.clo'.                      %
%                                                                      %
%    `ioplau2e.tex' itself uses LaTeX with `iopart.cls'                %
%                                                                      %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%
% First we have a character check
%
% ! exclamation mark    " double quote  
% # hash                ` opening quote (grave)
% & ampersand           ' closing quote (acute)
% $ dollar              % percent       
% ( open parenthesis    ) close paren.  
% - hyphen              = equals sign
% | vertical bar        ~ tilde         
% @ at sign             _ underscore
% { open curly brace    } close curly   
% [ open square         ] close square bracket
% + plus sign           ; semi-colon    
% * asterisk            : colon
% < open angle bracket  > close angle   
% , comma               . full stop
% ? question mark       / forward slash 
% \ backslash           ^ circumflex
%
% ABCDEFGHIJKLMNOPQRSTUVWXYZ 
% abcdefghijklmnopqrstuvwxyz 
% 1234567890
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\documentclass[12pt]{iopart}
\newcommand{\gguide}{{\it Preparing graphics for IOP Publishing journals}}
%Uncomment next line if AMS fonts required
%\usepackage{iopams}  
\begin{document}

\title[Author guidelines for IOP Publishing journals in  \LaTeXe]{How to prepare and submit an article for 
publication in an IOP Publishing journal using \LaTeXe}

\author{Content \& Services Team}

\address{IOP Publishing, Temple Circus, Temple Way, Bristol BS1 6HG, UK}
\ead{submissions@iop.org}
\vspace{10pt}
\begin{indented}
\item[]February 2014
\end{indented}

\begin{abstract}
This document describes the  preparation of an article using \LaTeXe\ and 
\verb"iopart.cls" (the IOP Publishing \LaTeXe\ preprint class file).
This class file is designed to help 
authors produce preprints in a form suitable for submission to any of the
journals listed in table~\ref{jlab1} on the next page.  You are not obliged to use this class file---we accept
submissions using all common \LaTeX\ class and style files.  The \verb"iopart.cls"
class file is supplied merely as a convenience for those authors who find it useful.
This document gives both general advice that applies whatever class file you use, and specific advice
that applies if you choose to use \verb"iopart.cls".

We also accept submissions in Word format.  Instructions for Word submissions are available via the `Author Guidelines' link at \verb"http://authors.iop.org".

If you have any queries about this document or any aspect of preparing your article for submission please contact us at the e-mail address given above.
\end{abstract}

% Uncomment for PACS numbers
%\pacs{00.00, 20.00, 42.10}
%
% Uncomment for keywords
%\vspace{2pc}
%\noindent{\it Keywords}: XXXXXX, YYYYYYYY, ZZZZZZZZZ
%
% Uncomment for Submitted to journal title message
%\submitto{\JPA}
%
% Uncomment if a separate title page is required
%\maketitle
% 
% For two-column output uncomment the next line and choose [10pt] rather than [12pt] in the \documentclass declaration
%\ioptwocol
%

\section{Introduction: file preparation and submission}

The \verb"iopart" \LaTeXe\ article class file is provided to help authors prepare articles for submission to IOP Publishing journals.
  This document gives advice on preparing your submission, and specific instructions on how to use \verb"iopart.cls" to follow this advice.  You
  do not have to use \verb"iopart.cls"; articles prepared using any other common class and style files can also be submitted.
    It is not necessary to mimic the appearance of a published article.

The advice
on \LaTeX\ file preparation in this document applies to
the journals listed in table~\ref{jlab1}.  If your journal is not listed please go to the journal website via \verb"http://iopscience.iop.org/journals" for specific
submission instructions.

\begin{table}
\caption{\label{jlab1}Journals to which this document applies, and macros for the abbreviated journal names in {\tt iopart.cls}. Macros for other journal titles are listed in appendix\,A.}
\footnotesize
\begin{tabular}{@{}llll}
\br
Short form of journal title&Macro name&Short form of journal title&Macro name\\
\mr
2D Mater.&\verb"\TDM"&Mater. Res. Express&\verb"\MRE"\\
Biofabrication&\verb"\BF"&Meas. Sci. Technol.$^c$&\verb"\MST"\\
Bioinspir. Biomim.&\verb"\BB"&Methods Appl. Fluoresc.&\verb"\MAF"\\
Biomed. Mater.&\verb"\BMM"&Modelling Simul. Mater. Sci. Eng.&\verb"\MSMSE"\\
Class. Quantum Grav.&\verb"\CQG"&Nucl. Fusion$^a$&\verb"\NF"\\
Comput. Sci. Disc.&\verb"\CSD"&New J. Phys.&\verb"\NJP"\\
Environ. Res. Lett.&\verb"\ERL"&Nonlinearity$^{a,b}$&\verb"\NL"\\
Eur. J. Phys.$^a$&\verb"\EJP"&Nanotechnology&\verb"\NT"\\
Inverse Problems$^{b,c}$&\verb"\IP"&Phys. Biol.$^c$&\verb"\PB"\\
J. Breath Res.&\verb"\JBR"&Phys. Educ.$^a$&\verb"\PED"\\
J. Geophys. Eng.$^a$&\verb"\JGE"&Physiol. Meas.$^{c,d,e}$&\verb"\PM"\\
J. Micromech. Microeng.&\verb"\JMM"&Phys. Med. Biol.$^{c,d,e}$&\verb"\PMB"\\
J. Neural Eng.$^c$&\verb"\JNE"&Plasma Phys. Control. Fusion&\verb"\PPCF"\\
J. Opt.&\verb"\JOPT"&Phys. Scr.&\verb"\PS"\\
J. Phys. A: Math. Theor.&\verb"\jpa"&Plasma Sources Sci. Technol.&\verb"\PSST"\\
J. Phys. B: At. Mol. Opt. Phys.$^a$&\verb"\jpb"&Rep. Prog. Phys.$^{e}$&\verb"\RPP"\\
J. Phys: Condens. Matter&\verb"\JPCM"&Semicond. Sci. Technol.&\verb"\SST"\\
J. Phys. D: Appl. Phys.$^a$&\verb"\JPD"&Smart Mater. Struct.&\verb"\SMS"\\
J. Phys. G: Nucl. Part. Phys.&\verb"\jpg"&Supercond. Sci. Technol.&\verb"\SUST"\\
J. Radiol. Prot.$^a$&\verb"\JRP"&Surf. Topogr.: Metrol. Prop.&\verb"\STMP"\\
Metrologia$^a$&\verb"\MET"&Transl. Mater. Res.&\verb"\TMR"\\
\br
\end{tabular}\\
$^{a}$UK spelling is required; $^{b}$MSC classification may be used as well as PACS; $^{c}$titles of articles are required in journal references; $^{d}$Harvard-style references must be used (see section \ref{except}); $^{e}$final page numbers of articles are required in journal references.

\end{table}
\normalsize

Any special submission requirements for the journals are indicated with footnotes in table~\ref{jlab1}.
Journals which require references in a particular format will need special care if you are using BibTeX, and you might need to use a \verb".bst" file
that gives slightly non-standard output in order to supply any extra information required.  It is not
necessary to give references in the exact style of references used in published articles, as long as all of
the required information is present.

Also note that there is an incompatibility
between \verb"amsmath.sty" and \verb"iopart.cls" which cannot be completely worked around.  If your article relies
on commands in \verb"amsmath.sty" that are not available in \verb"iopart.cls", you may wish to consider using a different
class file.

Whatever journal you are submitting to, please look at recent published articles (preferably
articles in your subject area) to familiarize yourself with the features of the journal.  We do not demand
that your \LaTeX\ file closely resembles a published article---a generic `preprint' appearance of the sort
commonly seen on \verb"arXiv.org" is fine---but your submission should be presented
in a way that makes it easy for the referees to form an opinion of whether it is suitable for the journal.
The generic advice in this document---on what to include in an abstract, how best to present complicated
mathematical expressions, and so on---applies whatever class file you are using.

\subsection{What you will need to supply}
Submissions to our journals are handled via the ScholarOne web-based submission system.  When you submit
a new article to us you need only submit a PDF of your article.  When you submit a revised version,
we ask you to submit the source files as well.  Upon acceptance for publication we will use the source files to produce a proof of your article in the journal style.

\subsubsection{Text.}When you send us the source files for a revised version of your submission,
you should send us the \LaTeX\ source code of your paper with all figures read in by 
the source code (see section \ref{figinc}).  Articles can be prepared using almost any version of \TeX\ or \LaTeX{},
not just \LaTeX\ with the class file \verb"iopart.cls".  You may split your \LaTeX\ file into several parts, but please show
which is the `master' \LaTeX\ file that reads in all of the other ones by naming it appropriately.  The `master'
\LaTeX\ file must read in all other \LaTeX\ and figure files from the current directory.  {\it Do not read in files from a different directory, e.g. \verb"\includegraphics{/figures/figure1.eps}" or
\verb"\include{../usr/home/smith/myfiles/macros.tex}"---we store submitted files
all together in a single directory with no subdirectories}.
\begin{itemize}
\item {\bf Using \LaTeX\ packages.} Most \LaTeXe\ packages can be used if they are 
available in common distributions of \LaTeXe; however, if it is essential to use 
a non-standard package then any extra files needed to process the article must 
also be supplied.  Try to avoid using any packages that manipulate or change the standard
\LaTeX\ fonts: published articles use fonts in the Times family, but we prefer that you 
use \LaTeX\ default Computer Modern fonts in your submission.  The use of \LaTeX\ 2.09, and of plain
\TeX\ and variants such as AMSTeX is acceptable, but a complete PDF of your submission should be supplied in these cases.
\end{itemize}
\subsubsection{Figures.} Figures should ideally be included in an article as encapsulated PostScript files
(see section \ref{figinc}) or created using standard \LaTeX\ drawing commands. 
 Please name all figure files using the http://gmail.com/guidelines in section \ref{fname}.
We accept submissions that use pdf\TeX\ to include
PDF or bitmap figures, but please ensure that you send us a PDF that uses PDF version 1.4 or lower
(to avoid problems in the ScholarOne system).
You can do this by putting \verb"\pdfminorversion=4" at the very start of your TeX file.

\label{fig1}All figures should be included within the body of the text 
at an appropriate point or grouped together with their captions at the end of the article. A standard graphics inclusion package such as \verb"graphicx" should be used for figure inclusion, and the package should be declared in the usual
way, for example with \verb"\usepackage{graphicx}", after the \verb"\documentclass" command.
Authors should avoid using special effects generated by including verbatim
PostScript code in the submitted \LaTeX\ file. Wherever possible, please try to use standard \LaTeX\ tools 
and packages.

\subsubsection{References.\label{bibby}}
You can produce your bibliography in the standard \LaTeX\ way using the \verb"\bibitem" command. Alternatively
you can use BibTeX: our preferred  \verb".bst" styles are:

\begin{itemize}
\item For the numerical (Vancouver) reference style we recommend that authors use 
 \verb"unsrt.bst"; this does not quite follow the style of published articles in our
 journals but this is not a problem.  Alternatively \verb"iopart-num.bst" created by Mark A Caprio
 produces a reference style that closely matches that in published articles.  The file is available from
\verb"http://ctan.org/tex-archive/biblio/bibtex/contrib/iopart-num/" .
\item For alphabetical (Harvard) style references we recommend that authors use the \verb"harvard.sty"
in conjunction with the \verb"jphysicsB.bst" BibTeX style file.  These, and accompanying documentation, can be downloaded
from \penalty-10000 \verb"http://www.ctan.org/tex-archive/macros/latex/contrib/harvard/".
Note that the \verb"jphysicsB.bst" bibliography style does not include article titles
in references to journal articles.http://gmail.com/
To include the titles of journal articles you can use the style \verb"dcu.bst" which is included
in the \verb"harvard.sty" package.  The output differs a little from the final journal reference
style, but all of the necessary information is present and the reference list will be formatted
into journal house style as part of the production process if your article is accepted for publication.
\end{itemize}

\noindent Please make sure that you include your \verb".bib" bibliographic database file(s) and any 
\verb".bst" style file(s) you have used.

\subsection{\label{copyright}Copyrighted material and ethical policy} If you wish to make use of previously published material for which you do not own the copyright then you must seek permission from the copyright holder, usually both the author and the publisher.  It is your responsibility to obtain copyright permissions and this should be done prior to submitting your article. If you have obtained permission, please provide full details of the permission granted---for example, copies of the text of any e-mails or a copy of any letters you may have received. Figure captions must include an acknowledgment of the original source of the material even when permission to reuse has been obtained.  Please read our ethical policy (available at \verb"http://authors.iop.org/ethicalpolicy") before writing your article.

\subsection{Naming your files}
\subsubsection{General.}
Please name all your files, both figures and text, as follows:
\begin{itemize}
\item Use only characters from the set a to z, A to Z, 0 to 9 and underscore (\_).
\item Do not use spaces or punctuation characters in file names.
\item Do not use any accented characters such as
\'a, \^e, \~n, \"o.
\item Include an extension to indicate the file type (e.g., \verb".tex", \verb".eps", \verb".txt", etc).
\item Use consistent upper and lower case in filenames and in your \LaTeX\ file.
If your \LaTeX\ file contains the line \verb"\includegraphics{fig1.eps}" the figure file must be called
\verb"fig1.eps" and not \verb"Fig1.eps" or \verb"fig1.EPS".  If you are on a Unix system, please ensure that
there are no pairs of figures whose names differ only in capitalization, such as \verb"fig_2a.eps" and \verb"fig_2A.eps",
as Windows systems will be unable to keep the two files in the same directory.
\end{itemize}
When you submit your article files, they are manipulated
and copied many times across multiple databases and file systems. Including non-standard
characters in your filenames will cause problems when processing your article.
\subsubsection{\label{fname}Naming your figure files.} In addition to the above points, please give each figure file a name which indicates the number of the figure it contains; for example, \verb"figure1.eps", \verb"figure2a.eps", etc. If the figure file contains a figure with multiple parts, for example figure 2(a) to 2(e), give it a name such as \verb"figure2a_2e.eps", and so forth.
\subsection{How to send your files}
Please send your submission via the ScholarOne submission system.  Go to the journal home
page, and use the 'Submit an article' link on the right-hand side.

\subsection{Obtaining the class file and documentation}
This documentation and the \verb"iopart.cls" class file and associated files are available by anonymous FTP
at \verb"ftp://ftp.iop.org/pub/journals/".  They are available in two archive
formats there, \verb"ioplatexguidelines.tar.gz" and \verb"ioplatexguidelines.zip".  The individual files
are present in the subdirectory \verb"latex2e".

\section{Preparing your article}

\subsection{Sample coding for the start of an article}
\label{startsample}
The code for the start of a title page of a typical paper in the \verb"iopart.cls" style might read:
\small\begin{verbatim}
\documentclass[12pt]{iopart}
\begin{document}
\title[The anomalous magnetic moment of the 
neutrino]{The anomalous magnetic moment of the 
neutrino and its relation to the solar neutrino problem}

\author{P J Smith$^1$, T M Collins$^2$, 
R J Jones$^3$\footnote{Present address:
Department of Physics, University of Bristol, Tyndalls Park Road, 
Bristol BS8 1TS, UK.} and Janet Williams$^3$}

\address{$^1$ Mathematics Faculty, Open University, 
Milton Keynes MK7~6AA, UK}
\address{$^2$ Department of Mathematics, 
Imperial College, Prince Consort Road, London SW7~2BZ, UK}
\address{$^3$ Department of Computer Science, 
University College London, Gower Street, London WC1E~6BT, UK}
\ead{williams@ucl.ac.uk}

\begin{abstract}
...
\end{abstract}
\pacs{1315, 9440T}
\keywords{magnetic moment, solar neutrinos, astrophysics}
\submitto{\jpg}
\maketitle
\end{verbatim}
\normalsize

At the start of the \LaTeX\ source code please include 
commented material to identify the journal, author, and (if you are sending a revised
version or a resubmission) the reference number that the journal
has given to the submission. The first non-commented line should be 
\verb"\documentclass[12pt]{iopart}"  to load the preprint class 
file.  The normal text will be in the Computer Modern 12pt font.
It is possible to specify 10pt font size by passing the option \verb"[10pt]" to the class file.
Although it is possible to choose a font other than Computer Modern by loading external packages, this is not recommended.

The article text begins after \verb"\begin{document}".
Authors of very long articles may find it convenient to separate 
their article into a series of \LaTeX\ files each containing one section, and each of which is called 
in turn by the primary file.  The files for each section should be read in from the current directory;
please name the primary file clearly so that we know to run \LaTeX\ on this file.

Authors may use any common \LaTeX\ \verb".sty" files.
Authors may also define their own macros and definitions either in the main article \LaTeX\ file
or in a separate \verb".tex" or \verb".sty" file that is read in by the
main file, provided they do not overwrite existing definitions.
It is helpful to the production staff if complicated author-defined macros are explained in a \LaTeX\ comment.
The article class \verb"iopart.cls" can be used with other package files such
as those loading the AMS extension fonts 
\verb"msam" and \verb"msbm", which provide the 
blackboard bold alphabet and various extra maths symbols as well as symbols useful in figure 
captions.  An extra style file \verb"iopams.sty" is provided to load these
packages and provide extra definitions for bold Greek letters.

\subsection{\label{dblcol}Double-column layout}
The \verb"iopart.cls" class file produces single-column output by default, but a two-column layout can be obtained by
using \verb"\documentclass[10pt]" at the start of the file and \verb"\ioptwocol" after the \verb"\maketitle" command.  Two-column output will begin
on a new page (unlike in published double-column articles, where the two-column material
starts on the same page as the abstract).

In general we prefer to receive submissions in single-column format even for journals
published in double-column style; however, the \verb"\ioptwocol" option may be useful to test figure sizes
and equation breaks for these journals.  When setting material
in two columns you can use the asterisked versions of \LaTeX\ commands such as \verb"\begin{figure*} ... \end{figure*}"
to set figures and tables across two columns.  If you have any problems or any queries about producing two-column output, please contact us at \verb"submissions@iop.org".

\section{The title and abstract page} 
If you use \verb"iopart.cls", the code for setting the title page information is slightly different from
the normal default in \LaTeX.  If you are using a different class file, you do not need to mimic the appearance of
an \verb"iopart.cls" title page, but please ensure that all of the necessary information is present.

\subsection{Titles and article types}
The title is set using the command
\verb"\title{#1}", where \verb"#1" is the title of the article. The
first letter 
of the title should be capitalized with the rest in lower case. 
The title appears in bold case, but mathematical expressions within the title may be left in light-face type.

If the title is too long to use as a running head at the top of each page (apart from the
first) a short
form can be provided as an optional argument (in square brackets)
before the full title, i.e.\ \verb"\title[Short title]{Full title}".

For article types other than papers, \verb"iopart.cls"
has a generic heading \verb"\article[Short title]{TYPE}{Full title}" 
and some specific definitions given in table~\ref{arttype}. In each case (apart from Letters
to the Editor and Fast Track Communications) an 
optional argument can be used immediately after the control sequence name
to specify the short title; where no short title is given, the full title
will be used as the running head.  Not every article type has its own macro---use \verb"\article" for
any not listed.  A full list of the types of articles published by a journal is given
in the submission information available via the journal home page.
%For Letters use \verb"\letter{Full title}"; no short title is required as 
%the running head is automatically defined to be {\it Letter to the Editor}.
The generic heading could be used for 
articles such as those presented at a conference or workshop, e.g.
\small\begin{verbatim}
\article[Short title]{Workshop on High-Energy Physics}{Title}
\end{verbatim}\normalsize
Footnotes to titles may be given by using \verb"\footnote{Text of footnote.}" immediately after the title.
Acknowledgment of funding should be included in the acknowledgments section rather than in a footnote.

\begin{table}
\caption{\label{arttype}Types of article defined in the {\tt iopart.cls} 
class file.}
\footnotesize\rm
\begin{tabular*}{\textwidth}{@{}l*{15}{@{\extracolsep{0pt plus12pt}}l}}
\br
Command& Article type\\
\mr
\verb"\title{#1}"&Paper (no surtitle on first page)\\
\verb"\ftc{#1}"&Fast Track Communication\\
\verb"\review{#1}"&Review\\
\verb"\topical{#1}"&Topical Review\\
\verb"\comment{#1}"&Comment\\
\verb"\note{#1}"&Note\\
\verb"\paper{#1}"&Paper (no surtitle on first page)\\
\verb"\prelim{#1}"&Preliminary Communication\\
\verb"\rapid{#1}"&Rapid Communication\\
\verb"\letter{#1}"&Letter to the Editor\\
\verb"\article{#1}{#2}"&Other articles\\\ & (use this for any other type of article; surtitle is whatever is entered as {\tt 
\#1})\\
\br
\end{tabular*}
\end{table}

\subsection{Authors' names and addresses}
For the authors' names type \verb"\author{#1}", 
where \verb"#1" is the 
list of all authors' names. Western-style names should be written as initials then
family name, with a comma after all but the last 
two names, which are separated by `and'. Initials should {\it not} be followed by full stops. First (given) names may be used if 
desired.  Names in Chinese, Japanese and Korean styles should be written as you want them to appear in the published article. Authors in all IOP Publishing journals have the option to include their names in Chinese, Japanese or Korean characters in addition to the English name: see appendix B for details.

If the authors are at different addresses a superscripted number, e.g. $^1$, \verb"$^1$", should be used after each 
name to reference the author to his/her address.
If an author has additional information to appear as a footnote, such as 
a permanent address, a normal \LaTeX\ footnote command
should be given after the family name and address marker 
with this extra information.

The authors' affiliations follow the list of authors. 
Each address is set by using
\verb"\address{#1}" with the address as the single parameter in braces. 
If there is more 
than one address then the appropriate superscripted number, followed by a space, should come at the start of
the address.

E-mail addresses are added by inserting the 
command \verb"\ead{#1}" after the postal address(es) where \verb"#1" is the e-mail address.  
See section~\ref{startsample} for sample coding. For more than one e-mail address, please use the command 
\verb"\eads{\mailto{#1}, \mailto{#2}}" with \verb"\mailto" surrounding each e-mail address.  Please ensure
that, at the very least, you state the e-mail address of the corresponding author.

\subsection{The abstract}
The abstract follows the addresses and
should give readers concise information about the content 
of the article and indicate the main results obtained and conclusions 
drawn. It should be self-contained---there should be no references to 
figures, tables, equations, bibliographic references etc.  It should be enclosed between \verb"\begin{abstract}"
and \verb"\end{abstract}" commands.  The abstract should normally be restricted 
to a single paragraph of around 200 words.

\subsection{Subject classification numbers}
For all of our journals, we ask that you should state Physics and Astronomy Classification System (PACS)
classification numbers, except for {\it Inverse Problems} and {\it Nonlinearity} whose authors may 
use either PACS or Mathematics Subject Classification (MSC) codes.  These codes
can greatly help in the choice of suitable referees and allocation of articles to subject areas.
See \verb"http://www.aip.org/pacs" and \verb"http://www.ams.org/msc" 
for full information on these codes.

PACS or MSC numbers are included after the abstract 
using \verb"\pacs{#1}" and \verb"\ams{#1}" respectively.

After any classification numbers the command
\verb"\submitto{#1}" can be inserted, where \verb"#1" is the journal name written in full or the appropriate control sequence as
given in table~\ref{jlab1}. This command is not essential to the running of the file and can be omitted.

\subsection{Keywords}
Keywords are required for all submissions. Authors should supply a minimum of three (maximum seven) keywords appropriate to their article as a new paragraph starting \verb"\noindent{\it Keywords\/}:" after the end of the abstract.

\subsection{Making a separate title page}
To keep the header material on a separate page from the
body of the text insert \verb"\maketitle" (or \verb"\newpage") before the start of the text. 
If \verb"\maketitle" is not included the text of the
article will start immediately after the abstract.

\section{The text}
\subsection{Sections, subsections and subsubsections}
The text of articles may be divided into sections, subsections and, where necessary, 
subsubsections. To start a new section, end the previous paragraph and 
then include \verb"\section" followed by the section heading within braces. 
Numbering of sections is done {\it automatically} in the headings: 
sections will be numbered 1, 2, 3, etc, subsections will be numbered 
2.1, 2.2,  3.1, etc, and subsubsections will be numbered 2.3.1, 2.3.2, 
etc.  Cross references to other sections in the text should, where
possible, be made using 
labels (see section~\ref{xrefs}) but can also
be made manually. See section~\ref{eqnum} for information on the numbering of displayed equations. Subsections and subsubsections are 
similar to sections but 
the commands are \verb"\subsection" and \verb"\subsubsection" respectively. 
Sections have a bold heading, subsections an italic heading and 
subsubsections an italic heading with the text following on directly.
\small\begin{verbatim}
\section{This is the section title}
\subsection{This is the subsection title}
\end{verbatim}\normalsize

The first section is normally an introduction,  which should state clearly 
the object of the work, its scope and the main advances reported, with 
brief references to relevant results by other workers. In long papers it is 
helpful to indicate the way in which the paper is arranged and the results 
presented.

Footnotes should be avoided whenever possible and can often be included in the text as phrases or sentences in parentheses. If required, they should be used only for brief notes that do not fit conveniently into the text. The use of 
displayed mathematics in footnotes should be avoided wherever possible and no equations within a footnote should be numbered. 
The standard \LaTeX\ macro \verb"\footnote" should be used.  Note that in \verb"iopart.cls" the \verb"\footnote" command
produces footnotes indexed by a variety of different symbols,
whereas in published articles we use numbered footnotes.  This
is not a problem: we will convert symbol-indexed footnotes to numbered ones during the production process.

\subsection{Acknowledgments}
Authors wishing to acknowledge assistance or encouragement from 
colleagues, special work by technical staff or financial support from 
organizations should do so in an unnumbered `Acknowledgments' section 
immediately following the last numbered section of the paper. In \verb"iopart.cls" the 
command \verb"\ack" sets the acknowledgments heading as an unnumbered
section.

Please ensure that you include all of the sources of funding and the funding contract reference numbers that you are contractually obliged to acknowledge. We often receive requests to add such information very late in the production process, or even after the article is published, and we cannot always do this. Please collect all of the necessary information from your co-authors and sponsors as early as possible.

\subsection{Appendices}
Technical detail that it is necessary to include, but that interrupts 
the flow of the article, may be consigned to an appendix. 
Any appendices should be included at the end of the main text of the paper, after the acknowledgments section (if any) but before the reference list.
If there are 
two or more appendices they should be called Appendix A, Appendix B, etc. 
Numbered equations will be in the form (A.1), (A.2), etc,
figures will appear as figure A1, figure B1, etc and tables as table A1,
table B1, etc.

The command \verb"\appendix" is used to signify the start of the
appendices. Thereafter \verb"\section", \verb"\subsection", etc, will 
give headings appropriate for an appendix. To obtain a simple heading of 
`Appendix' use the code \verb"\section*{Appendix}". If it contains
numbered equations, figures or tables the command \verb"\appendix" should
precede it and \verb"\setcounter{section}{1}" must follow it.

\subsection{Some matters of style}
It will help the readers if your article is written in a clear,
consistent and concise manner. During the production process
we will try to make sure that your work is presented to its
readers in the best possible way without sacrificing the individuality of
your writing. Some recommended 
points to note, however, are the following.  These apply to all of the journals listed
in table~\ref{jlab1}.
\begin{enumerate}
\item Authors are often inconsistent in the use of `ize' and `ise' endings.
We recommend using `-ize' spellings (diagonalize, 
renormalization, minimization, etc) but there are some common 
exceptions to this, for example: devise, 
promise and advise.

\item The words table and figure should be written 
in full and {\bf not} abbreviaged to tab. and fig. Do not include `eq.', `equation' etc before an equation number or `ref.'\, `reference' etc before a reference number.
\end{enumerate}

Please check your article carefully for accuracy, consistency and clarity before
submission. Remember that your article will probably be read by many
people whose native language is not English and who may not  
be aware of many of the subtle meanings of words or idiomatic phases
present in the English language. It therefore helps if you try to keep
sentences as short and simple as possible.  If you are not a native English speaker,
please ask a native English speaker to read your paper and check its grammar.

\section{Mathematics}
\subsection{Two-line constructions}
The great advantage of \LaTeX\ 
over other text processing systems is its 
ability to handle mathematics of almost any degree of complexity. However, 
in order to produce an article suitable for publication both within a print journal and online, 
authors should exercise some restraint on the constructions used. Some equations using very small characters which are clear in a preprint style article may be difficult read in a smaller format.

For simple fractions in the text the solidus \verb"/", as in 
$\lambda/2\pi$, should be used instead of \verb"\frac" or \verb"\over", 
using parentheses where necessary to avoid ambiguity, 
for example to distinguish between $1/(n-1)$ and $1/n-1$. Exceptions to 
this are the proper fractions $\frac12$, $\frac13$, $\frac34$, 
etc, which are better left in this form. In displayed equations 
horizontal lines are preferable to solidi provided the equation is 
kept within a height of two lines. A two-line solidus should be 
avoided where possible; the construction $(\ldots)^{-1}$ should be 
used instead. For example use:
\begin{equation*}
\frac{1}{M_{\rm a}}\left(\int^\infty_0{\rm d}
\omega\;\frac{|S_o|^2}{N}\right)^{-1}\qquad\mbox{instead of}\qquad
\frac{1}{M_{\rm a}}\biggl/\int^\infty_0{\rm d}
\omega\;\frac{|S_o|^2}{N}.
\end{equation*}

\subsection{Roman and italic in mathematics}
In mathematics mode \LaTeX\ automatically sets variables in an italic 
font. In most cases authors should accept this italicization. However, 
there are some cases where it is preferable to use a Roman font; for 
instance, a Roman d for a differential d, a Roman e 
for an exponential e and a Roman i for the square root of $-1$. To 
accommodate this and to simplify the  typing of equations, \verb"iopart.cls" provides
some extra definitions. \verb"\rmd", \verb"\rme" and \verb"\rmi" 
now give Roman d, e and i respectively for use in equations, 
e.g.\ $\rmi x\rme^{2x}\rmd x/\rmd y$ 
is obtained by typing \verb"$\rmi x\rme^{2x}\rmd x/\rmd y$".

Certain other common mathematical functions, such as cos, sin, det and 
ker, should appear in Roman type. Standard \LaTeX\ provides macros for 
most of these functions 
(in the cases above, \verb"\cos", \verb"\sin", \verb"\det" and \verb"\ker" 
respectively); \verb"iopart.cls" also provides 
additional definitions for $\Tr$, $\tr$ and 
$\Or$ (\verb"\Tr", \verb"\tr" and \verb"\Or", respectively).

Subscripts and superscripts should be in Roman type if they are labels 
rather than variables or characters that take values. For example in the 
equation
\[
\epsilon_m=-g\mu_{\rm n}Bm
\]
$m$, the $z$ component of the nuclear spin, is italic because it can have 
different values whereas n is Roman because it 
is a label meaning nuclear ($\mu_{\rm n}$ 
is the nuclear magneton).

\subsection{Displayed equations in double-column journals}
Authors should bear in mind that all mathematical formulae in double-column journals will need to fit
into the width of a single column.  You may find it helpful to use a two-column layout (such as the two-column
option in \verb"iopart.cls") in your submission so that you can check the width of equations.

\subsection{Special characters for mathematics}
Bold italic characters can be used in our journals to signify vectors (rather
than using an upright bold or an over arrow). To obtain this effect when using \verb"iopart.cls",
use \verb"\bi{#1}" within maths mode, e.g. $\bi{ABCdef}$. Similarly, in \verb"iopart.cls", if upright 
bold characters are required in maths, use \verb"\mathbf{#1}" within maths
mode, e.g. $\mathbf{XYZabc}$. The calligraphic (script) uppercase alphabet
is obtained with \verb"\mathcal{AB}" or \verb"\cal{CD}" 
($\mathcal{AB}\cal{CD}$).

The American Mathematical Society provides a series of extra symbol fonts
to use with \LaTeX\ and packages containing the character definitions to
use these fonts. Authors wishing to use Fraktur 
\ifiopams$\mathfrak{ABC}$ \fi
or Blackboard Bold \ifiopams$\mathbb{XYZ}$ \fi can include the appropriate
AMS package (e.g. \verb"amsgen.sty", \verb"amsfonts.sty", \verb"amsbsy.sty", \verb"amssymb.sty") with a 
\verb"\usepackage" command or add the command \verb"\usepackage{iopams}"
which loads the four AMS packages mentioned above and also provides
definitions for extra bold characters (all Greek letters and some other
additional symbols).

The package \verb"iopams.sty" uses the definition \verb"\boldsymbol" in \verb"amsbsy.sty"
which allows individual non-alphabetical symbols and Greek letters to be 
made bold within equations.
The bold Greek lowercase letters \ifiopams$\balpha \ldots\bomega$,\fi 
are obtained with the commands 
\verb"\balpha" \dots\ \verb"\bomega" (but note that
bold eta\ifiopams, $\bfeta$,\fi\ is \verb"\bfeta" rather than \verb"\beta")
and the capitals\ifiopams, $\bGamma\ldots\bOmega$,\fi\ with commands 
\verb"\bGamma" \dots\
\verb"\bOmega". Bold versions of the following symbols are
predefined in \verb"iopams.sty": 
bold partial\ifiopams, $\bpartial$,\fi\ \verb"\bpartial",
bold `ell'\ifiopams, $\bell$,\fi\  \verb"\bell", 
bold imath\ifiopams, $\bimath$,\fi\  \verb"\bimath", 
bold jmath\ifiopams, $\bjmath$,\fi\  \verb"\bjmath", 
bold infinity\ifiopams, $\binfty$,\fi\ \verb"\binfty", 
bold nabla\ifiopams, $\bnabla$,\fi\ \verb"\bnabla", 
bold centred dot\ifiopams, $\bdot$,\fi\  \verb"\bdot". Other 
characters are made bold using 
\verb"\boldsymbol{\symbolname}".

Please do not use the style file \verb"amsmath.sty" (part of the AMSTeX package) in conjunction with \verb"iopart.cls". This will result in several errors. To make use of the macros defined in \verb"amsmath.sty", \verb"iopart.cls" provides the file \verb"setstack.sty" which reproduces the following useful macros from \verb"amsmath.sty":
\small\begin{verbatim}
\overset \underset \sideset \substack \boxed   \leftroot
\uproot  \dddot    \ddddot  \varrow   \harrow
\end{verbatim}\normalsize

If the mathematical notation
that you need is best handled in \verb"amsmath.sty" you might want to consider using an article class
other than \verb"iopart.cls". We accept submissions using any class or style files.

Table~\ref{math-tab2} lists some other macros for use in 
mathematics with a brief description of their purpose.

\begin{table}
\caption{\label{math-tab2}Other macros defined in {\tt iopart.cls} for use in maths.}
\begin{tabular*}{\textwidth}{@{}l*{15}{@{\extracolsep{0pt plus
12pt}}l}}
\br
Macro&Result&Description\\
\mr
\verb"\fl"&&Start line of equation full left\\
\verb"\case{#1}{#2}"&$\case{\#1}{\#2}$&Text style fraction in display\\
\verb"\Tr"&$\Tr$&Roman Tr (Trace)\\
\verb"\tr"&$\tr$&Roman tr (trace)\\
\verb"\Or"&$\Or$&Roman O (of order of)\\
\verb"\tdot{#1}"&$\tdot{x}$&Triple dot over character\\
\verb"\lshad"&$\lshad$&Text size left shadow bracket\\
\verb"\rshad"&$\rshad$&Text size right shadow bracket\\
\br
\end{tabular*}
\end{table}

\subsection{Alignment of displayed equations}

The normal style for aligning displayed equations in our published journal articles is to align them left rather than centre. The \verb"iopart.cls" class file automatically does this and indents each line of a display.  In \verb"iopart.cls", to make any line start at the left margin of the page, add \verb"\fl" at start of the line (to indicate full left).

Using the \verb"eqnarray" environment equations will naturally be aligned left and indented without the use of any ampersands for alignment, see equations (\ref{eq1}) and (\ref{eq2})
\begin{eqnarray}
\alpha + \beta =\gamma^2, \label{eq1}\\
\alpha^2 + 2\gamma + \cos\theta = \delta. \label{eq2} 
\end{eqnarray}
This is the normal equation style for our journals.

Where some secondary alignment is needed, for instance a second part of an equation on a second line, a single ampersand is added at the point of alignment in each line  (see  (\ref{eq3}) and (\ref{eq4})).
\begin{eqnarray}
\alpha &=2\gamma^2 + \cos\theta + \frac{XY \sin\theta}{X+ Y\cos\theta} \label{eq3}\\
 & = \delta\theta PQ \cos\gamma. \label{eq4} 
\end{eqnarray}

Two points of alignment are possible using two ampersands for alignment (see  (\ref{eq5}) and (\ref{eq6})).  Note in this case extra space \verb"\qquad" is added before the second ampersand in the longest line (the top one) to separate the condition from the equation. 
\begin{eqnarray}
\alpha &=2\gamma^2 + \cos\theta + \frac{XY \sin\theta}{X+ Y\cos\theta}\qquad& \theta > 1 \label{eq5}\\
 & = \delta\theta PQ \cos\gamma &\theta \leq 1.\label{eq6} 
\end{eqnarray}

For a long equation which has to be split over more than one line the first line should start at the left margin, this is achieved by inserting \verb"\fl" (full left) at the start of the line. The use of the alignment parameter \verb"&" is not necessary unless some secondary alignment is needed.
\begin{eqnarray}
\fl \alpha + 2\gamma^2 = \cos\theta + \frac{XY \sin\theta}{X+ Y\cos\theta} +  \frac{XY \sin\theta}{X- Y\cos\theta} +
+ \left(\frac{XY \sin\theta}{X+ Y\cos\theta}\right)^2 \nonumber\\
+  \left(\frac{XY \sin\theta}{X- Y\cos\theta}\right)^2.\label{eq7} 
\end{eqnarray}

The plain \TeX\ command \verb"\eqalign" can be used within an \verb"equation" environment to obtain a multiline equation with a single centred number, for example
\begin{equation}
\eqalign{\alpha + \beta =\gamma^2 \cr
\alpha^2 + 2\gamma + \cos\theta = \delta.} 
\end{equation}

During the production process we will break equations as appropriate for the page layout of the journal. If you are submitting to a double-column journal and wish to review how your equations will break, you may find the double-column layout described in section \ref{dblcol} useful.

\subsection{Miscellaneous points}
The following points on the layout of mathematics apply whichever class file you use.

Exponential expressions, especially those containing subscripts or 
superscripts, are clearer if the notation $\exp(\ldots)$ is used, except for 
simple examples. For instance $\exp[\rmi(kx-\omega t)]$ and $\exp(z^2)$ are 
preferred to $\e^{\rmi(kx-\omega t)}$ and $\e^{z^2}$, but 
$\e^x$ 
is acceptable.

Similarly the square root sign $\sqrt{\phantom{b}}$ should 
only be used with relatively
simple expressions, e.g.\ $\sqrt2$ and $\sqrt{a^2+b^2}$;
in other cases the 
power $1/2$ should be used; for example, $[(x^2+y^2)/xy(x-y)]^{1/2}$.

It is important to distinguish between $\ln = \log_\e$ and $\lg 
=\log_{10}$. Braces, brackets and parentheses should be used in the 
following order: $\{[(\;)]\}$. The same ordering of brackets should be 
used within each size. However, this ordering can be ignored if the
brackets have a 
special meaning (e.g.\ if they denote an average or a function).

Decimal fractions should always be preceded by a zero: for example 0.123 {\bf not} .123.
For long numbers use thin spaces after every third character away from the position of the decimal point, unless 
this leaves a single separated character: e.g.\ $60\,000$, $0.123\,456\,78$ 
but 4321 and 0.7325.

Equations should be followed by a full stop (periods) when at the end
of a sentence.

\subsection{Equation numbering and layout in {\tt iopart.cls}}
\label{eqnum}
\LaTeX\ provides facilities for automatically numbering equations 
and these should be used where possible. Sequential numbering (1), (2), 
etc, is the default numbering system although in \verb"iopart.cls", if the command
\verb"\eqnobysec" is included in the preamble, equation numbering
by section is obtained, e.g.\ 
(2.1), (2.2), etc. Equation numbering by section is used in appendices automatically when the \verb"\appendix" command is used, even if sequential numbering has been used in the rest of the article. 
Refer to equations in the text using the equation number in parentheses. It is not normally necessary to include the word equation before the number; and abbreviations such as eqn or eq should not be used.
In \verb"iopart.cls", there are alternatives to the standard \verb"\ref" command that you might
find useful---see \tref{abrefs}.

Sometimes it is useful to number equations as parts of the same
basic equation. This can be accomplished in \verb"iopart.cls" by inserting the 
commands \verb"\numparts" before the equations concerned and 
\verb"\endnumparts" when reverting to the normal sequential numbering.
For example using \verb"\numparts \begin{eqnarray}" ... \verb"\end{eqnarray} \endnumparts":

\numparts
\begin{eqnarray}
T_{11}&=(1+P_\e)I_{\uparrow\uparrow}-(1-P_\e)
I_{\uparrow\downarrow},\label{second}\\
T_{-1-1}&=(1+P_\e)I_{\downarrow\downarrow}-(1-P_\e)I_{\uparrow\downarrow},\\
S_{11}&=(3+P_\e)I_{\downarrow\uparrow}-(3-P_e)I_{\uparrow\uparrow},\\
S_{-1-1}&=(3+P_\e)I_{\uparrow\downarrow}-(3-P_\e)
I_{\downarrow\downarrow}.
\end{eqnarray}
\endnumparts

Equation labels within the \verb"\eqnarray" environment will be referenced
as subequations, e.g. (\ref{second}).

\subsection{Miscellaneous extra commands for displayed equations}
The \verb"\cases" command has been amended slightly in \verb"iopart.cls" to 
increase the space between the equation and the condition. 
\Eref{cases} 
demonstrates simply the output from the \verb"\cases" command
\begin{equation}
\label{cases}
X=\cases{1&for $x \ge 0$\\
-1&for $x<0$\\}
\end{equation}
%The code used was:
%\small\begin{verbatim}
%\begin{equation}
%\label{cases}
%X=\cases{1&for $x \ge 0$\\
%-1&for $x<0$\\}
%\end{equation}
%\end{verbatim}
%\normalsize

To obtain text style fractions within displayed maths the command 
\verb"\case{#1}{#2}" can be used instead
of the usual \verb"\frac{#1}{#2}" command or \verb"{#1 \over #2}".

When two or more short equations are on the same line they should be 
separated by a `qquad space' (\verb"\qquad"), rather than
\verb"\quad" or any combination of \verb"\,", \verb"\>", \verb"\;" 
and \verb"\ ".

\section{Referencing\label{except}}
Two different styles of referencing are in common use: 
the Harvard alphabetical system and the Vancouver numerical system. 
All journals to which this document applies allow the use of either the Harvard or Vancouver system, 
except for {\it Physics in Medicine and Biology} and {\it Physiological Measurement}
for which authors {\it must\/} use the Harvard referencing style (with the titles of journal
articles given, and final page numbers given).

\subsection{Harvard (alphabetical) system}
In the Harvard system the name of the author appears in the text together 
with the year of publication. As appropriate, either the date or the name 
and date are included within parentheses. Where there are only two authors 
both names should be given in the text; if there are more than two 
authors only the first name should appear followed by `{\it et al}' 
(which can be obtained in \verb"iopart.cls" by typing \verb"\etal"). When two or 
more references to work by one author or group of authors occur for the 
same year they should be identified by including a, b, etc after the date 
(e.g.\ 2012a). If several references to different pages of the same article 
occur the appropriate page number may be given in the text, e.g.\ Kitchen 
(2011, p 39).

The reference list at the end of an article consists of an 
unnumbered `References' section containing an
alphabetical listing by authors' names. References with the same author list are ordered by date, oldest first.
The reference list in the 
preprint style is started in \verb"iopart.cls" by including the command \verb"\section*{References}" and then
\verb"\begin{harvard}".
Individual references start with \verb"\item[]" and the reference list is completed with \verb"\end{harvard}".
There is also a shortened form of the coding: \verb"\section*{References}"
and \verb"\begin{harvard}" can be replaced by the single command
\verb"\References", and \verb"\end{harvard}" can be shortened to
\verb"\endrefs".

\subsection{Vancouver (numerical) system}
In the Vancouver system references are numbered sequentially 
throughout the text. The numbers occur within square brackets and one 
number can be used to designate several references. A numerical 
reference list in the \verb"iopart" style is started by including the 
command \verb"\section*{References}" and then
\verb"\begin{thebibliography}{<num>}", where \verb"<num>" is the largest
number in the reference list (or any other number with the same number
of digits).  The 
reference list gives the references in 
numerical order, individual references start with \verb"\bibitem{label}". The list is completed by
\verb"\end{thebibliography}". Short forms of the commands are again
available: \verb"\Bibliography{<num>}" can be used at the start of the
references section and \verb"\endbib" at the end.

A variant of this system is to use labels instead of numbers within 
square brackets, in this case references in the list should start with \verb"\bibitem[label-text]". This method is allowed for all journals that accept numerical references.

\subsection{BibTeX\label{bibtex}}
If you are using BibTeX, see the earlier section \ref{bibby} for information on what \verb".bst" file to use.
The output that you get will differ slightly from that specified in the rest of this section,
but this is not a problem as long as all the relevant information is present.

\subsection{References, general}
A complete reference should provide the reader with enough information to 
locate the item concerned. Up to ten authors may be given in a particular reference; where 
there are more than ten only the first should be given followed by 
`{\it et al}'.  If you are using BibTeX
and the \verb".bst" file that you are using includes more than 10 authors, do not worry about this:
we can correct this during the production process.  Abbreviate a journal name only in accordance with the journal's
own recommendations for abbreviation---if in doubt, leave it unabbreviated.

The terms {\it loc.\ cit.}\ and {\it ibid}.\ should not be used. 
Unpublished conferences and reports should generally not be included 
in the reference list if a published version of the work exists. Articles in the course of publication should 
include the article title and the journal of publication, if known. 
A reference to a thesis submitted for a higher degree may be included 
if it has not been superseded by a published 
paper---please state the institution where the work was submitted.

The basic structure of a reference in the reference list is the same in both the alphabetical and numerical systems, the only difference the code at the start of the reference. Alphabetical references are preceded by \verb"\item[]", numerical by \verb"\bibitem{label}" or just \verb"\item" to generate a number or \verb"\nonum" where a reference is not the first in a group of references under the same number.

Note that footnotes to the text should not be 
included in the reference list, but should appear at the bottom of the relevant page by using the \verb"\footnote" command.

\subsection{References to journal articles}
The following guidance applies if you are producing your reference list `by hand'; that is,
without the help of BibTeX.  See section~\ref{bibby} for BibTeX help.

Article references in published articles in our journals contain three changes of 
font:
the authors and date appear in Roman type, the journal title in 
italic, the volume number in bold and the page numbers in Roman again. 
A typical journal entry would be:

\smallskip
\begin{harvard}
\item[] Spicer P E, Nijhoff F W and van der Kamp P H 2011 {\it Nonlinearity} {\bf 24} 2229
\end{harvard}
\smallskip

\noindent which would be obtained by typing, within the references
environment 
\small\begin{verbatim}
\item[] Spicer P E, Nijhoff F W and van der Kamp P H 2011 {\it Nonlinearity} 
{\bf 24} 2229
\end{verbatim}\normalsize

\noindent Features to note are the following.

\begin{enumerate}
\item The authors should be in the form of surname (with only the first 
letter capitalized) followed by the initials with no 
periods after the initials. Authors should be separated by a comma 
except for the last two which should be separated by `and' with no 
comma preceding it.

\item The year of publication follows the authors and is not in parentheses.

\item Titles of journal articles can also be included (in Roman (upright) text after the year). Article titles are required in reference lists for {\it Inverse Problems, Journal of Neural Engineering, Measurement Science and Technology, Physical Biology, Physics in Medicine and Biology\/} and {\it Physiological Measurement}.

\item The journal is in italic and is abbreviated. If a journal has several parts denoted by 
different letters the part letter should be inserted after the journal in Roman type (e.g.\ 
{\it Phys.\ Rev.\ \rm A}). \verb"iopart.cls" includes macros for abbreviated titles of all journals handled by IOP Publishing (see table~\ref{jlab2}) and some other common titles (table \ref{jlab3}).

\item The volume number is bold; the page number is Roman.
 Both the initial and final page numbers should be given where possible---note that for {\it Reports on Progress in Physics\/}, {\it Physiological Measurement} and {\it Physics in Medicine and Biology}
 the final page number is {\it required}. The final page number should be in 
the shortest possible form and separated from the initial page number by an en rule (\verb"--"), e.g.\ 1203--14.

\item Where there are two or more references with identical authors, 
the authors' names should be repeated for the second and subsequent references. Each individual publication should be presented as a separate reference, although in the numerical system one number can be used for several references. This facilitates linking in the online journal. 
\end{enumerate}

\subsubsection{Article numbering.}
Many journals now use article-numbering systems that do not fit the conventional {\it year-journal-volume-page numbers} pattern. Some examples are:

\numrefs{1}
\item Carlip S and Vera R 1998 {\it Phys. Rev.} D {\bf 58} 011345 
\item Davies K and Brown G 1997 {\it J. High Energy Phys.} JHEP12(1997)002
\item Hannestad S 2005 {\it J. Cosmol. Astropart. Phys.} JCAP02(2005)011
\item Hilhorst H J 2005 {\it J. Stat. Mech.} L02003
\item Gundlach C 1999 {\it Liv. Rev. Rel.} 1994-4
\endnumrefs

\noindent The website of the journal you are citing should state the correct format for citations.

\subsection{Preprint references}
Preprints may be referenced but if the article concerned has been published in a peer-reviewed journal, that reference should take precedence. If only a preprint reference can be given, it is helpful to include the article title. Examples are:
\vskip6pt
\numrefs{1}
\item Neilson D and Choptuik M 2000 {\it Class. Quantum Grav.} {\bf 17} 761 (arXiv:gr-qc/9812053)
\item Sundu H, Azizi K, S\"ung\"u J Y and Yinelek N 2013 Properties of $D_{s2}^*(2573)$ charmed-strange tensor meson arXiv:1307.6058
\endnumrefs

\noindent For preprints added to arXiv.org after April 2007 it is not necessary to include the subject area, however this information can be included in square brackets after the number if desired, e.g.
\numrefs{1}
\item Sundu H, Azizi K, S\"ung\"u J Y and Yinelek N 2013 Properties of $D_{s2}^*(2573)$ charmed-strange tensor meson arXiv:1307.6058 [hep-ph]
\endnumrefs

\subsection{References to books, conference proceedings and reports}

References to books, proceedings and reports are similar, but have only two
changes of font. The authors and date of publication are in Roman, the 
title of the book is in italic, and the editors, publisher, 
town of publication 
and page number are in Roman. A typical reference to a book and a
conference paper might be

\smallskip
\begin{harvard}
\item[] Dorman L I 1975 {\it Variations of Galactic Cosmic Rays} 
(Moscow: Moscow State University Press) p~103
\item[] Caplar R and Kulisic P 1973 {\it Proc.\
Int.\ Conf.\ on Nuclear Physics (Munich)} vol~1 (Amsterdam:  
North-Holland/American Elsevier) p~517
\end{harvard}
\smallskip

\noindent which would be obtained by with the code
\small\begin{verbatim}
\item[] Dorman L I 1975 {\it Variations of Galactic Cosmic Rays} 
(Moscow: Moscow State University Press) p~103
\item[] Caplar R and Kulisic P 1973 {\it Proc. Int. Conf. on Nuclear 
Physics (Munich)} vol~1 (Amsterdam: North-Holland/American 
Elsevier) p~517
\end{verbatim}\normalsize
\noindent

\noindent Features to note are the following.
\begin{enumerate}
\item Book titles are in italic and should be spelt out in full with 
initial capital letters for all except minor words. Words such as 
Proceedings, Symposium, International, Conference, Second, etc should 
be abbreviated to Proc., Symp., Int., Conf., 2nd, 
respectively, but the rest of the title should be given in full, 
followed by the date of the conference and the 
town or city where the conference was held. For 
laboratory reports the laboratory should be spelt out wherever 
possible, e.g.\ {\it Argonne National Laboratory Report}.

\item The volume number, for example, vol~2, should be followed by 
the editors, if any, in the form ed~A~J~Smith and P~R~Jones. Use 
\etal if there are more than two editors. Next comes the town of 
publication and publisher, within brackets and separated by a colon, 
and finally the page numbers preceded by p if only one number is given 
or pp if both the initial and final numbers are given.

\item If a book is part of a series (for examples, {\it Springer Tracts in Modern Physics\/}), the series title and volume number is given in parentheses after the book title. Whereas for an individual volume in a multivolume set, the set title is given first, then the volume title. 
\end{enumerate}
\smallskip
\begin{harvard}
\item[]Morse M 1996 Supersonic beam sources {\it Atomic Molecular and Optical Physics\/} ({\it Experimental Methods in the Physical Sciences\/} vol 29) ed F B Dunning and R Hulet (San Diego, CA: Academic)  
\item[]Fulco C E, Liverman C T and Sox H C (eds) 2000 {\it Gulf War and Health\/} vol 1 {\it Depleted Uranium, Pyridostigmine Bromide, Sarin, and Vaccines\/} (Washington, DC: The National Academies Press)
\end{harvard}

\section{Cross-referencing\label{xrefs}}
The facility to cross reference items in the text is very useful when 
composing articles as the precise form of the article may be uncertain at the start 
and  revisions and amendments may subsequently be made. 
\LaTeX\ provides excellent facilities for doing cross-referencing
and these can be very useful in preparing articles.

\subsection{References}
\label{refs}
Cross referencing is useful for numeric reference lists because, if it 
is used, adding 
another reference to the list does not then involve renumbering all 
subsequent references. It is not necessary for referencing 
in the Harvard system where the final reference list is alphabetical 
and normally no other changes are necessary when a reference is added or
deleted.
When using \LaTeX , two passes (under certain circumstances, three passes)
are necessary initially to get the cross references right 
but once they are correct a single run is usually sufficient provided an 
\verb".aux" file is available and the file 
is run to the end each time.
If the 
reference list contains an entry \verb"\bibitem{label}", 
this command 
will produce the correct number in the reference list and 
\verb"\cite{label}" will produce the number within square brackets in the 
text. \verb"label" may contain letters, numbers 
or punctuation characters but must not contain spaces or commas. It is also
recommended that the underscore character \_{} is not used in cross
referencing. 
Thus labels of the form 
\verb"eq:partial", \verb"fig:run1", \verb"eq:dy'", 
etc, may be used. When several 
references occur together in the text \verb"\cite" may be used with 
multiple labels with commas but no spaces separating them; 
the output will be the 
numbers within a single pair of square brackets with a comma and a 
thin space separating the numbers. Thus \verb"\cite{label1,label2,label4}"
would give [1,\,2,\,4]. Note that no attempt is made by the style file to sort the 
labels and no shortening of groups of consecutive numbers is done.
Authors should therefore either try to use multiple labels in the correct 
order, or use a package such as \verb"cite.sty" that reorders labels
correctly.

The numbers for the cross referencing are generated in the order the 
references appear in the reference list, so that if the entries in the 
list are not in the order in which the references appear in the text 
then the 
numbering within the text will not be sequential. To correct this 
change the ordering of the entries in the reference list and then 
rerun the \LaTeX\ file {\it twice}.  Please ensure that all references resolve correctly: check the \verb".log" file
for undefined or multiply-defined citations, and check that the output does not contain question
marks that indicate unresolved references.

\subsection{Equation numbers, sections, subsections, figures and 
tables}
Labels for equation numbers, sections, subsections, figures and tables 
are all defined with the \verb"\label{label}" command and cross references 
to them are made with the \verb"\ref{label}" command.

Any section, subsection, subsubsection, appendix or subappendix 
command defines a section type label, e.g. 1, 2.2, A2, A1.2 depending 
on context. A typical article might have in the code of its introduction 
`The results are discussed in section\verb"~\ref{disc}".' and
the heading for the discussion section would be:
\small\begin{verbatim}
\section{Results}\label{disc}
\end{verbatim}\normalsize
Labels to sections, etc, may occur anywhere within that section except
within another numbered environment. 
Within a maths environment labels can be used to tag equations which are 
referred to within the text.

In addition to the standard \verb"\ref{<label>}", in \verb"iopart.cls" the abbreviated
forms given in \tref{abrefs}
are available for reference to standard parts of the text.

\Table{\label{abrefs}Alternatives to the normal references command {\tt $\backslash$ref} available in {\tt iopart.cls},
and the text generated by
them. Note it is not normally necessary to include the word equation
before an equation number except where the number starts a sentence. The
versions producing an initial capital should only be used at the start of
sentences.} 
\br
Reference&Text produced\\
\mr
\verb"\eref{<label>}"&(\verb"<num>")\\
\verb"\Eref{<label>}"&Equation (\verb"<num>")\\
\verb"\fref{<label>}"&figure \verb"<num>"\\
\verb"\Fref{<label>}"&Figure \verb"<num>"\\
\verb"\sref{<label>}"&section \verb"<num>"\\
\verb"\Sref{<label>}"&Section \verb"<num>"\\
\verb"\tref{<label>}"&table \verb"<num>"\\
\verb"\Tref{<label>}"&Table \verb"<num>"\\
\br
\endTable

\section{Tables and table captions}
Tables are numbered serially and referred to in the text 
by number (table 1, etc, {\bf not} tab. 1). Each table should have an 
explanatory caption which should be as concise as possible. If a table 
is divided into parts these should be labelled \pt(a), \pt(b), 
\pt(c), etc but there should be only one caption for the whole 
table, not separate ones for each part.

In the preprint style the tables may be included in the text 
or listed separately after the reference list 
starting on a new page.

\subsection{The basic table format}
The standard form for a table in \verb"iopart.cls" is:
\small\begin{verbatim}
\begin{table}
\caption{\label{label}Table caption.}
\begin{indented}
\item[]\begin{tabular}{@{}llll}
\br
Head 1&Head 2&Head 3&Head 4\\
\mr
1.1&1.2&1.3&1.4\\
2.1&2.2&2.3&2.4\\
\br
\end{tabular}
\end{indented}
\end{table}
\end{verbatim}\normalsize

\noindent Points to note are:
\begin{enumerate}
\item The caption comes before the table. It should have a period at
the end.

\item Tables are normally set in a smaller type than the text.
The normal style is for tables to be indented. This is accomplished
by using \verb"\begin{indented}" \dots\ \verb"\end{indented}"
and putting \verb"\item[]" before the start of the tabular environment.
Omit these
commands for any tables which will not fit on the page when indented.

\item The default is for columns to be aligned left and 
adding \verb"@{}" omits the extra space before the first column.

\item Tables have only horizontal rules and no vertical ones. The rules at
the top and bottom are thicker than internal rules and are set with
\verb"\br" (bold rule). 
The rule separating the headings from the entries is set with
\verb"\mr" (medium rule).  These are special \verb"iopart.cls" commands.

\item Numbers in columns should be aligned on the decimal point;
to help do this a control sequence \verb"\lineup" has been defined
in \verb"iopart.cls"
which sets \verb"\0" equal to a space the size of a digit, \verb"\m"
to be a space the width of a minus sign, and \verb"\-" to be a left
overlapping minus sign. \verb"\-" is for use in text mode while the other
two commands may be used in maths or text.
(\verb"\lineup" should only be used within a table
environment after the caption so that \verb"\-" has its normal meaning
elsewhere.) See table~\ref{tabone} for an example of a table where
\verb"\lineup" has been used.
\end{enumerate}

\begin{table}
\caption{\label{tabone}A simple example produced using the standard table commands 
and $\backslash${\tt lineup} to assist in aligning columns on the 
decimal point. The width of the 
table and rules is set automatically by the 
preamble.}

\begin{indented}
\lineup
\item[]\begin{tabular}{@{}*{7}{l}}
\br                              
$\0\0A$&$B$&$C$&\m$D$&\m$E$&$F$&$\0G$\cr 
\mr
\0\023.5&60  &0.53&$-20.2$&$-0.22$ &\01.7&\014.5\cr
\0\039.7&\-60&0.74&$-51.9$&$-0.208$&47.2 &146\cr 
\0123.7 &\00 &0.75&$-57.2$&\m---   &---  &---\cr 
3241.56 &60  &0.60&$-48.1$&$-0.29$ &41   &\015\cr 
\br
\end{tabular}
\end{indented}
\end{table}

\subsection{Simplified coding and extra features for tables}
The basic coding format can be simplified using extra commands provided in
the \verb"iopart" class file. The commands up to and including 
the start of the tabular environment
can be replaced by
\small\begin{verbatim}
\Table{\label{label}Table caption}
\end{verbatim}\normalsize
and this also activates the definitions within \verb"\lineup".
The final three lines can also be reduced to \verb"\endTable" or
\verb"\endtab". Similarly for a table which does not fit on the page when indented
\verb"\fulltable{\label{label}caption}" \dots\ \verb"\endfulltable"
can be used. \LaTeX\ optional positional parameters can, if desired, be added after 
\verb"\Table{\label{label}caption}" and \verb"\fulltable{\label{label}caption}".

\verb"\centre{#1}{#2}" can be used to centre a heading 
\verb"#2" over \verb"#1" 
columns and \verb"\crule{#1}" puts a rule across 
\verb"#1" columns. A negative 
space \verb"\ns" is usually useful to reduce the space between a centred 
heading and a centred rule. \verb"\ns" should occur immediately after the 
\verb"\\" of the row containing the centred heading (see code for
\tref{tabl3}). A small space can be 
inserted between rows of the table 
with \verb"\ms" and a half line space with \verb"\bs" 
(both must follow a \verb"\\" but should not have a 
\verb"\\" following them).

\Table{\label{tabl3}A table with headings spanning two columns and containing notes. 
To improve the 
visual effect a negative skip ($\backslash${\tt ns})
has been put in between the lines of the 
headings. Commands set-up by $\backslash${\tt lineup} are used to aid 
alignment in columns. $\backslash${\tt lineup} is defined within
the $\backslash${\tt Table} definition.}
\br
&&&\centre{2}{Separation energies}\\
\ns
&Thickness&&\crule{2}\\
Nucleus&(mg\,cm$^{-2}$)&Composition&$\gamma$, n (MeV)&$\gamma$, 2n (MeV)\\
\mr
$^{181}$Ta&$19.3\0\pm 0.1^{\rm a}$&Natural&7.6&14.2\\
$^{208}$Pb&$\03.8\0\pm 0.8^{\rm b}$&99\%\ enriched&7.4&14.1\\
$^{209}$Bi&$\02.86\pm 0.01^{\rm b}$&Natural&7.5&14.4\\
\br
\end{tabular}
\item[] $^{\rm a}$ Self-supporting.
\item[] $^{\rm b}$ Deposited over Al backing.
\end{indented}
\end{table}

Units should not normally be given within the body of a table but 
given in brackets following the column heading; however, they can be 
included in the caption for long column headings or complicated units. 
Where possible tables should not be broken over pages. 
If a table has related notes these should appear directly below the table
rather than at the bottom of the page. Notes can be designated with
footnote symbols (preferable when there are only a few notes) or
superscripted small roman letters. The notes are set to the same width as
the table and in normal tables follow after \verb"\end{tabular}", each
note preceded by \verb"\item[]". For a full width table \verb"\noindent"
should precede the note rather than \verb"\item[]". To simplify the coding 
\verb"\tabnotes" can, if desired, replace \verb"\end{tabular}" and 
\verb"\endtabnotes" replaces
\verb"\end{indented}\end{table}".

If all the tables are grouped at the end of a document
the command \verb"\Tables" is used to start a new page and 
set a heading `Tables and table captions'. If the tables follow an appendix then add the command \verb"\noappendix" to revert to normal style numbering.

\section{Figures and figure captions}

Figures (with their captions) can be incorporated into the text at the appropriate position or grouped together
at the end of the article. If the figures are at the end of the article and follow an appendix then in \verb"iopart.cls" you can add the command \verb"\noappendix" to revert to normal style numbering.  We remind you that you must seek permission
to reuse any previously-published figures, and acknowledge their use correctly---see section \ref{copyright}.

\subsection{Inclusion of graphics files\label{figinc}}
Using the \verb"graphicx" package graphics files can 
be included within figure and center environments at an 
appropriate point within the text using code such as:
\small\begin{verbatim}
\includegraphics{file.eps}
\end{verbatim}\normalsize
The \verb"graphicx" package supports various optional arguments
to control the appearance of the figure. Other similar 
packages can also be used (e.g. \verb"graphics", \verb"epsf").   Whatever package you use,
you must include it explicitly after the \verb"\documentclass" declaration using (say)
\verb"\usepackage{graphicx}".

For more detail about graphics inclusion see the documentation 
of the \verb"graphicx" package, refer to one of the books on \LaTeX , e.g. Goosens M, Rahtz S and Mittelbach F 1997 {\it The }\LaTeX\ {\it Graphics Companion\/} 
(Reading, MA: Addison-Wesley),
or download some of the excellent free documentation available via the Comprehensive
TeX Archive Network (CTAN) \verb"http://www.ctan.org"---{in particular see Reckdahl K 2006 {\it Using Imported Graphics in }\LaTeX\ {\it and }pdf\LaTeX\ \verb"http://www.ctan.org/tex-archive/info/epslatex".

IOP Publishing's graphics guidelines (available via the `Author Guidelines' link at \verb"http://authors.iop.org") provide further information on preparing \verb".eps" files.

We prefer you to use \verb".eps" files for your graphics, but we realise that converting other
formats of graphics to \verb".eps" format can be troublesome.  If you use PDF or bitmap-format graphics
such as JPG or PNG that need to be included using the pdf\TeX\ package, this is OK, but please bear
in mind that the PDF you submit should use PDF standard 1.4 or lower (use \verb"\pdfminorversion=4" at the
start of the file).

The main \LaTeX\ file must read in graphics files and subsidiary \LaTeX\ files from the current directory,
{\it not} from a subdirectory.  Your submission files are stored on our systems in a single location and we will not be able to process your
TeX file automatically if it relies on organization of the files into subdirectories.

\subsection{Captions}
Below each figure should be a brief caption describing it and, if 
necessary, interpreting the various lines and symbols on the figure. 
As much lettering as possible should be removed from the figure itself 
and included in the caption. If a figure has parts, these should be 
labelled ($a$), ($b$), ($c$), etc and all parts should be described 
within a single caption. \Tref{blobs} gives the definitions for describing 
symbols and lines often used within figure captions (more symbols are 
available when using the optional packages loading the AMS extension fonts).

\subsection{Supplementary Data}
All of our journals encourage authors to submit supplementary data attachments to 
enhance the online versions of published research articles. Supplementary data 
enhancements typically consist of video clips, animations or
data files, tables of extra information or extra figures. They can 
add to the reader's understanding and present results in attractive ways that go 
beyond what can be presented in the PDF version of the article. 
Guidelines on supplementary data file formats 
are available via the `Author Guidelines' link at \verb"http://authors.iop.org".

Software, in the form of input scripts for mathematical packages (such as Mathematica notebook files), or
source code that can be interpreted or compiled (such as Python scripts or Fortran or C programs), or executable
files, can sometimes be accepted as supplementary data, but the journal may ask you for assurances about
the software and distribute them from the article web page only subject to a disclaimer.  Contact the journal
in the first instance if you want to submit software.

\begin{table}[t]
\caption{\label{blobs}Control sequences to describe lines and symbols in figure 
captions.}
\begin{indented}
\item[]\begin{tabular}{@{}lllll}
\br
Control sequence&Output&&Control sequence&Output\\
\mr
\verb"\dotted"&\dotted        &&\verb"\opencircle"&\opencircle\\
\verb"\dashed"&\dashed        &&\verb"\opentriangle"&\opentriangle\\
\verb"\broken"&\broken&&\verb"\opentriangledown"&\opentriangledown\\
\verb"\longbroken"&\longbroken&&\verb"\fullsquare"&\fullsquare\\
\verb"\chain"&\chain          &&\verb"\opensquare"&\opensquare\\
\verb"\dashddot"&\dashddot    &&\verb"\fullcircle"&\fullcircle\\
\verb"\full"&\full            &&\verb"\opendiamond"&\opendiamond\\
\br
\end{tabular}
\end{indented}
\end{table}

\clearpage

\appendix

\section{List of macros for formatting text, figures and tables}

\begin{table}[hb]
\caption{Macros available for use in text in {\tt iopart.cls}. Parameters in square brackets are optional.}
\footnotesize\rm
\begin{tabular}{@{}*{7}{l}}
\br
Macro name&Purpose\\
\mr
\verb"\title[#1]{#2}"&Title of article and short title (optional)\\
\verb"\paper[#1]{#2}"&Title of paper and short title (optional)\\
\verb"\letter{#1}"&Title of Letter to the Editor\\
\verb"\ftc{#1}"&Title of Fast Track Communication\\
\verb"\rapid[#1]{#2}"&Title of Rapid Communication and short title (optional)\\
\verb"\comment[#1]{#2}"&Title of Comment and short title (optional)\\
\verb"\topical[#1]{#2}"&Title of Topical Review and short title 
(optional)\\
\verb"\review[#1]{#2}"&Title of review article and short title (optional)\\
\verb"\note[#1]{#2}"&Title of Note and short title (optional)\\
\verb"\prelim[#1]{#2}"&Title of Preliminary Communication \& short title\\
\verb"\author{#1}"&List of all authors\\
\verb"\article[#1]{#2}{#3}"&Type and title of other articles and 
short title (optional)\\
\verb"\address{#1}"&Address of author\\
\verb"\pacs{#1}"&PACS classification codes\\
\verb"\pacno{#1}"&Single PACS classification code\\
\verb"\ams{#1}"&Mathematics Classification Scheme\\
\verb"\submitto{#1}"&`Submitted to' message\\
\verb"\maketitle"&Creates title page\\
\verb"\begin{abstract}"&Start of abstract\\
\verb"\end{abstract}"&End of abstract\\
\verb"\nosections"&Inserts space before text when no sections\\
\verb"\section{#1}"&Section heading\\
\verb"\subsection{#1}"&Subsection heading\\
\verb"\subsubsection{#1}"&Subsubsection heading\\
\verb"\appendix"&Start of appendixes\\
\verb"\ack"&Acknowledgments heading\\
\verb"\References"&Heading for reference list\\
\verb"\begin{harvard}"&Start of alphabetic reference list\\
\verb"\end{harvard}"&End of alphabetic reference list\\
\verb"\begin{thebibliography}{#1}"&Start of numeric reference list\\
\verb"\end{thebibliography}"&End of numeric reference list\\
\verb"\etal"&\etal for text and reference lists\\
\verb"\nonum"&Unnumbered entry in numerical reference list\\
\br
\end{tabular}
\end{table}

\clearpage

\begin{table}
\caption{Macros defined within {\tt iopart.cls}
for use with figures and tables.}
\begin{indented}
\item[]\begin{tabular}{@{}l*{15}{l}}
\br
Macro name&Purpose\\
\mr
\verb"\Figures"&Heading for list of figure captions\\
\verb"\Figure{#1}"&Figure caption\\
\verb"\Tables"&Heading for tables and table captions\\
\verb"\Table{#1}"&Table caption\\
\verb"\fulltable{#1}"&Table caption for full width table\\
\verb"\endTable"&End of table created with \verb"\Table"\\
\verb"\endfulltab"&End of table created with \verb"\fulltable"\\
\verb"\endtab"&End of table\\
\verb"\br"&Bold rule for tables\\
\verb"\mr"&Medium rule for tables\\
\verb"\ns"&Small negative space for use in table\\
\verb"\centre{#1}{#2}"&Centre heading over columns\\
\verb"\crule{#1}"&Centre rule over columns\\
\verb"\lineup"&Set macros for alignment in columns\\
\verb"\m"&Space equal to width of minus sign\\
\verb"\-"&Left overhanging minus sign\\
\verb"\0"&Space equal to width of a digit\\
\br
\end{tabular}
\end{indented}
\end{table}

\clearpage

\begin{table}[hb]
\caption{\label{jlab2}Abbreviations in {\tt iopart.cls} for journals handled by IOP Publishing.}
\footnotesize
\begin{tabular}{@{}llll}
\br
{\rm Short form of journal title} & Macro & {\rm Short form of journal title} & Macro \\
\mr
2D Mater.&\verb"\TDM"&J. Radiol. Prot.&\verb"\JRP"\\
AJ&\verb"\AJ"&J. Semicond.&\verb"\JOS"\\
ApJ&\verb"\APJ"&J. Stat. Mech.&\verb"\JSTAT"\\
ApJL&\verb"\APJL"&Laser Phys.&\verb"\LP"\\
ApJS&\verb"\APJS"&Laser Phys. Lett.&\verb"\LPL"\\
Adv. Nat. Sci: Nanosci. Nanotechnol.&\verb"\ANSN"&Metrologia&\verb"\MET"\\
Appl. Phys. Express&\verb"\APEX"&Mater. Res. Express&\verb"\MRE"\\
Biofabrication&\verb"\BF"&Meas. Sci. Technol.&\verb"\MST"\\
Bioinspir. Biomim.&\verb"\BB"&Methods Appl. Fluoresc.&\verb"\MAF"\\
Biomed. Mater.&\verb"\BMM"&Modelling Simul. Mater. Sci. Eng.&\verb"\MSMSE"\\
Chin. J. Chem. Phys.&\verb"\CJCP"&Nucl. Fusion&\verb"\NF"\\
Chinese Phys. B&\verb"\CPB"&New J. Phys.&\verb"\NJP"\\
Chinese Phys. C&\verb"\CPC"&Nonlinearity&\verb"\NL"\\
Chinese Phys. Lett.&\verb"\CPL"&Nanotechnology&\verb"\NT"\\
Class. Quantum Grav.&\verb"\CQG"&Phys. Biol.&\verb"\PB"\\
Commun. Theor. Phys.&\verb"\CTP"&Phys. Educ.&\verb"\PED"\\
Comput. Sci. Disc.&\verb"\CSD"&Phys.-Usp.&\verb"\PHU"\\
Environ. Res. Lett.&\verb"\ERL"&Physiol. Meas.&\verb"\PM"\\
EPL&\verb"\EPL"&Phys. Med. Biol.&\verb"\PMB"\\
Eur. J. Phys.&\verb"\EJP"&Phys. Scr.&\verb"\PS"\\
Fluid Dyn. Res.&\verb"\FDR"&Plasma Phys. Control. Fusion&\verb"\PPCF"\\
Inverse Problems&\verb"\IP"&Plasma Sci. Technol.&\verb"\PST"\\
Izv. Math.&\verb"\IZV"&Plasma Sources Sci. Technol.&\verb"\PSST"\\
Jpn. J. Appl. Phys.&\verb"\JJAP"&Quantum Electron.&\verb"\QEL"\\
J. Breath Res.&\verb"\JBR"&Rep. Prog. Phys.&\verb"\RPP"\\
JCAP&\verb"\JCAP"&Res. Astron. Astrophys.&\verb"\RAA"\\
J. Geophys. Eng.&\verb"\JGE"&Russ. Chem. Rev.&\verb"\RCR"\\
JINST&\verb"\JINST"&Russ. Math. Surv.&\verb"\RMS"\\
J. Micromech. Microeng.&\verb"\JMM"&Sb. Math.&\verb"\MSB"\\
J. Neural Eng.&\verb"\JNE"&Science Foundation in China&\verb"\SFC"\\
J. Opt.&\verb"\JOPT"&Sci. Technol. Adv. Mater.&\verb"\STAM"\\
J. Phys. A: Math. Theor.&\verb"\jpa"&Semicond. Sci. Technol.&\verb"\SST"\\
J. Phys. B: At. Mol. Opt. Phys.&\verb"\jpb"&Smart Mater. Struct.&\verb"\SMS"\\
J. Phys: Condens. Matter&\verb"\JPCM"&Supercond. Sci. Technol.&\verb"\SUST"\\
J. Phys. D: Appl. Phys.&\verb"\JPD"&Surf. Topogr.: Metrol. Prop.&\verb"\STMP"\\
J. Phys. G: Nucl. Part. Phys.&\verb"\jpg"&Transl. Mater. Res.&\verb"\TMR"\\
\mr
{\it IOP Conference Series journals}\\
\mr
J. Phys.: Conf. Ser.&\verb"\JPCS"\\
IOP Conf. Ser.: Earth Environ. Sci.&\verb"\EES"\\
IOP Conf. Ser.: Mater. Sci. Eng.&\verb"\MSE"\\

\br
\end{tabular}

\end{table}

\clearpage

\begin{table}

\caption{\label{jlab2b}Abbreviations for IOP Publishing journals that are no longer published.}
\begin{indented}
\item[]\begin{tabular}{@{}lll}
\br
{\rm Short form of journal title} & Macro name & Years relevant\\
\mr
J. Phys. A: Math. Gen.&\verb"\JPA"&1975--2006\\
J. Phys. B: At. Mol. Phys.&\verb"\JPB"&1968--1987\\
J. Phys. C: Solid State Phys.&\verb"\JPC"&1968--1988\\
J. Phys. E: Sci. Instrum.&\verb"\JPE"&1968--1989\\
J. Phys. F: Met. Phys.&\verb"\JPF"&1971--1988\\
J. Phys. G: Nucl. Phys.&\verb"\JPG"&1975--1988\\
Pure Appl. Opt.&\verb"\PAO"&1992--1998\\
Quantum Opt.&\verb"\QO"&1989--1994\\
Quantum Semiclass. Opt.&\verb"\QSO"&1995--1998\\
J. Opt. A: Pure Appl. Opt.&\verb"\JOA"&1999--2009\\
J. Opt. B: Quantum Semiclass. Opt.&\verb"\JOB"&1999--2005\\
\br
\end{tabular}
\end{indented}
\end{table}

\begin{table}[h]

\caption{\label{jlab3}Abbreviations in {\tt iopart.cls} for some 
common journals not handled by IOP Publishing.}
\begin{indented}
\item[]\begin{tabular}{@{}llll}
\br
Short form of journal & Macro & Short form of Journal & Macro\\
\mr
Acta Crystallogr.&\verb"\AC"&J. Quant. Spectrosc. Radiat. Transfer&\verb"\JQSRT"\\
Acta Metall.&\verb"\AM"&Nuovo Cimento&\verb"\NC"\\
Ann. Phys., Lpz&\verb"\AP"&Nucl. Instrum. Methods&\verb"\NIM"\\
Ann. Phys., NY&\verb"\APNY"&Nucl. Phys.&\verb"\NP"\\
Ann. Phys., Paris&\verb"\APP"&Phys. Fluids&\verb"\PF"\\
Can. J. Phys.&\verb"\CJP"&Phys. Lett.&\verb"\PL"\\
Gen. Rel. Grav.&\verb"\GRG"&Phys. Rev.&\verb"\PR"\\
J. Appl. Phys.&\verb"\JAP"&Phys. Rev. Lett.&\verb"\PRL"\\
J. Chem. Phys.&\verb"\JCP"&Proc. R. Soc.&\verb"\PRS"\\
J. High Energy Phys.&\verb"\JHEP"&Phys. Status Solidi&\verb"\PSS"\\
J. Magn. Magn. Mater.&\verb"\JMMM"&Phil. Trans. R. Soc.&\verb"\PTRS"\\
J. Math. Phys.&\verb"\JMP"&Rev. Mod. Phys.&\verb"\RMP"\\
J. Opt. Soc. Am.&\verb"\JOSA"&Rev. Sci. Instrum.&\verb"\RSI"\\
J. Physique&\verb"\JP"&Solid State Commun.&\verb"\SSC"\\
J. Phys. Chem.&\verb"\JPhCh"&Sov. Phys.--JETP&\verb"\SPJ"\\
J. Phys. Soc. Jpn.&\verb"\JPSJ"&Z. Phys.&\verb"\ZP"\\

\br
\end{tabular}
\end{indented}
\end{table}

\clearpage

\section{Including author names using Chinese, Japanese and Korean characters in submissions to IOP Publishing journals}
Authors in all IOP Publishing journals have the option to include names in Chinese, Japanese or Korean (CJK) characters in addition to the English name. The names will be displayed in the print issue and the online PDF, abstract and table of contents, in parentheses after the English name.

It is the decision of the individual authors whether or not to include a CJK version of their names; for a single article it is not necessary for all authors to include a CJK name if only one author wishes to do so. It is the responsibility of the authors to check the accuracy and formatting of the names in the final proofs that they receive prior to publication.

To include names in CJK characters, authors should use the \verb"cjk.sty" package, available from \verb"http://www.ctan.org/tex-archive/language/chinese/CJK/". Users should be aware that this is a very large and complicated package which relies on a large number of fonts.  We recommend using a TeX package that includes this package and all of the fonts by default, so that manual configuration is not required (e.g. the TeXLive distribution, which is available on all platforms (Macintosh, Windows and Linux)).

The documentation for the \verb"cjk.sty" package gives information on how CJK characters can be included in TeX files.  Most authors will find it convenient to include the characters in one of the standard encodings such as UTF-8, GB or JIS, if they have access to a text editor that supports such encodings.

Example TeX coding might be:
\begin{verbatim}
\documentclass[12pt]{iopart}
\usepackage{CJK}
.
.
.
\begin{document}
\begin{CJK*}{GBK}{ }

\title[]{Title of article}
\author{Author Name (CJK characters)}
\address{Department, University, City, Country}
.
.
.
\end{CJK*}
\end{verbatim}

To avoid potential problems in handling the CJK characters in submissions, authors should always include a PDF of the full version of their papers (including all figure files, tables, references etc) with the CJK characters in it.

\end{document}

### 2016-12-19 pasp/paperclimate.tex
dont steal plox

\documentclass[12pt]{iopart}
\newcommand{\gguide}{{\it Preparing graphics for IOP Publishing journals}}
\usepackage{harvard} 
\usepackage{nicefrac}

%Uncomment next line if AMS fonts required
%\usepackage{iopams}  
\begin{document}

\title[Site Location Study for a Millimeter Observatory in the Colombian Andes I: In Situ]{Site location study for a high-mountain millimeter observatory in the Colombian Andes I: In situ data}

\author{Germ¬\'an Chaparro Molano$^{1}$}
\address{$^1$Grupo de Simulaci\'on, An\'alisis y Modelado, Vicerrector\'ia de Investigaci\'on, Universidad ECCI, Bogot\'a, Colombia}
\ead{gchaparrom@ecci.edu.co}
\author{Oscar Leonardo Ram\'irez Su\'arez$^{1}$}
\address{$^1$Grupo de Simulaci\'on, An\'alisis y Modelado, Vicerrector\'ia de Investigaci\'on, Universidad ECCI, Bogot\'a, Colombia}
\ead{oramirezs@ecci.edu.co}
\author{Oscar Restrepo$^{1}$}
\address{$^1$Grupo de Simulaci\'on, An\'alisis y Modelado, Vicerrector\'ia de Investigaci\'on, Universidad ECCI, Bogot\'a, Colombia}
\ead{orestrepog@ecci.edu.co}
\author{Alexander Mart\'inez$^{1,2}$}
\address{$^1$Grupo de Simulaci\'on, An\'alisis y Modelado, Vicerrector\'ia de Investigaci\'on, Universidad ECCI, Bogot\'a, Colombia}
\address{$^2$Instituto de Hidrolog\'ia, Meteorolog\'ia y Estudios Ambientales, Bogot\'a, Colombia}
\vspace{10pt}
\begin{indented}
\item[]December 2016
\end{indented}

\begin{abstract}
This document describes the  preparation of an article using \LaTeXe\ and 
\verb"iopart.cls" (the IOP Publishing \LaTeXe\ preprint class file).
This class file is designed to help 
authors produce preprints in a form suitable for submission to any of the
journals listed in table~\ref{jlab1} on the next page.  You are not obliged to use this class file---we accept
submissions using all common \LaTeX\ class and style files.  The \verb"iopart.cls"
class file is supplied merely as a convenience for those authors who find it useful.
This document gives both general advice that applies whatever class file you use, and specific advice
that applies if you choose to use \verb"iopart.cls".

We also accept submissions in Word format.  Instructions for Word submissions are available via the `Author Guidelines' link at \verb"http://authors.iop.org".

If you have any queries about this document or any aspect of preparing your article for submission please contact us at the e-mail address given above.
\end{abstract}

% Uncomment for PACS numbers
%\pacs{00.00, 20.00, 42.10}
%
% Uncomment for keywords
%\vspace{2pc}
%\noindent{\it Keywords}: XXXXXX, YYYYYYYY, ZZZZZZZZZ
%
% Uncomment for Submitted to journal title message
\submitto{PASP}
%
% Uncomment if a separate title page is required
%\maketitle
% 
% For two-column output uncomment the next line and choose [10pt] rather than [12pt] in the \documentclass declaration
%\ioptwocol
%

\section{Introduction}

The \verb"iopart" \LaTeXe\ article class file is provided to help authors prepare articles for submission to IOP Publishing journals.
  This document gives advice on preparing your submission, and specific instructions on how to use \verb"iopart.cls" to follow this advice.  You \cite{nakano}

\section{Methods}

The meteorological data used in this study was gathered for 30 years (1980-2010) in 2046 Instituto de Hidrolog\'ia, Meteorología y Estudios Ambientales (IDEAM) weather stations across Colombia (Figure \ref{totmap}). The  variables that we considered relevant to our work were: Precipitation (mm/mo), Rainy Days (d/mo),  Relative Humidity (\%), and Sunshine (h/d). These variables are reported as multi-annual monthly data, i.e. each variable is reported monthly from January to December averaged over the 1980-2010 range. For precipitation, each monthly datum corresponds to the cumulative monthly value.  For rainy days, each monthly datum corresponds to the number of rainy days for that month.  For relative humidity, each monthly datum corresponds to the  average daily value averaged over each month. For sunshine, each monthly datum corresponds to the cumulative daily value averaged over each month. All stations reported precipitation values, although only 2002 reported rainy days values, 445 reported relative humidity values, and 336 reported sunshine values. This means that not all available stations registered all meteorological variables relevant to this study. We classify stations according to which variables they measured (Table \ref{tabstype}).\\

\begin{table}
\caption{\label{tabstype}Types of stations classified according to measured variables.}
\begin{indented}
\item[]\begin{tabular}{@{}cccccc}
\br
Type&Precip.&Rainy &Relative&Sunshine&No. of Stations\\
&&Days&Humidity&&\\
\mr
$T_1$&Y&N&N&N&29\\
$T_2$&Y&Y&N&N&1563\\
$T_3$&Y&N&Y&N&1\\
$T_4$&Y&Y&Y&N&117\\
$T_5$&Y&N&N&Y&1\\
$T_6$&Y&Y&N&Y&8\\
$T_7$&Y&N&Y&Y&13\\
$T_8$&Y&Y&Y&Y&314\\
\br
\end{tabular}
\end{indented}
\end{table}

Since the data represent multi-annual seasonal variations, we wish to classify and group together stations that show a similar climatological behavior for each variable, regardless of the station's location, elevation or type. This way, we can relate weather patterns of regions which are not obviously related climatologically but might show a similar behavior, e.g. dry weather patterns in Guajira (a desert region in the north of Colombia) with similar dry conditions present at high-mountain sites in the Andes mountain range.\\

We classified stations using a low-dimensional embedding \cite{embed} for each variable, which ensures that a classification algorithm can uncover features in the data across a number of dimensions that is lower than the native dimensionality of the data (which in this case is monthly, i.e. $N_\mathrm{dim}=12$). This involves a dimensionality reduction algorithm followed by a clustering algorithm. Given that we wish to make a grouping of stations showing similar climatological patterns without making too many assumptions on the data, we decided to use unsupervised learning techniques.  In our case, we chose Principal Component Analysis (PCA) covering 95\% of the variance followed by a Gaussian Mixture Model yielding a low/locally minimum Bayesian Information Criterion (BIC) value.

\subsection{Principal Component Analysis}

We were able to reduce the dimensionality of the data for each variable from  $N_\mathrm{dim}=12$ to  $N_\mathrm{dim}^\mathrm{PCA}\le3$, while ensuring that at least 95\% of the variance of the data is explained by the least amount of components. For precipitation, rainy days and sunshine, $N_\mathrm{dim}^\mathrm{PCA}(95\%)=3$ and for relative humidity $N_\mathrm{dim}^\mathrm{PCA}(95\%)=2$. To do this we used the Python \texttt{sklearn.decomposition.PCA} module \cite{sklearn}.

\subsection{Gaussian Mixture Models}

After projecting the data along the components found by the PCA algorithm, we clustered the data using a Gaussian Mixture Model, which uses an Expectation-Maximization algorithm to find a Maximum Likelihood model made of a given number $N_\mathrm{G-C}$ of Gaussian distributions, each of which is expected to represent a cluster (component) in the data \cite{gmm}. Depending on the restrictions put on the covariance of each Gaussian component, the number of free parameters can go from $N_\mathrm{G-C}(N_\mathrm{dim}+1)$ (spherical covariance) to  $N_\mathrm{G-C}(3N_\mathrm{dim}+1)$ (full covariance). Since the number of components and covariance restriction is given by the user and not \emph{a posteriori} by the algorithm, me need to make sure that the Gaussian components do not over-fit the data, e.g. a model which yields one cluster per datum should be disallowed.\\

In order to achieve this, we compare the Bayesian Information Criterion values \cite{bicref} for a grid of Gaussian components ($N_\mathrm{G-C}<20$) and covariance restriction methods for each variable (Figure \ref{bic}). We selected the models which yielded the lowest BIC value and plotted the climatological clusters in Figures \ref{clusth}-\ref{clustb}. To do this we used the Python \texttt{sklearn.mixture.GMM} module \cite{sklearn}.\\

Table \ref{tabclu} shows the selected number of clusters and covariance method for each variable along with our preferred clusters that correlate with dry, sunny climate. From here on, if a station appears in one of our preferred clusters, we will say that it satisfies our criterion for that specific variable.

\begin{table}
\caption{\label{tabclu}Climate clusters according to a lowest BIC-based selection of Gaussian Mixture Models.}
\begin{indented}
\item[]\begin{tabular}{@{}lclll}
\br
Variable&Symbol&Lowest BIC&Lowest BIC&Preferred Clusters\\
&&No. of Clusters&Cov. Method &\\
\mr
Precipitation&$P$&11&Full&2,3,7,11\\
Rainy Days&$D$&6&Full&1,4,6\\
Relative Humidity&$H$&2&Full&2\\
Sunshine&$S$&5&Tied&1,4,5\\
\br
\end{tabular}
\end{indented}
\end{table}

\section{Quality Index}

We wish to identify all stations for whom all four criteria for identifying a region as having a stable sunny, dry climate can be met.  However, since only a few stations measured all variables, it is not straightforward to accept or reject stations based on their satisfing a criterion, i.e. belonging to one of our preferred clusters. This requires setting up a measure of quality for individual stations depending on their location and the variables they measured. We will describe the selection process and probabilistic analysis that allows us to quantify a station's capability for indicating regions with an unusually dry, sunny climate.\\

We start by creating a shortlist of stations that meet criteria for (and only for) the variables they measured. 665 stations across Colombia showed such behavior. However, since we are interested in high-mountain regions, we included a criterion for elevation limiting our list to 119 stations located above 2000 masl. This shortlist was further reduced by requiring that for a given station, the total average plus one  standard deviation for precipitation, rainy days, and relative humidity correspond to less than 4.3 mm/d, 17 d/mo, and 81\% respectively, and for sunshine, to more than 4 h/d. This left us with 83 weather stations. \\

However, we want to give a measure of the quality of the station vis-a-vis our criteria, i.e. a station that meets less than all four criteria should still be considered in our analysis, but it cannot be given the same importance as a station that meets all four criteria. In order to account for this, and controlling for the fact that stations are not distributed evenly in elevation (Figure \ref{figelev}), we propose a quality measure based on a probabilistic analysis of the data.

\subsection{Probabilistic Analysis}

We base our analysis on the probability for our hypothesis that a station which appears in our shortlist meets all four criteria, given that it is located at a given elevation. We can write down this probability in terms of the following events:

\begin{itemize}
\item $C$ is the event that a given station meets all four criteria, i.e. belongs to our preferred clusters for the four variables relevant to this study.
\begin{itemize}
\item $C$ is actually the conjunction of all four of the $P,D,H,S$ events ``The station belongs to a preferred cluster for variable $X$'', i.e. $C=P\cap R\cap H \cap S$.
\end{itemize}
\item $A$ is the event that a given station appears in our shortlist.
\item $T_i$ is the event that a given station is of type $T_i$ (See Table \ref{tabstype}).
\begin{itemize}
\item For example, if $M_X$ is the event ``the station measured variable $X$'', then the event $T_7$ is equivalent to $M_P\cap \neg M_R\cap M_H \cap M_S$
\end{itemize}
\item $h$ is the event that a given station is located at an elevation $h$.
\end{itemize}

Thus, the probability that a station $j=(0,1,...,N_\mathrm{stations})$ of type $T_i$, meets all four criteria ($C$) and appears in our shortlist ($A$) given that it is located at an elevation $h$, can be written as $P_j=P(C\cap A\cap T_i\ |\ h)$. We cannot  calculate exactly this probability, as it depends on unknown unknowns, i.e. whether a station that did not measure a given variable in reality belongs to one of our preferred climatological clusters for that variable. Thus we will use Bayes' theorem to help estimate this probability using simpler conditional probabilities. First we control for the inhomogeneous distribution of elevations $P(h)$,
\begin{equation}
P(C\cap A\cap T_i\ |\ h)=\frac{P(C\cap A\cap T_i\cap h)}{P(h)}\ .
\end{equation}
The joint probability of all events $C\cap A\cap T_i\cap h$ can be rewritten as,
\begin{equation}
P(C\cap A\cap T_i\ |\ h)=\frac{P(h\ |\ C\cap A\cap T_i)}{P(h)}P(C\cap A\cap T_i)\ .
\end{equation}
If a station satisfies all four criteria, $P(C\cap A\cap T_i)=1$. This probability can be expressed in terms of simpler conditional probabilities,
\begin{equation}
P(C\cap A\cap T_i\ |\ h)=\frac{P(h\ |\ C\cap A\cap T_i)}{P(h)}P(C\ |\ A\cap T_i)P(A\ |\ T_i)P(T_i)\ .
\end{equation}
Here $P(h\ |\ C\cap A\cap T_i$ is the distribution of elevations for a station type ($T_i$) on our shortlist ($C\cap A$), $P(A\ |\ T_i)$ is the probability that a station of type $T_i$ is in our shortlist, $P(T_i)$ is the probability that a station type is $T_i$, and $P(C\ |\ A\cap T_i)$ is the probability that a station meets all four criteria ($C$) given that it is in our shortlist ($A$) and is of type $T_i$. All of these probabilities except for the last one can be computed directly from the data.  The probability $P(C\ |\ A\cap T_i)$ has to be estimated indirectly, as we cannot presume to know under which conditions a station can satisfy all our criteria. We can rewrite this probability as,
\begin{eqnarray}
P(C\ |\ A\cap T_i)&=P(P\cap D\cap H\cap S\ |\ A\cap T_i)\ , \\
\label{pclanti}
&=\frac{P(P\cap D\cap H\cap S\cap A\cap T_i)}{P(A\cap T_i)}\ . 
\end{eqnarray}
In order to estimate $P(C\ |\ A\cap T_i)$ we assume that the probability of a station meeting our criteria for a given number of variables is independent of the station type. To illustrate this, let us assume that a given station is of type 1, so it only measured precipitation and rainy days, i.e.,
\begin{equation}
T_1=M_P\cap M_D\cap\neg M_H\cap\neg M_S\ .
\end{equation}
If this station of type $T_1$ is in our shortlist ($A$), the $P$ and $D$ criteria are met. Thus,
\begin{equation}
P(P\cap D\cap H\cap S\cap A\cap T_1)=P( H\cap S\ |\ A\cap T_1)\ .
\end{equation}
Substituting this into Eq. (\ref{pclanti}),
\begin{equation}
P(C\ |\ A\cap T_1)=\frac{P( H\cap S\ |\ A\cap T_1)}{P(A\cap T_1)}=P(H\cap S\ |\ A\cap T_1)
\end{equation}
The last term in the previous expression is unknown, but we can approximate it assuming, as mentioned above, that it is independent of the station type. Therefore, 
\begin{equation}
P(C\ |\ A\cap T_i)\simeq P(H\cap S)\ .
\end{equation}
The probability on the right hand side of the previous equation can be directly computed from the data, as it only requires counting how many stations meet both $H$ and $S$ criteria. This procedure can be extended to all other station types.\\

Since the probabilities for stations in our shortlist vary by orders of magnitude, for clarity's sake we decided to use a logarithmic quality index for a station in our shortlist. For $j=(0,1,...,N_\mathrm{stations})$,
\begin{equation}
Q_j=9\ \frac{\log (\nicefrac{P_\mathrm{j}}{P_\mathrm{min}})}{\log (\nicefrac{P_\mathrm{max}}{P_\mathrm{min}})}+1\ .
\end{equation}
Here $P_\mathrm{min},P_\mathrm{max}$ are the lowest and highest values for the probabilities for the stations in the first shorlist of 665 stations that satisfy criteria for all measured variables. Thus, if $P_j=P_\mathrm{max}$, $Q_j=10$, and if $P_j=P_\mathrm{min}$, $Q_j=1$.

\section{Results}

We selected clusters of climatologically similar stations across Colombia using unsupervised learning methods for each measured variable. From these clusters, 83 weather stations came up as potential candidates for identifying high-mountain regions (at an elevation greater that 2000 masl) with unusually dry, sunny weather (Figure \ref{mapsl}).\\

Given that many of our candidate stations were lacking in relative humidity and sunshine data, we extrapolated for them using data from nearby stations (at less than 4 km away and at an elevation difference of less than 100 m). We made sure that in general, for stations less than 5 km away, the difference in relative humidity and sunshine data is less than 15\%. The extrapolated data suggested that 4 of the stations in the shortlist did not satisfy the relative humidity and sunshine criteria.  The remaining 73 stations are plotted in Figure \ref{73sl}.\\

All stations outside of the region located within the latitude, longitude range $[(4.3,6.2),(-72.4,-74.5)]$ seem to be outliers, and while they could indicate unusually dry, sunny regions, there are not enough stations nearby to say anything conclusively. Furthermore, those stations only measured one or two variables, which leads us to reject them from our shortlist. \\

The final 70 stations located nearby the Cundinamarca-Boyac\'a region of Colombia are shown in more detail in Figure \ref{boycun}. This Figure shows two tentative locations in the northern region (Boyac\'a department) appear, where the quality index $Q_j$ (proportional to the point size) indicates that all four criteria are satisfied. In order to see if those regions are geographically correlated, we grouped together these stations using a lowest-BIC spherical covariance Gaussian Mixture Model on their coordinates (Figure \ref{biglatlon}).\\

The stations are thus classified, and the colored points in Figure \ref{gmmlatlon} show 6 candidate regions. However, in order to see if the prediction from the Gaussian Mixture Models is indicative of regions with similar (dry, sunny) weather, we plotted the predicted variance regions for each cluster along with other stations in our sample (rejects) in Figure \ref{latlonreg}. From this figure, we can reject regions where the number of rejected stations within the predicted variance regions is similar to or higher than the number of stations in our shortlist for a specified cluster.

\subsection{La Sabana}

The region located to the west of Bogot\'a ()is a hilly region at an elevation above 2600 masl, with a large variety of microclimates, which explains the significant amount of stations from this region in our shortlist. However, Figure \ref{lasabana} shows that within this predicted variance region there are more stations that do not satisfy our criteria than stations that do. Furthermore, stations that meet and do not meet our criteria are homogeneously distributed. This, compounded with the fact that most (70\%) of these stations have a $Q$ value lower than 5.6 due to the lack of humidity/sunshine data, implies that at the moment this is not a strong candidate region for our purposes.
%light blue
\subsection{Valle de Ubat\'e}

This region, located to the NNE of Bogot\'a s a hilly region along a valley at an elevation above 2600 masl. The ratio of number of stations that do not meet our criteria to the stations in this predicted variance region is low (0.3), and almost half of these stations (\%46) have a $Q$ value higher than 6. This indicates a potential candidate region, although more  sunshine and humidity data are needed. In our next paper we will study the amount of atmospheric water vapor near this region.
%cyan

\subsection{Villa de Leyva}

The region near the town of Villa de Leyva is known for its dry weather, and the stations located within this region seem to support this belief. The ratio of stations not in our shortlist vs. the stations in this predicted variance region is not very low (0.4). However, Figure \ref{villadeley} seems to indicate that the actual region of interest is narrower than the predicted variance region, as the stations that do not satisfy our criteria are on the NNW and SSW fringes of said region. The presence of one very high-$Q$ station is very suggestive of this region being a strong candidate. However, its relatively low elevation (near 2200 masl) can signify the presence of too much atmospheric water vapor above the surface. This will be covered in our next paper.
%red

\subsection{Tunja Canton}

The variance predicted region west of Tunja is located in the historic Tunja Canton, and even though the ratio of stations not in our shortlist to stations in this predicted region is low (0.31), there are simply not enough humidity and sunshine data to make any strong conclusion regarding this region. However, the high-elevation location of some of these stations is tantalizing, which is why we will keep this as a region of interest for our next paper.

% dark blue

\subsection{Valle del Sol}

This is one of the most promising regions in our sample. Located in a wide, sunny valley (hence the name) at an elevation of 2600 masl, it is surrounded by mountains, and rains are not as common as elsewhere in the country. This region has the lowest ratio of not in our shortlist stations to stations in this predicted variance region (0.2), $\%40$ of the stations have a $Q$ value higher than 6, and one of the stations has a very high-$Q$ value. This will be one of the regions of interest for our next paper, and we think that it warrants radiometer measurements to be carried out.

\subsection{Pisba National Park}

This is a \emph{p\'aramo} region, characterized by high-mountain tundra weather. The ratio of stations not in our shortlist to station in this predicted variance region is high (0.5), and it is unlikely that the mist allows for a low-atmospheric water vapor region to be located here, even if rain is sporadic. The lack of a high-$Q$ station means that more sunshine-humidity data is needed, but this is not one a strong candidate region.\\

%figures for all these

% tables for our three favorites, Ubat\'e, Villa, Sol.

One, to the NWW of Bogot\'a, another to the N of Bogot\'a, in the region known as Valle de Ubat\'e,

4000 masl might be interesting as well, but there are not enough stations.

 \section*{References}
 \bibliography{paperclimate}
\bibliographystyle{jphysicsB}

  \end{document}

### 2017-03-02 pasp/paperclimate.tex
1st plot
intro

@@
 \usepackage{nicefrac} 
-
+\usepackage{graphicx}
+\usepackage{epstopdf}
 %Uncomment next line if AMS fonts required
@@
 
+% kw: water vapor, atmospheric effects, site testing, absorption, integrated water vapor, radio astronomy, satellite communication, atmospheric propagation
+
 \section{Introduction}
 
-The \verb"iopart" \LaTeXe\ article class file is provided to help authors prepare articles for submission to IOP Publishing journals.
-  This document gives advice on preparing your submission, and specific instructions on how to use \verb"iopart.cls" to follow this advice.  You \cite{nakano}
+Astronomical observations in the millimeter and sub-millimeter wavelength range require that atmospheric effects affecting absorption at these wavelengths are kept to a minimum.
+The main factor contributing to the atmospheric opacity is water vapor, which very efficiently absorbs light in the THz range due to BLAHBLAHBLAH.
+In order to characterize a site according to its atmospheric transparency to THz radiation, it is necessary to retrieve the precipitable water vapor profile either via remote sensing techniques such as microwave radiometry, satellite measurements, or indirectly via models of in situ climatological measurements or GPS-delay studies.
+\\
+
+Most of the precipitable water vapor that can potentially affect the path of a THz photon hoping to get through the atmosphere, actually exists in the troposphere.
+The lower the elevation of a potential site, the longer the path becomes for that photon, increasing its chances of being absorbed by a water vapor molecule.
+For this reason, the geographic characterization of the quality of a site is given in terms of the local water vapor column above the potential site.
+This water vapor column is usually expressed as the amount of precipitable water vapor, given in mm.\\
+
+The water vapor column above a potential site can change rapidly (during the course of the day) or it can change seasonally.
+This is why medium-to-long term monitoring of the atmosphere above a site is required before building an expensive mm/sub-mm wave radio telescope.\\
+
+You \cite{embed}
 
@@
 We classify stations according to which variables they measured (Table \ref{tabstype}).\\
+
+\begin{figure}
+\begin{center}
+\includegraphics[scale=0.8]{totmap.eps}
+\caption{Say What}
+\end{center}
+\end{figure}
 

### 2017-03-02 pasp/paperclimate.tex
intro

@@
 
-Astronomical observations in the millimeter and sub-millimeter wavelength range require that atmospheric effects affecting absorption at these wavelengths are kept to a minimum.
+Astronomical observations in the millimeter and sub-millimeter wavelength range require that atmospheric effects affecting absorption at these wavelengths are kept to a minimum \cite{chajpwv,radford2016}.
 The main factor contributing to the atmospheric opacity is water vapor, which very efficiently absorbs light in the THz range due to BLAHBLAHBLAH.
-In order to characterize a site according to its atmospheric transparency to THz radiation, it is necessary to retrieve the precipitable water vapor profile either via remote sensing techniques such as microwave radiometry, satellite measurements, or indirectly via models of in situ climatological measurements or GPS-delay studies.
+In order to characterize a site according to its atmospheric transparency to THz radiation, it is necessary to retrieve the precipitable water vapor profile either via remote sensing techniques such as microwave radiometry \cite{radiometer}, satellite measurements, or indirectly via models of in situ climatological measurements or GPS-delay studies.
 \\
 
-Most of the precipitable water vapor that can potentially affect the path of a THz photon hoping to get through the atmosphere, actually exists in the troposphere.
+Most of the precipitable water vapor that can potentially affect the path of a THz photon hoping to get through the atmosphere actually exists in the troposphere.
 The lower the elevation of a potential site, the longer the path becomes for that photon, increasing its chances of being absorbed by a water vapor molecule.

### 2017-03-03 pasp/paperclimate.tex
some mods

@@
 
+Besides astronomical observations, satellite communications in the THz frequency range are a growing field, given that the ``classical radio or whatevs''
+
 Astronomical observations in the millimeter and sub-millimeter wavelength range require that atmospheric effects affecting absorption at these wavelengths are kept to a minimum \cite{chajpwv,radford2016}.
 The main factor contributing to the atmospheric opacity is water vapor, which very efficiently absorbs light in the THz range due to BLAHBLAHBLAH.
-In order to characterize a site according to its atmospheric transparency to THz radiation, it is necessary to retrieve the precipitable water vapor profile either via remote sensing techniques such as microwave radiometry \cite{radiometer}, satellite measurements, or indirectly via models of in situ climatological measurements or GPS-delay studies.
+In order to characterize a site according to its atmospheric transparency to THz radiation, it is necessary to retrieve the precipitable water vapor profile either via remote sensing techniques such as microwave radiometry \cite{radiometer}, satellite measurements \cite{spaceradio}, or indirectly via models of in situ climatological measurements or GPS-delay studies.
 \\
@@
 
+ %c y después ok
+
  \section*{References}

### 2017-03-24 pasp/paperclimate.tex
99% intro

--- 2017-03-02 pasp/paperclimate.tex
@@
 
-Astronomical observations in the millimeter and sub-millimeter wavelength range require that atmospheric effects affecting absorption at these wavelengths are kept to a minimum \cite{chajpwv,radford2016}.
-The main factor contributing to the atmospheric opacity is water vapor, which very efficiently absorbs light in the THz range due to BLAHBLAHBLAH.
-In order to characterize a site according to its atmospheric transparency to THz radiation, it is necessary to retrieve the precipitable water vapor profile either via remote sensing techniques such as microwave radiometry \cite{radiometer}, satellite measurements, or indirectly via models of in situ climatological measurements or GPS-delay studies.
+Besides astronomical observations, satellite communications technology in the THz frequency range is a rapidly growing field \cite{newera,suen2016}.
+Atmospheric models for absorption of THz photons \cite{rosenkranz,lababs} and artifact removal models \cite{removal} have been developed recently, specially considering that the ``classical radio or whatevs'' BLAHBLAH.
+Also Colombia and previous work \cite{pinzon}.
+\\
+
+Astronomical observations in the millimeter and sub-millimeter wavelength range require that atmospheric effects affecting absorption at these wavelengths are kept to a minimum \cite{southpole,radford2016,chajpwv}.
+The main factor contributing to the atmospheric opacity is water vapor, which very efficiently absorbs light in the THz range due to BLAHBLAHBLAH \cite{pardocerni,pardocerni2,cont350}.
+In order to characterize a site according to its atmospheric transparency to THz radiation, it is necessary to retrieve the precipitable water vapor profile either via remote sensing techniques such as microwave radiometry \cite{radiometer,paine2000fourier,southpole2}, satellite measurements \cite{aqua,spacemols,suen2014,spaceradio}, radiosonde humidity measurements \cite{radiosonde,radiosonde2} or indirectly via models of in situ climatological measurements \cite{climatology} or GPS-delay studies \cite{gpsorig,gpsmap}.
 \\
@@
 Most of the precipitable water vapor that can potentially affect the path of a THz photon hoping to get through the atmosphere actually exists in the troposphere.
-The lower the elevation of a potential site, the longer the path becomes for that photon, increasing its chances of being absorbed by a water vapor molecule.
-For this reason, the geographic characterization of the quality of a site is given in terms of the local water vapor column above the potential site.
+The lower the elevation of a potential site, the longer the path becomes for that photon, increasing its chances of being absorbed by a water vapor molecule \cite{liebe1989mpm,linecont,lababs}, either by line absorbtion \cite{submmlines} or by the continuum \cite{turner2009}.
+For this reason, the characterization of the quality of a site is given in terms of the local water vapor column above the potential site \cite{mkradio,southpole3,he2012,atacama}.
 This water vapor column is usually expressed as the amount of precipitable water vapor, given in mm.\\
 
-The water vapor column above a potential site can change rapidly (during the course of the day) or it can change seasonally.
-This is why medium-to-long term monitoring of the atmosphere above a site is required before building an expensive mm/sub-mm wave radio telescope.\\
+The water vapor column above a potential site can change rapidly (during the course of the day) or it can change seasonally \cite{arm2013,caumont2016}.
+This is why medium-to-long term monitoring of the atmosphere above a site is required before building an expensive mm/sub-mm wave radio telescope.
+However, even in best-case scenarios, earth-based telescopes will always be limited by atmospheric absorption of light specially at THz frequencies, where windows of observations are few and in some cases, narrow \cite{arrlim,limits}.
+\\
 
-You \cite{embed}
+Radiometers are complex and expensive \cite{radiopro,receiver}.
+We are poor and for this reason we did what we did BLAHBLAH.
+In an upcoming paper we analize satellite data in order to retrieve seasonal precipitable water vapor measurements for the regions this initial study suggests.
 
@@
 
-We classified stations using a low-dimensional embedding \cite{embed} for each variable, which ensures that a classification algorithm can uncover features in the data across a number of dimensions that is lower than the native dimensionality of the data (which in this case is monthly, i.e.
+We classified stations using a low-dimensional embedding of the data \cite{embed} for each variable.
+This ensures that a classification algorithm can uncover features in the data across a number of dimensions that is lower than the native dimensionality of the data (which in this case is monthly, i.e.
 $N_\mathrm{dim}=12$).
@@
 
-One, to the NWW of Bogot\'a, another to the N of Bogot\'a, in the region known as Valle de Ubat\'e,
+One, to the NWW of Bogot\'a, another to the N of Bogot\'a, in the region known as Valle de Ubat\'e, \\
 
-4000 masl might be interesting as well, but there are not enough stations.
+4000 masl might be interesting as well, but there are not enough stations.\\
+
+Seasonality plots\\
 

### 2017-04-20 pasp/paperclimate.tex
more plots

@@
 \title[Site Location Study for a Millimeter Observatory in the Colombian Andes I: In Situ]{Site location study for a high-mountain millimeter observatory in the Colombian Andes I: In situ data}
+
+% Low Dimensional Embedding of Climate Data for Radio Astronomical Site Testing in the Colombian Andes
 
@@
 %\vspace{2pc}
-%\noindent{\it Keywords}: XXXXXX, YYYYYYYY, ZZZZZZZZZ
+\noindent{\it Keywords}: water vapor, atmospheric effects, site testing, absorption, integrated water vapor, radio astronomy, satellite communication, atmospheric propagation\\
 %
@@
 
-% kw: water vapor, atmospheric effects, site testing, absorption, integrated water vapor, radio astronomy, satellite communication, atmospheric propagation
+% kw:
 
@@
 
-Besides astronomical observations, satellite communications technology in the THz frequency range is a rapidly growing field \cite{newera,suen2016}.
-Atmospheric models for absorption of THz photons \cite{rosenkranz,lababs} and artifact removal models \cite{removal} have been developed recently, specially considering that the ``classical radio or whatevs'' BLAHBLAH.
-Also Colombia and previous work \cite{pinzon}.
+The development of astronomical instrumentation technology in the 0.2-2 THz range has been rapidly growing in recentl years along with THz satellite communications technology \cite{newera} specially considering that the ``classical radio window'' in telecommunications has become saturated.
+Atmospheric models for absorption of THz photons \cite{rosenkranz,lababs} and artifact removal models \cite{removal} have been developed recently.
+Near-equatorial countries face a challenge in using this frequency band for astronomical observations and telecommunications using terabit satellite links  \cite{suen2016} due to the presence of a tropical belt of dense water vapor which efficiently absorbs THz radiation \cite{tropicalbelt}.
+Previous studies of astronomical site testing in the visible range have yielded that nighttime humidity conditions across the Northern Andes are only suitable for educational observatories \cite{pinzon}.
+However, considering that millimeter/sub-millimeter observations need not be done during the night, unusually dry, clear-daytime-sky, high altitude regions in the Northern Andes could be suitable for a world-class millimeter-wave observatory.
 \\
@@
 Astronomical observations in the millimeter and sub-millimeter wavelength range require that atmospheric effects affecting absorption at these wavelengths are kept to a minimum \cite{southpole,radford2016,chajpwv}.
-The main factor contributing to the atmospheric opacity is water vapor, which very efficiently absorbs light in the THz range due to BLAHBLAHBLAH \cite{pardocerni,pardocerni2,cont350}.
+The main factor contributing to the atmospheric opacity is water vapor, which very efficiently absorbs light in the THz range \cite{pardocerni,pardocerni2,cont350} due to a continuum absorption spectrum formed by collisionally broadened absorption lines of water vapor in this frequency range  \cite{linecont,submmlines,turner2009}.
 In order to characterize a site according to its atmospheric transparency to THz radiation, it is necessary to retrieve the precipitable water vapor profile either via remote sensing techniques such as microwave radiometry \cite{radiometer,paine2000fourier,southpole2}, satellite measurements \cite{aqua,spacemols,suen2014,spaceradio}, radiosonde humidity measurements \cite{radiosonde,radiosonde2} or indirectly via models of in situ climatological measurements \cite{climatology} or GPS-delay studies \cite{gpsorig,gpsmap}.
@@
 Most of the precipitable water vapor that can potentially affect the path of a THz photon hoping to get through the atmosphere actually exists in the troposphere.
-The lower the elevation of a potential site, the longer the path becomes for that photon, increasing its chances of being absorbed by a water vapor molecule \cite{liebe1989mpm,linecont,lababs}, either by line absorbtion \cite{submmlines} or by the continuum \cite{turner2009}.
-For this reason, the characterization of the quality of a site is given in terms of the local water vapor column above the potential site \cite{mkradio,southpole3,he2012,atacama}.
+The lower the elevation of a potential site, the longer the path becomes for that photon, increasing its chances of being absorbed by a water vapor molecule \cite{liebe1989mpm,linecont,lababs}.
+For this reason, the quality of a site is determined by the atmospheric water vapor column above the potential site \cite{mkradio,southpole3,he2012,atacama}.
 This water vapor column is usually expressed as the amount of precipitable water vapor, given in mm.\\
@@
 
-Radiometers are complex and expensive \cite{radiopro,receiver}.
-We are poor and for this reason we did what we did BLAHBLAH.
-In an upcoming paper we analize satellite data in order to retrieve seasonal precipitable water vapor measurements for the regions this initial study suggests.
+Although radiometer measurements are desirable for directly measuring the atmospheric water vapor column, their development and/or deployment can be complex and expensive \cite{radiopro,receiver}.
+For this reason we screened historic climate data for evidence of local long-term clear sky, low humidity conditions as proxies for a locally low precipitable water vapor column.
+By low-dimensionally embedding the data, we are able to reduce the dimensionality of multi-annual monthly data from $N_\mathrm{dim}=12$ to $N_\mathrm{dim}\le3$.
+Thus, we can correlate unusually dry regions regardless of their geographical location.
+Given that the climatological variables were sparsely measured by weather stations across Colombia, we evaluated the aptness of a given location using a Bayesian probability-derived quality index.
+Finally, we geographically clustered our candidate locations in order to identify regions of interest.
+We compared our regions of interest with regions with unusually low precipitable water vapor in low-resolution water vapor satellite maps \cite{suen2016} and find a possibly significative geographical correlation.
+In an upcoming paper we will analize satellite data in order to retrieve high-resolution seasonal precipitable water vapor maps for our regions of interest.\\
 
-\section{Methods}
+This paper is organized as follows.
 
-The meteorological data used in this study was gathered for 30 years (1980-2010) in 2046 Instituto de Hidrolog\'ia, Meteorología y Estudios Ambientales (IDEAM) weather stations across Colombia (Figure \ref{totmap}).
-The  variables that we considered relevant to our work were: Precipitation (mm/mo), Rainy Days (d/mo),  Relative Humidity (\%), and Sunshine (h/d).
+\section{The Dataset}
+
+The meteorological data used in this study were gathered over 30 years (1980-2010) in 2046 weather stations monitored by the Instituto de Hidrolog\'ia, Meteorolog\'ia y Estudios Ambientales (IDEAM) in Colombia (Figure \ref{totmap}).
+The  variables that we considered relevant to our work were: Precipitation (mm/mo), Rain Days (d/mo),  Relative Humidity (\%), and Sunshine Duration (h/d).
 These variables are reported as multi-annual monthly data, i.e.
-each variable is reported monthly from January to December averaged over the 1980-2010 range.
+each variable is reported monthly from January to December averaged over the 30 year range.
 For precipitation, each monthly datum corresponds to the cumulative monthly value.
-For rainy days, each monthly datum corresponds to the number of rainy days for that month.
+For rain days, each monthly datum corresponds to the number of rain days for that month.
 For relative humidity, each monthly datum corresponds to the  average daily value averaged over each month.
-For sunshine, each monthly datum corresponds to the cumulative daily value averaged over each month.
-All stations reported precipitation values, although only 2002 reported rainy days values, 445 reported relative humidity values, and 336 reported sunshine values.
-This means that not all available stations registered all meteorological variables relevant to this study.
-We classify stations according to which variables they measured (Table \ref{tabstype}).\\
+For sunshine duration, each monthly datum corresponds to the daily value averaged over each month.
+All stations reported precipitation values, although only 2002 reported rain days, 445 reported relative humidity, and 336 reported sunshine duration.
+This means that not all stations registered all the meteorological variables relevant to this work.
+Thus, we classify stations according to which variables they measured (Table \ref{tabstype}).\\
 
@@
 \includegraphics[scale=0.8]{totmap.eps}
-\caption{Say What}
+\caption{1$\sigma$ (solid) and 2$\sigma$ (dashed) countours of a gaussian kernel density estimation of the geographical distribution of weather stations (small green points) from the IDEAM 1980-2010 database.
+The estimation was done using the Python \texttt{scipy.stats.gaussian\_kde}  function.}\label{totmap}
 \end{center}
@@
 \begin{table}
-\caption{\label{tabstype}Types of stations classified according to measured variables.}
+\caption{\label{tabstype}Types of stations classified according to its measured variables.}
 \begin{indented}
@@
 \br
-Type&Precip.&Rainy &Relative&Sunshine&No. of Stations\\
-&&Days&Humidity&&\\
+Type&Precip.&Rain &Relative&Sunshine&No. of Stations\\
+&&Days&Humidity&Duration&\\
 \mr
@@
 
-Since the data represent multi-annual seasonal variations, we wish to classify and group together stations that show a similar climatological behavior for each variable, regardless of the station's location, elevation or type.
-This way, we can relate weather patterns of regions which are not obviously related climatologically but might show a similar behavior, e.g.
-dry weather patterns in Guajira (a desert region in the north of Colombia) with similar dry conditions present at high-mountain sites in the Andes mountain range.\\
+\section{Low-Dimensional Embedding of Climatological Data}
+
+Our goal is to classify and group together stations that show a similar climatological behavior for each variable, regardless of the station's location, elevation or type.
+Since climate data track multi-annual seasonal variations, we can correlate weather patterns of regions which are not obviously related climatologically but might show a similar behavior, e.g.
+dry weather patterns in Guajira (a desert region in the north of Colombia) with similar dry conditions present at high-mountain sites in the Andes.\\
 
 We classified stations using a low-dimensional embedding of the data \cite{embed} for each variable.
-This ensures that a classification algorithm can uncover features in the data across a number of dimensions that is lower than the native dimensionality of the data (which in this case is monthly, i.e.
+This ensures that a classification algorithm can uncover features in the data even even when the data are projected across a number of dimensions that is lower than the native dimensionality of the data (which in this case is monthly, i.e.
 $N_\mathrm{dim}=12$).
-This involves a dimensionality reduction algorithm followed by a clustering algorithm.
+This involves applying a dimensionality reduction algorithm followed by a clustering algorithm.
 Given that we wish to make a grouping of stations showing similar climatological patterns without making too many assumptions on the data, we decided to use unsupervised learning techniques.
-In our case, we chose Principal Component Analysis (PCA) covering 95\% of the variance followed by a Gaussian Mixture Model yielding a low/locally minimum Bayesian Information Criterion (BIC) value.
+In our case, we chose Principal Component Analysis (PCA) covering 95\% of the variance followed by a Gaussian Mixture Model selected by a low/locally minimum Bayesian Information Criterion (BIC).
 
@@
 We were able to reduce the dimensionality of the data for each variable from  $N_\mathrm{dim}=12$ to  $N_\mathrm{dim}^\mathrm{PCA}\le3$, while ensuring that at least 95\% of the variance of the data is explained by the least amount of components.
-For precipitation, rainy days and sunshine, $N_\mathrm{dim}^\mathrm{PCA}(95\%)=3$ and for relative humidity $N_\mathrm{dim}^\mathrm{PCA}(95\%)=2$.
+For precipitation, rain days and sunshine duration, $N_\mathrm{dim}^\mathrm{PCA}(95\%)=3$ and for relative humidity $N_\mathrm{dim}^\mathrm{PCA}(95\%)=2$.
 To do this we used the Python \texttt{sklearn.decomposition.PCA} module \cite{sklearn}.
+Figure \ref{pcah} shows the relative humidity data projected across 2 principal components explaining 95\% of the variance, thus preserving most of the original structure of the data.
+
+\begin{figure}
+\begin{center}
+\includegraphics[scale=0.8]{pcah.eps}
+\caption{Relative humidity data projected across 2 principal components. The dimensionality of the data has been reduced from 12 to 2 while covering 95\% of the variance of the data. }\label{pcah}
+\end{center}
+\end{figure}
 
@@
 
-After projecting the data along the components found by the PCA algorithm, we clustered the data using a Gaussian Mixture Model, which uses an Expectation-Maximization algorithm to find a Maximum Likelihood model made of a given number $N_\mathrm{G-C}$ of Gaussian distributions, each of which is expected to represent a cluster (component) in the data \cite{gmm}.
-Depending on the restrictions put on the covariance of each Gaussian component, the number of free parameters can go from $N_\mathrm{G-C}(N_\mathrm{dim}+1)$ (spherical covariance) to  $N_\mathrm{G-C}(3N_\mathrm{dim}+1)$ (full covariance).
+After projecting the data along the components found by the PCA algorithm, we clustered the dimensionally-reduced data using a Gaussian Mixture Model, which uses an Expectation-Maximization algorithm to find a Maximum Likelihood model composed of a number $N_\mathrm{G-C}$ of Gaussian distributions, each of which represents a cluster in the data \cite{gmm}.
+Depending on the restrictions put on the covariance of each Gaussian distribution across the data space, the number of free parameters can go from $N_\mathrm{G-C}(N_\mathrm{dim}+1)$ (spherical covariance) to  $N_\mathrm{G-C}(3N_\mathrm{dim}+1)$ (full covariance).
 Since the number of components and covariance restriction is given by the user and not \emph{a posteriori} by the algorithm, me need to make sure that the Gaussian components do not over-fit the data, e.g.
@@
 
-In order to achieve this, we compare the Bayesian Information Criterion values \cite{bicref} for a grid of Gaussian components ($N_\mathrm{G-C}<20$) and covariance restriction methods for each variable (Figure \ref{bic}).
+In order to achieve this, we compare the Bayesian Information Criterion \cite{bicref} for a grid of Gaussian clusters ($N_\mathrm{G-C}<20$) and covariance restriction methods for each variable (Figure \ref{bic}).
 We selected the models which yielded the lowest BIC value and plotted the climatological clusters in Figures \ref{clusth}-\ref{clustb}.
 To do this we used the Python \texttt{sklearn.mixture.GMM} module \cite{sklearn}.\\
+
+\begin{figure}
+\begin{center}
+\includegraphics[scale=0.5]{bich.eps}
+\includegraphics[scale=0.5]{bicl.eps}\\
+\includegraphics[scale=0.5]{bicd.eps}
+\includegraphics[scale=0.5]{bicb.eps}
+\caption{Bayesian Information Criterion (BIC) results for Gaussian Mixture Models using a different number of clusters and using different covariance restriction methods applied on each measured variable dataset.}\label{bic}
+\end{center}
+\end{figure}
 
@@
 Precipitation&$P$&11&Full&2,3,7,11\\
-Rainy Days&$D$&6&Full&1,4,6\\
+Rain Days&$D$&6&Full&1,4,6\\
 Relative Humidity&$H$&2&Full&2\\
-Sunshine&$S$&5&Tied&1,4,5\\
+Sunshine Duration&$S$&5&Tied&1,4,5\\
 \br
@@
 
-\section{Quality Index}
+\begin{figure}
+\begin{center}
+\includegraphics[scale=0.5]{ldeh.eps}
+\includegraphics[scale=0.5]{ldel.eps}
+\includegraphics[scale=0.5]{lded.eps}
+\includegraphics[scale=0.5]{ldeb.eps}
+\caption{Lowest BIC Gaussian Mixture Model clustering of reduced dimensionality climate data. Colors indicate the clustering classification.}\label{lde}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[scale=0.5]{gmmh0.eps}
+\includegraphics[scale=0.5]{gmmh1.eps}
+\caption{Climate clusters as predicted by the lowest-BIC Gaussian Mixture Model for relative humidity.
+Left column: Average monthly relative humidity and standard deviation (error bars) for each climate cluster.
+Right column: Geographical location and altitude of weather stations in each relative humidity cluster.}\label{clusth}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[scale=0.5]{gmml0.eps}
+\includegraphics[scale=0.5]{gmml1.eps}
+\includegraphics[scale=0.5]{gmml2.eps}
+\includegraphics[scale=0.5]{gmml3.eps}
+\includegraphics[scale=0.5]{gmml4.eps}
+\includegraphics[scale=0.5]{gmml5.eps}
+\includegraphics[scale=0.5]{gmml6.eps}
+\includegraphics[scale=0.5]{gmml7.eps}
+\includegraphics[scale=0.5]{gmml8.eps}
+\includegraphics[scale=0.5]{gmml9.eps}
+\includegraphics[scale=0.5]{gmml10.eps}
+\caption{Climate clusters as predicted by the lowest-BIC Gaussian Mixture Model for precipitation.
+Left column: Average monthly precipitation and standard deviation (error bars) for each climate cluster.
+Right column: Geographical location and altitude of weather stations in each precipitation cluster.}\label{clustl}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[scale=0.5]{gmmd0.eps}
+\includegraphics[scale=0.5]{gmmd1.eps}
+\includegraphics[scale=0.5]{gmmd2.eps}
+\includegraphics[scale=0.5]{gmmd3.eps}
+\includegraphics[scale=0.5]{gmmd4.eps}
+\includegraphics[scale=0.5]{gmmd5.eps}
+\caption{Climate clusters as predicted by the lowest-BIC Gaussian Mixture Model for rain days.
+Left column: Average monthly rain days and standard deviation (error bars) for each climate cluster.
+Right column: Geographical location and altitude of weather stations in each rain days cluster.}\label{clustl}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[scale=0.5]{gmmb0.eps}
+\includegraphics[scale=0.5]{gmmb1.eps}
+\includegraphics[scale=0.5]{gmmb2.eps}
+\includegraphics[scale=0.5]{gmmb3.eps}
+\includegraphics[scale=0.5]{gmmb4.eps}
+\caption{Climate clusters as predicted by the lowest-BIC Gaussian Mixture Model for sunshine duration.
+Left column: Average monthly sunshine duration and standard deviation (error bars) for each climate cluster.
+Right column: Geographical location and altitude of weather stations in each sunshine duration cluster.}\label{clustl}
+\end{center}
+\end{figure}
+
+\section{Quality Index for Regions of Interest}
 
@@
 However, since we are interested in high-mountain regions, we included a criterion for elevation limiting our list to 119 stations located above 2000 masl.
-This shortlist was further reduced by requiring that for a given station, the total average plus one  standard deviation for precipitation, rainy days, and relative humidity correspond to less than 4.3 mm/d, 17 d/mo, and 81\% respectively, and for sunshine, to more than 4 h/d.
+This shortlist was further reduced by requiring that for a given station, the total average plus one  standard deviation for precipitation, rain days, and relative humidity correspond to less than 4.3 mm/d, 17 d/mo, and 81\% respectively, and for sunshine duration, to more than 4 h/d.
 This left us with 83 weather stations.
@@
 In order to estimate $P(C\ |\ A\cap T_i)$ we assume that the probability of a station meeting our criteria for a given number of variables is independent of the station type.
-To illustrate this, let us assume that a given station is of type 1, so it only measured precipitation and rainy days, i.e.,
+To illustrate this, let us assume that a given station is of type 1, so it only measured precipitation and rain days, i.e.,
 \begin{equation}
@@
 
-We selected clusters of climatologically similar stations across Colombia using unsupervised learning methods for each measured variable.
-From these clusters, 83 weather stations came up as potential candidates for identifying high-mountain regions (at an elevation greater that 2000 masl) with unusually dry, sunny weather (Figure \ref{mapsl}).\\
+We selected clusters of climatologically similar stations across Colombia using an unsupervised learning low-dimensional embedding algorithm for each measured variable.
+From these clusters, 83 weather stations came up as potential candidates for identifying high-mountain sites (at an elevation greater that 2000 masl) with an unusually dry, sunny weather (Figure \ref{mapsl}).\\
 
-Given that many of our candidate stations were lacking in relative humidity and sunshine data, we extrapolated for them using data from nearby stations (at less than 4 km away and at an elevation difference of less than 100 m).
-We made sure that in general, for stations less than 5 km away, the difference in relative humidity and sunshine data is less than 15\%.
-The extrapolated data suggested that 4 of the stations in the shortlist did not satisfy the relative humidity and sunshine criteria.
+Given that many of our candidate stations were lacking in relative humidity and sunshine duration data, we extrapolated using data from nearby stations (at less than 4 km away and at an elevation difference of less than 100 m).
+We made sure that in general, for stations less than 5 km away, the difference in relative humidity and sunshine duration data are less than 15\%.
+The extrapolated data suggested that 4 of the stations in the shortlist did not satisfy the relative humidity and sunshine duration criteria.
 The remaining 73 stations are plotted in Figure \ref{73sl}.\\
@@
 Furthermore, stations that meet and do not meet our criteria are homogeneously distributed.
-This, compounded with the fact that most (70\%) of these stations have a $Q$ value lower than 5.6 due to the lack of humidity/sunshine data, implies that at the moment this is not a strong candidate region for our purposes.
+This, compounded with the fact that most (70\%) of these stations have a $Q$ value lower than 5.6 due to the lack of humidity/sunshine data, implies that at the moment this is not a strong candidate region of interest.
 %light blue
@@
 The ratio of stations not in our shortlist to station in this predicted variance region is high (0.5), and it is unlikely that the mist allows for a low-atmospheric water vapor region to be located here, even if rain is sporadic.
-The lack of a high-$Q$ station means that more sunshine-humidity data is needed, but this is not one a strong candidate region.\\
+The lack of a high-$Q$ station means that more sunshine-humidity data are needed, but this is possibly not a region of interest.\\
 

### 2017-04-20 pasp/paperclimate.tex
more plots in paper

@@
 
-% Low Dimensional Embedding of Climate Data for Radio Astronomical Site Testing in the Colombian Andes
+% low dimensional embedding of climatological data for astronomical site testing in the Colombian andes
 
@@
 
-The development of astronomical instrumentation technology in the 0.2-2 THz range has been rapidly growing in recentl years along with THz satellite communications technology \cite{newera} specially considering that the ``classical radio window'' in telecommunications has become saturated.
-Atmospheric models for absorption of THz photons \cite{rosenkranz,lababs} and artifact removal models \cite{removal} have been developed recently.
-Near-equatorial countries face a challenge in using this frequency band for astronomical observations and telecommunications using terabit satellite links  \cite{suen2016} due to the presence of a tropical belt of dense water vapor which efficiently absorbs THz radiation \cite{tropicalbelt}.
-Previous studies of astronomical site testing in the visible range have yielded that nighttime humidity conditions across the Northern Andes are only suitable for educational observatories \cite{pinzon}.
+Besides astronomical observations, satellite communications technology in the THz frequency range is a rapidly growing field \cite{newera}.
+Atmospheric models for absorption of THz photons \cite{rosenkranz,lababs} and artifact removal models \cite{removal} have been developed recently, specially considering that the ``classical radio window'' in telecommunications has become saturated, and terabit satellite links could be used in the near future \cite{suen2016}.
+Near-equatorial countries face a challenge in using this frequency band for satellite telecommunications and astronomical observations due to the presence of a tropical belt of dense water vapor, which efficiently absorbs THz radiation \cite{tropicalbelt}.
+Previous studies of astronomical site testing in the visible range have yielded that nighttime humidity conditions across the northern Andes are only suitable for educational observatories \cite{pinzon}.
 However, considering that millimeter/sub-millimeter observations need not be done during the night, unusually dry, clear-daytime-sky, high altitude regions in the Northern Andes could be suitable for a world-class millimeter-wave observatory.
@@
 Astronomical observations in the millimeter and sub-millimeter wavelength range require that atmospheric effects affecting absorption at these wavelengths are kept to a minimum \cite{southpole,radford2016,chajpwv}.
-The main factor contributing to the atmospheric opacity is water vapor, which very efficiently absorbs light in the THz range \cite{pardocerni,pardocerni2,cont350} due to a continuum absorption spectrum formed by collisionally broadened absorption lines of water vapor in this frequency range  \cite{linecont,submmlines,turner2009}.
+The main factor contributing to the atmospheric opacity is water vapor, which very efficiently absorbs light in the THz range \cite{pardocerni,pardocerni2,cont350} due to a continuum absorption spectrum formed by collisionally broadened absorption lines of water vapor in this frequency range  \cite{linecont}.
 In order to characterize a site according to its atmospheric transparency to THz radiation, it is necessary to retrieve the precipitable water vapor profile either via remote sensing techniques such as microwave radiometry \cite{radiometer,paine2000fourier,southpole2}, satellite measurements \cite{aqua,spacemols,suen2014,spaceradio}, radiosonde humidity measurements \cite{radiosonde,radiosonde2} or indirectly via models of in situ climatological measurements \cite{climatology} or GPS-delay studies \cite{gpsorig,gpsmap}.
@@
 Most of the precipitable water vapor that can potentially affect the path of a THz photon hoping to get through the atmosphere actually exists in the troposphere.
-The lower the elevation of a potential site, the longer the path becomes for that photon, increasing its chances of being absorbed by a water vapor molecule \cite{liebe1989mpm,linecont,lababs}.
-For this reason, the quality of a site is determined by the atmospheric water vapor column above the potential site \cite{mkradio,southpole3,he2012,atacama}.
+The lower the elevation of a potential site, the longer the path becomes for that photon, increasing its chances of being absorbed by a water vapor molecule \cite{liebe1989mpm,linecont,lababs}, either by line absorbtion \cite{submmlines} or by the continuum \cite{turner2009}.
+For this reason, the characterization of the quality of a site is given in terms of the local water vapor column above the potential site \cite{mkradio,southpole3,he2012,atacama}.
 This water vapor column is usually expressed as the amount of precipitable water vapor, given in mm.\\
@@
 
-Although radiometer measurements are desirable for directly measuring the atmospheric water vapor column, their development and/or deployment can be complex and expensive \cite{radiopro,receiver}.
-For this reason we screened historic climate data for evidence of local long-term clear sky, low humidity conditions as proxies for a locally low precipitable water vapor column.
-By low-dimensionally embedding the data, we are able to reduce the dimensionality of multi-annual monthly data from $N_\mathrm{dim}=12$ to $N_\mathrm{dim}\le3$.
-Thus, we can correlate unusually dry regions regardless of their geographical location.
+Although radiometer measurements are desirable for directly measuring atmospheric water vapor column, they are complex and expensive \cite{radiopro,receiver}.
+For this reason, we decided to screen climatological data for measurements indicative of local long-term clear sky, low humidity conditions as proxies for a locally low precipitable water vapor column.
+By low-dimensionally embedding the data, we are able to correlate unusually dry regions regardless of their geographical location.
 Given that the climatological variables were sparsely measured by weather stations across Colombia, we evaluated the aptness of a given location using a Bayesian probability-derived quality index.
@@
 
-\section{The Dataset}
+\section{Climatological data}
 
@@
 For sunshine duration, each monthly datum corresponds to the daily value averaged over each month.
-All stations reported precipitation values, although only 2002 reported rain days, 445 reported relative humidity, and 336 reported sunshine duration.
-This means that not all stations registered all the meteorological variables relevant to this work.
-Thus, we classify stations according to which variables they measured (Table \ref{tabstype}).\\
+All stations reported precipitation values, although only 2002 reported rain days, 445 reported relative humidity , and 336 reported sunshine duration.
+This means that not all available stations registered all meteorological variables relevant to this work.
+Thus, we classify stations according to which variables were measured (Table \ref{tabstype}).\\
 
@@
 \end{table}
-
-\begin{figure}
-\begin{center}
-\includegraphics[scale=0.5]{ldeh.eps}
-\includegraphics[scale=0.5]{ldel.eps}
-\includegraphics[scale=0.5]{lded.eps}
-\includegraphics[scale=0.5]{ldeb.eps}
-\caption{Lowest BIC Gaussian Mixture Model clustering of reduced dimensionality climate data. Colors indicate the clustering classification.}\label{lde}
-\end{center}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-\includegraphics[scale=0.5]{gmmh0.eps}
-\includegraphics[scale=0.5]{gmmh1.eps}
-\caption{Climate clusters as predicted by the lowest-BIC Gaussian Mixture Model for relative humidity.
-Left column: Average monthly relative humidity and standard deviation (error bars) for each climate cluster.
-Right column: Geographical location and altitude of weather stations in each relative humidity cluster.}\label{clusth}
-\end{center}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-\includegraphics[scale=0.5]{gmml0.eps}
-\includegraphics[scale=0.5]{gmml1.eps}
-\includegraphics[scale=0.5]{gmml2.eps}
-\includegraphics[scale=0.5]{gmml3.eps}
-\includegraphics[scale=0.5]{gmml4.eps}
-\includegraphics[scale=0.5]{gmml5.eps}
-\includegraphics[scale=0.5]{gmml6.eps}
-\includegraphics[scale=0.5]{gmml7.eps}
-\includegraphics[scale=0.5]{gmml8.eps}
-\includegraphics[scale=0.5]{gmml9.eps}
-\includegraphics[scale=0.5]{gmml10.eps}
-\caption{Climate clusters as predicted by the lowest-BIC Gaussian Mixture Model for precipitation.
-Left column: Average monthly precipitation and standard deviation (error bars) for each climate cluster.
-Right column: Geographical location and altitude of weather stations in each precipitation cluster.}\label{clustl}
-\end{center}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-\includegraphics[scale=0.5]{gmmd0.eps}
-\includegraphics[scale=0.5]{gmmd1.eps}
-\includegraphics[scale=0.5]{gmmd2.eps}
-\includegraphics[scale=0.5]{gmmd3.eps}
-\includegraphics[scale=0.5]{gmmd4.eps}
-\includegraphics[scale=0.5]{gmmd5.eps}
-\caption{Climate clusters as predicted by the lowest-BIC Gaussian Mixture Model for rain days.
-Left column: Average monthly rain days and standard deviation (error bars) for each climate cluster.
-Right column: Geographical location and altitude of weather stations in each rain days cluster.}\label{clustl}
-\end{center}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-\includegraphics[scale=0.5]{gmmb0.eps}
-\includegraphics[scale=0.5]{gmmb1.eps}
-\includegraphics[scale=0.5]{gmmb2.eps}
-\includegraphics[scale=0.5]{gmmb3.eps}
-\includegraphics[scale=0.5]{gmmb4.eps}
-\caption{Climate clusters as predicted by the lowest-BIC Gaussian Mixture Model for sunshine duration.
-Left column: Average monthly sunshine duration and standard deviation (error bars) for each climate cluster.
-Right column: Geographical location and altitude of weather stations in each sunshine duration cluster.}\label{clustl}
-\end{center}
-\end{figure}
 

### 2017-04-20 pasp/paperclimate.tex
more plots in paper

@@
 
-Besides astronomical observations, satellite communications technology in the THz frequency range is a rapidly growing field \cite{newera}.
+Besides astronomical observations in the THz frequency range, development of THz satellite communications technology is also a rapidly growing field \cite{newera}.
 Atmospheric models for absorption of THz photons \cite{rosenkranz,lababs} and artifact removal models \cite{removal} have been developed recently, specially considering that the ``classical radio window'' in telecommunications has become saturated, and terabit satellite links could be used in the near future \cite{suen2016}.

### 2017-04-21 pasp/paperclimate.tex
big break

--- 2017-04-20 pasp/paperclimate.tex
@@
 %\vspace{2pc}
-\noindent{\it Keywords}: water vapor, atmospheric effects, site testing, absorption, integrated water vapor, radio astronomy, satellite communication, atmospheric propagation\\
+\noindent{\it Keywords}: Water Vapor, Atmospheric Effects, Site Testing, Radio Astronomy, Satellite Communication, \\
 %
@@
 
-The development of astronomical instrumentation technology in the 0.2-2 THz range has been rapidly growing in recentl years along with THz satellite communications technology \cite{newera} specially considering that the ``classical radio window'' in telecommunications has become saturated.
-Atmospheric models for absorption of THz photons \cite{rosenkranz,lababs} and artifact removal models \cite{removal} have been developed recently.
+The development of astronomical instrumentation technology in the 0.2-2 THz range has been rapidly growing in recent years.
+Also, interest in THz technology, specially applied to satellite telecommunications has also grown due to the saturation of the ``classical radio window'' in telecommunications  \cite{newera}, and atmospheric models for absorption of THz photons \cite{rosenkranz,lababs} and artifact removal models \cite{removal} have been developed recently due to this interest.
 Near-equatorial countries face a challenge in using this frequency band for astronomical observations and telecommunications using terabit satellite links  \cite{suen2016} due to the presence of a tropical belt of dense water vapor which efficiently absorbs THz radiation \cite{tropicalbelt}.
-Previous studies of astronomical site testing in the visible range have yielded that nighttime humidity conditions across the Northern Andes are only suitable for educational observatories \cite{pinzon}.
-However, considering that millimeter/sub-millimeter observations need not be done during the night, unusually dry, clear-daytime-sky, high altitude regions in the Northern Andes could be suitable for a world-class millimeter-wave observatory.
+In the Northern Andes, previous studies of astronomical site testing in the visible range have shown that high nighttime humidity conditions make this region suitable only for educational observatories \cite{pinzon}.
+However, considering that millimeter/sub-millimeter observations need not be done during the night, unusually dry, clear-daytime-sky, high altitude regions in the Northern Andes could be suitable candidates for a world-class millimeter-wave observatory.
 \\
@@
 \begin{center}
-\includegraphics[scale=0.8]{totmap.eps}
-\caption{1$\sigma$ (solid) and 2$\sigma$ (dashed) countours of a gaussian kernel density estimation of the geographical distribution of weather stations (small green points) from the IDEAM 1980-2010 database.
-The estimation was done using the Python \texttt{scipy.stats.gaussian\_kde}  function.}\label{totmap}
+\includegraphics[scale=0.5]{totmap.pdf}
+\includegraphics[scale=0.5]{elev.pdf}
+\caption{Distribution of weather stations in the dataset.
+(a) 1$\sigma$ (solid) and 2$\sigma$ (dashed) countours of a gaussian kernel density estimation of the geographical distribution of weather stations (small green points) from the IDEAM 1980-2010 database.
+The estimation was done using the Python \texttt{scipy.stats.gaussian\_kde}  function.
+(b) Altitude histogram for all stations in the dataset.}\label{totmap}
 \end{center}
@@
 \begin{center}
-\includegraphics[scale=0.8]{pcah.eps}
+\includegraphics[scale=0.8]{pcah.pdf}
 \caption{Relative humidity data projected across 2 principal components. The dimensionality of the data has been reduced from 12 to 2 while covering 95\% of the variance of the data. }\label{pcah}
@@
 In order to achieve this, we compare the Bayesian Information Criterion \cite{bicref} for a grid of Gaussian clusters ($N_\mathrm{G-C}<20$) and covariance restriction methods for each variable (Figure \ref{bic}).
-We selected the models which yielded the lowest BIC value and plotted the climatological clusters in Figures \ref{clusth}-\ref{clustb}.
-To do this we used the Python \texttt{sklearn.mixture.GMM} module \cite{sklearn}.\\
+To do this we used the Python \texttt{sklearn.mixture.GMM} module \cite{sklearn}.
+We selected the models which produced the lowest BIC value, and plotted the GMM-classified, PCA-projected climate data in Figure \ref{lde}.
+\\
+
+From the climatological clusters in Figures \ref{clusth}-\ref{clustb} we selected clusters that indicate a dry, sunny climate.
+Table \ref{tabclu} shows the selected number of clusters and covariance method for each variable along with our preferred clusters .
+From here on, if a station appears in one of our preferred clusters, we will say that it satisfies our criterion for that specific variable.\\
 
@@
 \begin{center}
-\includegraphics[scale=0.5]{bich.eps}
-\includegraphics[scale=0.5]{bicl.eps}\\
-\includegraphics[scale=0.5]{bicd.eps}
-\includegraphics[scale=0.5]{bicb.eps}
+\includegraphics[scale=0.5]{bich.pdf}
+\includegraphics[scale=0.5]{bicl.pdf}\\
+\includegraphics[scale=0.5]{bicd.pdf}
+\includegraphics[scale=0.5]{bicb.pdf}
 \caption{Bayesian Information Criterion (BIC) results for Gaussian Mixture Models using a different number of clusters and using different covariance restriction methods applied on each measured variable dataset.}\label{bic}
@@
 
-Table \ref{tabclu} shows the selected number of clusters and covariance method for each variable along with our preferred clusters that correlate with dry, sunny climate.
-From here on, if a station appears in one of our preferred clusters, we will say that it satisfies our criterion for that specific variable.
+\begin{figure}
+\begin{center}
+\includegraphics[scale=0.5]{ldeh.pdf}
+\includegraphics[scale=0.5]{ldel.pdf}
+\includegraphics[scale=0.5]{lded.pdf}
+\includegraphics[scale=0.5]{ldeb.pdf}
+\caption{Lowest BIC Gaussian Mixture Model clustering of reduced dimensionality climate data. Colors indicate the results of the clustering classification.}\label{lde}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[scale=0.5]{gmmh0.pdf}
+\includegraphics[scale=0.5]{gmmh1.pdf}
+\caption{Climate clusters as predicted by the lowest-BIC Gaussian Mixture Model for relative humidity.
+Left column: Average monthly relative humidity and standard deviation (error bars) for each climate cluster.
+Right column: Geographical location and altitude of weather stations in each relative humidity cluster.}\label{clusth}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[scale=0.5]{gmml0.pdf}
+\includegraphics[scale=0.5]{gmml1.pdf}
+\includegraphics[scale=0.5]{gmml2.pdf}
+\includegraphics[scale=0.5]{gmml3.pdf}
+\includegraphics[scale=0.5]{gmml4.pdf}
+\includegraphics[scale=0.5]{gmml5.pdf}
+\includegraphics[scale=0.5]{gmml6.pdf}
+\includegraphics[scale=0.5]{gmml7.pdf}
+\includegraphics[scale=0.5]{gmml8.pdf}
+\includegraphics[scale=0.5]{gmml9.pdf}
+\includegraphics[scale=0.5]{gmml10.pdf}
+\caption{Climate clusters as predicted by the lowest-BIC Gaussian Mixture Model for precipitation.
+Left column: Average monthly precipitation and standard deviation (error bars) for each climate cluster.
+Right column: Geographical location and altitude of weather stations in each precipitation cluster.}\label{clustl}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[scale=0.5]{gmmd0.pdf}
+\includegraphics[scale=0.5]{gmmd1.pdf}
+\includegraphics[scale=0.5]{gmmd2.pdf}
+\includegraphics[scale=0.5]{gmmd3.pdf}
+\includegraphics[scale=0.5]{gmmd4.pdf}
+\includegraphics[scale=0.5]{gmmd5.pdf}
+\caption{Climate clusters as predicted by the lowest-BIC Gaussian Mixture Model for rain days.
+Left column: Average monthly rain days and standard deviation (error bars) for each climate cluster.
+Right column: Geographical location and altitude of weather stations in each rain days cluster.}\label{clustl}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[scale=0.5]{gmmb0.pdf}
+\includegraphics[scale=0.5]{gmmb1.pdf}
+\includegraphics[scale=0.5]{gmmb2.pdf}
+\includegraphics[scale=0.5]{gmmb3.pdf}
+\includegraphics[scale=0.5]{gmmb4.pdf}
+\caption{Climate clusters as predicted by the lowest-BIC Gaussian Mixture Model for sunshine duration.
+Left column: Average monthly sunshine duration and standard deviation (error bars) for each climate cluster.
+Right column: Geographical location and altitude of weather stations in each sunshine duration cluster.}\label{clustl}
+\end{center}
+\end{figure}
 
 \begin{table}
-\caption{\label{tabclu}Climate clusters according to a lowest BIC-based selection of Gaussian Mixture Models.}
+\caption{\label{tabclu}Climate clusters according to a lowest BIC-based selection of Gaussian Mixture Models. See Figure \ref{bic}.}
 \begin{indented}
@@
 
-\begin{figure}
-\begin{center}
-\includegraphics[scale=0.5]{ldeh.eps}
-\includegraphics[scale=0.5]{ldel.eps}
-\includegraphics[scale=0.5]{lded.eps}
-\includegraphics[scale=0.5]{ldeb.eps}
-\caption{Lowest BIC Gaussian Mixture Model clustering of reduced dimensionality climate data. Colors indicate the clustering classification.}\label{lde}
-\end{center}
-\end{figure}
+\section{Discussion}
 
-\begin{figure}
-\begin{center}
-\includegraphics[scale=0.5]{gmmh0.eps}
-\includegraphics[scale=0.5]{gmmh1.eps}
-\caption{Climate clusters as predicted by the lowest-BIC Gaussian Mixture Model for relative humidity.
-Left column: Average monthly relative humidity and standard deviation (error bars) for each climate cluster.
-Right column: Geographical location and altitude of weather stations in each relative humidity cluster.}\label{clusth}
-\end{center}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-\includegraphics[scale=0.5]{gmml0.eps}
-\includegraphics[scale=0.5]{gmml1.eps}
-\includegraphics[scale=0.5]{gmml2.eps}
-\includegraphics[scale=0.5]{gmml3.eps}
-\includegraphics[scale=0.5]{gmml4.eps}
-\includegraphics[scale=0.5]{gmml5.eps}
-\includegraphics[scale=0.5]{gmml6.eps}
-\includegraphics[scale=0.5]{gmml7.eps}
-\includegraphics[scale=0.5]{gmml8.eps}
-\includegraphics[scale=0.5]{gmml9.eps}
-\includegraphics[scale=0.5]{gmml10.eps}
-\caption{Climate clusters as predicted by the lowest-BIC Gaussian Mixture Model for precipitation.
-Left column: Average monthly precipitation and standard deviation (error bars) for each climate cluster.
-Right column: Geographical location and altitude of weather stations in each precipitation cluster.}\label{clustl}
-\end{center}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-\includegraphics[scale=0.5]{gmmd0.eps}
-\includegraphics[scale=0.5]{gmmd1.eps}
-\includegraphics[scale=0.5]{gmmd2.eps}
-\includegraphics[scale=0.5]{gmmd3.eps}
-\includegraphics[scale=0.5]{gmmd4.eps}
-\includegraphics[scale=0.5]{gmmd5.eps}
-\caption{Climate clusters as predicted by the lowest-BIC Gaussian Mixture Model for rain days.
-Left column: Average monthly rain days and standard deviation (error bars) for each climate cluster.
-Right column: Geographical location and altitude of weather stations in each rain days cluster.}\label{clustl}
-\end{center}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-\includegraphics[scale=0.5]{gmmb0.eps}
-\includegraphics[scale=0.5]{gmmb1.eps}
-\includegraphics[scale=0.5]{gmmb2.eps}
-\includegraphics[scale=0.5]{gmmb3.eps}
-\includegraphics[scale=0.5]{gmmb4.eps}
-\caption{Climate clusters as predicted by the lowest-BIC Gaussian Mixture Model for sunshine duration.
-Left column: Average monthly sunshine duration and standard deviation (error bars) for each climate cluster.
-Right column: Geographical location and altitude of weather stations in each sunshine duration cluster.}\label{clustl}
-\end{center}
-\end{figure}
-
-\section{Quality Index for Regions of Interest}
-
-We wish to identify all stations for whom all four criteria for identifying a region as having a stable sunny, dry climate can be met.
+We wish to identify all stations for whom all four criteria (humidity, precipitation, rain days, sunshine duration) for identifying a region as having a stable sunny, dry climate can be met.
 However, since only a few stations measured all variables, it is not straightforward to accept or reject stations based on their satisfing a criterion, i.e.
@@
 We start by creating a shortlist of stations that meet criteria for (and only for) the variables they measured.
+This means that if a station has precipitation and humidity data, it should belong to a humidity and precipitation cluster listed in Table \ref{tabclu} or else it is rejected.
 665 stations across Colombia showed such behavior.
 However, since we are interested in high-mountain regions, we included a criterion for elevation limiting our list to 119 stations located above 2000 masl.
-This shortlist was further reduced by requiring that for a given station, the total average plus one  standard deviation for precipitation, rain days, and relative humidity correspond to less than 4.3 mm/d, 17 d/mo, and 81\% respectively, and for sunshine duration, to more than 4 h/d.
+This shortlist was further reduced by requiring that the total average plus 1$\sigma$ for precipitation, rain days, and relative humidity is less than 4.3 mm/d, 17 d/mo, and 81\% respectively, and for sunshine duration, more than 4 h/d for any given station.
 This left us with 83 weather stations.
@@
 a station that meets less than all four criteria should still be considered in our analysis, but it cannot be given the same importance as a station that meets all four criteria.
-In order to account for this, and controlling for the fact that stations are not distributed evenly in elevation (Figure \ref{figelev}), we propose a quality measure based on a probabilistic analysis of the data.
+In order to account for this, and controlling for the fact that stations are not distributed evenly in elevation (Figure \ref{totmap}b), we propose a quality measure based on a probabilistic analysis of the data.
 
@@
 
+\begin{figure}
+\begin{center}
+\includegraphics[scale=0.5]{slist.pdf}
+\includegraphics[scale=0.5]{slistz.pdf}
+\caption{Map of stations in our final shortlist.
+The highlighted region on (a) is seen in more detail in (b), which corresponds approximately to the Colombian \emph{altiplano} region.
+Marker size is proportional to the quality index $Q_j$, where the largest points correspond to a $Q_j=10$.}\label{shortlist}
+\end{center}
+\end{figure}
+
 We selected clusters of climatologically similar stations across Colombia using an unsupervised learning low-dimensional embedding algorithm for each measured variable.
-From these clusters, 83 weather stations came up as potential candidates for identifying high-mountain sites (at an elevation greater that 2000 masl) with an unusually dry, sunny weather (Figure \ref{mapsl}).\\
+From these clusters, 83 weather stations came up as potential candidates for identifying high-mountain sites (at an elevation greater that 2000 masl) with an unusually dry, sunny weather.\\
 
@@
 The extrapolated data suggested that 4 of the stations in the shortlist did not satisfy the relative humidity and sunshine duration criteria.
-The remaining 73 stations are plotted in Figure \ref{73sl}.\\
+The remaining 79 stations are mapped in Figure \ref{shortlist}a.\\
 
@@
 
-The final 70 stations located nearby the Cundinamarca-Boyac\'a region of Colombia are shown in more detail in Figure \ref{boycun}.
-This Figure shows two tentative locations in the northern region (Boyac\'a department) appear, where the quality index $Q_j$ (proportional to the point size) indicates that all four criteria are satisfied.
-In order to see if those regions are geographically correlated, we grouped together these stations using a lowest-BIC spherical covariance Gaussian Mixture Model on their coordinates (Figure \ref{biglatlon}).\\
+The final 70 stations located nearby the Cundinamarca-Boyac\'a region of Colombia are shown in more detail in Figure \ref{shortlist}b.
+This figure shows two candidate locations in the northern region (Boyac\'a department), where the quality index $Q_j$ (proportional to the point size) indicates that all four criteria are satisfied.
+In order to see if those regions are geographically correlated, we grouped together these stations in a simplified manner using a lowest-BIC spherical covariance Gaussian Mixture Model on their coordinates \ref{geobic}.\\
 
-The stations are thus classified, and the colored points in Figure \ref{gmmlatlon} show 6 candidate regions.
-However, in order to see if the prediction from the Gaussian Mixture Models is indicative of regions with similar (dry, sunny) weather, we plotted the predicted variance regions for each cluster along with other stations in our sample (rejects) in Figure \ref{latlonreg}.
+ show 6 candidate regions we plotted the predicted variance regions for each cluster along with other stations in our sample (rejects) in Figure \ref{latlonreg}.
 From this figure, we can reject regions where the number of rejected stations within the predicted variance regions is similar to or higher than the number of stations in our shortlist for a specified cluster.
@@
 
-The region located to the west of Bogot\'a ()is a hilly region at an elevation above 2600 masl, with a large variety of microclimates, which explains the significant amount of stations from this region in our shortlist.
+The region located to the west of Bogot\'a is a hilly region at an elevation above 2600 masl, with a large variety of microclimates, which explains the significant amount of stations from this region in our shortlist.
 However, Figure \ref{lasabana} shows that within this predicted variance region there are more stations that do not satisfy our criteria than stations that do.

### 2017-04-22 pasp/paperclimate.tex
geoclustering

@@
 The highlighted region on (a) is seen in more detail in (b), which corresponds approximately to the Colombian \emph{altiplano} region.
-Marker size is proportional to the quality index $Q_j$, where the largest points correspond to a $Q_j=10$.}\label{shortlist}
+Marker size is proportional to the probability $P_j$, where the largest points correspond to a $Q_j=10$.}\label{shortlist}
 \end{center}
@@
 
-All stations outside of the region located within the latitude, longitude range $[(4.3,6.2),(-72.4,-74.5)]$ seem to be outliers, and while they could indicate unusually dry, sunny regions, there are not enough stations nearby to say anything conclusively.
-Furthermore, those stations only measured one or two variables, which leads us to reject them from our shortlist.
+All stations outside of the region located within the latitude, longitude range $[(4.3,6.2),(-72.4,-74.5)]$ seem to be outliers, and while they could indicate unusually dry, sunny regions, there are not enough stations nearby to say anything conclusively as can be seen in Figure \ref{outliers}.
+Furthermore, those stations only measured precipitation ($Q_j=2.2$), which leads us to reject them from our shortlist.
 \\
+%The points of interest outside this region are too sparsely located to suggest a candidate region of interest, \\
 
@@
 In order to see if those regions are geographically correlated, we grouped together these stations in a simplified manner using a lowest-BIC spherical covariance Gaussian Mixture Model on their coordinates \ref{geobic}.\\
+
+\begin{figure}
+\begin{center}
+\includegraphics[scale=0.55]{outliers.pdf}
+\caption{Map of stations in our final shortlist but outside of the Colombian \emph{altiplano} region highlighted in Figure \ref{shortlist}.
+Marker size is proportional to the Probability $P_j$, where the smallest points correspond to a $Q_j=2.2$.}\label{outliers}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[scale=0.55]{geobic.pdf}
+\includegraphics[scale=0.55]{geobic2.pdf}
+\caption{Results  for a lowest-BIC spherical covariance Gaussian Mixture Model geographical clustering on the stations located inside the red circle in Figure \ref{shortlist}b.
+(a) BIC results for different number of clusters.
+(b) Lowest-BIC GMM geographical clustering classification.}\label{geobic}
+\end{center}
+\end{figure}
 

### 2017-04-25 pasp/paperclimate.tex
fire

@@
 
-The meteorological data used in this study were gathered over 30 years (1980-2010) in 2046 weather stations monitored by the Instituto de Hidrolog\'ia, Meteorolog\'ia y Estudios Ambientales (IDEAM) in Colombia (Figure \ref{totmap}).
+The meteorological data used in this study were gathered over 30 years (1981-2010) in 2046 weather stations monitored by the Instituto de Hidrolog\'ia, Meteorolog\'ia y Estudios Ambientales (IDEAM) in Colombia (Figure \ref{totmap}).
 The  variables that we considered relevant to our work were: Precipitation (mm/mo), Rain Days (d/mo),  Relative Humidity (\%), and Sunshine Duration (h/d).
@@
 \caption{Distribution of weather stations in the dataset.
-(a) 1$\sigma$ (solid) and 2$\sigma$ (dashed) countours of a gaussian kernel density estimation of the geographical distribution of weather stations (small green points) from the IDEAM 1980-2010 database.
+(a) 1$\sigma$ (solid) and 2$\sigma$ (dashed) countours of a gaussian kernel density estimation of the geographical distribution of weather stations (small green points) from the IDEAM 1981-2010 database.
 The estimation was done using the Python \texttt{scipy.stats.gaussian\_kde}  function.
@@
 
-\subsection{Probabilistic Analysis}
+\subsection{Bayesian probabilistic analysis}
 
-We base our analysis on the probability for our hypothesis that a station which appears in our shortlist meets all four criteria, given that it is located at a given elevation.
+We quantify the quality of a given site using the probability for the hypothesis that a station which appears in our shortlist meets all four criteria, given that it is located at a given elevation.
 We can write down this probability in terms of the following events:
@@
 
-\section{Results}
+\subsection{Final selection}
 
@@
 The highlighted region on (a) is seen in more detail in (b), which corresponds approximately to the Colombian \emph{altiplano} region.
-Marker size is proportional to the probability $P_j$, where the largest points correspond to a $Q_j=10$.}\label{shortlist}
+Marker size is proportional to the probability $P_j$, where the largest size corresponds to $Q_j=10$.}\label{shortlist}
 \end{center}
@@
 
-We selected clusters of climatologically similar stations across Colombia using an unsupervised learning low-dimensional embedding algorithm for each measured variable.
-From these clusters, 83 weather stations came up as potential candidates for identifying high-mountain sites (at an elevation greater that 2000 masl) with an unusually dry, sunny weather.\\
-
-Given that many of our candidate stations were lacking in relative humidity and sunshine duration data, we extrapolated using data from nearby stations (at less than 4 km away and at an elevation difference of less than 100 m).
-We made sure that in general, for stations less than 5 km away, the difference in relative humidity and sunshine duration data are less than 15\%.
-The extrapolated data suggested that 4 of the stations in the shortlist did not satisfy the relative humidity and sunshine duration criteria.
+Given that many of our 83 candidate stations were lacking in relative humidity and sunshine duration data, we extrapolated using data from nearby stations (at less than 4 km away and at an elevation difference of less than 100 m).
+We made sure that in general, for stations less than 5 km away, the difference in relative humidity and sunshine duration data is less than 15\%.
+The extrapolated data evidenced that 4 of the stations in the shortlist did not satisfy the relative humidity and sunshine duration criteria.
 The remaining 79 stations are mapped in Figure \ref{shortlist}a.\\
 
-All stations outside of the region located within the latitude, longitude range $[(4.3,6.2),(-72.4,-74.5)]$ seem to be outliers, and while they could indicate unusually dry, sunny regions, there are not enough stations nearby to say anything conclusively as can be seen in Figure \ref{outliers}.
-Furthermore, those stations only measured precipitation ($Q_j=2.2$), which leads us to reject them from our shortlist.
+We rejected 9 of the stations in Figure \ref{shortlist}a, i.e.
+those located outside of the region located within the latitude, longitude range $[(4.3,6.2),(-72.4,-74.5)]$.
+Even tough they could be indicative of unusually dry, sunny regions, there are simply not enough stations nearby to say anything conclusively, as evidenced in Figure \ref{outliers}.
+Adding to this, the only variable they measured was precipitation ($Q_j=2.2$), which leads us to reject them from our shortlist.
 \\
-%The points of interest outside this region are too sparsely located to suggest a candidate region of interest, \\
 
-The final 70 stations located nearby the Cundinamarca-Boyac\'a region of Colombia are shown in more detail in Figure \ref{shortlist}b.
+The final 70 stations located in the Cundinamarca-Boyac\'a region of Colombia are shown in more detail in Figure \ref{shortlist}b.
 This figure shows two candidate locations in the northern region (Boyac\'a department), where the quality index $Q_j$ (proportional to the point size) indicates that all four criteria are satisfied.
-In order to see if those regions are geographically correlated, we grouped together these stations in a simplified manner using a lowest-BIC spherical covariance Gaussian Mixture Model on their coordinates \ref{geobic}.\\
+In order to see if those regions are geographically correlated, we grouped together these stations in a simplified manner using a lowest-BIC spherical covariance Gaussian Mixture Model on their coordinates \ref{geobic}.
 
@@
 \caption{Map of stations in our final shortlist but outside of the Colombian \emph{altiplano} region highlighted in Figure \ref{shortlist}.
-Marker size is proportional to the Probability $P_j$, where the smallest points correspond to a $Q_j=2.2$.}\label{outliers}
+Marker size is proportional to the probability $P_j$, where the smallest size corresponds to $Q_j=2.2$.}\label{outliers}
 \end{center}
@@
 \includegraphics[scale=0.55]{geobic2.pdf}
-\caption{Results  for a lowest-BIC spherical covariance Gaussian Mixture Model geographical clustering on the stations located inside the red circle in Figure \ref{shortlist}b.
+\caption{Geographical clustering results  for a lowest-BIC spherical covariance Gaussian Mixture Model of the stations located inside the red circle in Figure \ref{shortlist}b.
 (a) BIC results for different number of clusters.
-(b) Lowest-BIC GMM geographical clustering classification.}\label{geobic}
+(b) Lowest-BIC GMM geographical clustering classification.
+Marker size is proportional to the probability $P_j$, where the largest size corresponds to  $Q_j=10$ and the smallest size corresponds to $Q_j=2.2$.
+}\label{geobic}
 \end{center}
@@
 
- show 6 candidate regions we plotted the predicted variance regions for each cluster along with other stations in our sample (rejects) in Figure \ref{latlonreg}.
-From this figure, we can reject regions where the number of rejected stations within the predicted variance regions is similar to or higher than the number of stations in our shortlist for a specified cluster.
+\section{Results}
+
+We selected clusters of climatologically similar stations across Colombia using an unsupervised learning low-dimensional embedding algorithm for each measured variable.
+From these clusters, 70 weather stations came up as potential candidates for identifying high-mountain regions (at an elevation greater that 2000 masl) with an unusually dry, sunny weather.\\
+
+Figure \ref{geobic}b shows 6 candidate regions of interest, where weather stations indicating dry, sunny climate are regionally correlated.
+In order to visualize the extent of those regions we plotted the predicted variance regions for each cluster along with other (rejected) stations in our sample in Figure \ref{latlonreg}.
+From this figure, we can reject regions where the number of rejected stations within the predicted variance regions is too similar to the number of stations in our shortlist for a specified candidate region in Figure \ref{geobic}b.
+We discuss the suitability of each of these regions below, from South to North.
+
+\begin{figure}
+\begin{center}
+\includegraphics[scale=0.55]{latlonreg.pdf}
+\caption{Geographical clustering results  for the lowest-BIC spherical covariance Gaussian Mixture Model shown in Figure \ref{geobic}b, indicating the predicted variance regions for each cluster (red circles), and plotted along rejected stations in our original dataset (black points).}\label{latlonreg}
+\end{center}
+\end{figure}
 
@@
 
+\section{Conclusions}
+
+%The points of interest outside this region are too sparsely located to suggest a candidate region of interest, \\
+
  \section*{References}

### 2017-04-25 pasp/paperclimate.tex
falta la tabla

@@
 
-\subsection{Bayesian probabilistic analysis}
+\subsection{Bayesian probabilistic analysis}\label{bpa}
 
@@
 This figure shows two candidate locations in the northern region (Boyac\'a department), where the quality index $Q_j$ (proportional to the point size) indicates that all four criteria are satisfied.
-In order to see if those regions are geographically correlated, we grouped together these stations in a simplified manner using a lowest-BIC spherical covariance Gaussian Mixture Model on their coordinates \ref{geobic}.
+In order to see if those regions are geographically correlated, we grouped together these stations in a simplified manner using a lowest-BIC spherical covariance Gaussian Mixture Model on their coordinates \ref{unif}.
 
@@
 
-\begin{figure}
-\begin{center}
-\includegraphics[scale=0.55]{geobic.pdf}
-\includegraphics[scale=0.55]{geobic2.pdf}
-\caption{Geographical clustering results  for a lowest-BIC spherical covariance Gaussian Mixture Model of the stations located inside the red circle in Figure \ref{shortlist}b.
-(a) BIC results for different number of clusters.
-(b) Lowest-BIC GMM geographical clustering classification.
-Marker size is proportional to the probability $P_j$, where the largest size corresponds to  $Q_j=10$ and the smallest size corresponds to $Q_j=2.2$.
-}\label{geobic}
-\end{center}
-\end{figure}
-
 \section{Results}
@@
 
-Figure \ref{geobic}b shows 6 candidate regions of interest, where weather stations indicating dry, sunny climate are regionally correlated.
-In order to visualize the extent of those regions we plotted the predicted variance regions for each cluster along with other (rejected) stations in our sample in Figure \ref{latlonreg}.
-From this figure, we can reject regions where the number of rejected stations within the predicted variance regions is too similar to the number of stations in our shortlist for a specified candidate region in Figure \ref{geobic}b.
-We discuss the suitability of each of these regions below, from South to North.
+Figure \ref{unif} shows 6 candidate regions of interest, where weather stations indicating dry, sunny climate are regionally correlated.
+In order to visualize the extent of those regions we plotted the predicted variance regions for each cluster along with other (rejected) stations in our sample in Figure \ref{unif}.
+It should be noted that the spherical variance GMM geographical clustering overestimates the extent of each region, and tells us nothing of their actual shape.
+We can now reject regions where the number of rejected stations within the predicted variance regions is too similar to the number of stations in our shortlist in Figure \ref{unif}.
+We discuss the suitability of each of these regions below, from South-West to North-East based on our Bayesian probabilistic quality index (Section \ref{bpa}).
 
@@
 \begin{center}
-\includegraphics[scale=0.55]{latlonreg.pdf}
-\caption{Geographical clustering results  for the lowest-BIC spherical covariance Gaussian Mixture Model shown in Figure \ref{geobic}b, indicating the predicted variance regions for each cluster (red circles), and plotted along rejected stations in our original dataset (black points).}\label{latlonreg}
+\includegraphics[scale=0.65]{unif.pdf}
+\caption{Geographical clustering results  for a lowest-BIC spherical covariance Gaussian Mixture Model of the stations located inside the red circle in Figure \ref{shortlist}b, indicating the predicted variance regions for each cluster (red circles), and plotted along rejected stations in our original dataset (black points).
+Marker size is proportional to the probability $P_j$, where the largest size corresponds to  $Q_j=10$ and the smallest size corresponds to $Q_j=2.2$.
+Embedded plot: BIC results for different number of clusters.
+}\label{unif}
 \end{center}
@@
 
-The region located to the west of Bogot\'a is a hilly region at an elevation above 2600 masl, with a large variety of microclimates, which explains the significant amount of stations from this region in our shortlist.
-However, Figure \ref{lasabana} shows that within this predicted variance region there are more stations that do not satisfy our criteria than stations that do.
-Furthermore, stations that meet and do not meet our criteria are homogeneously distributed.
+The first cluster is located at ($4.77^\circ$N, $74.18^\circ$W) and covers an radius of 15.3 km$^2$.
+This is a region located to the west of Bogot\'a at an elevation above 2600 masl, with a large variety of microclimates.
+This explains the significant amount of stations from this region in our shortlist.
+However, Figure \ref{unif} shows that within this predicted variance region there are more stations that do not satisfy our criteria than stations that do, and no single sub-region can be identified.
 This, compounded with the fact that most (70\%) of these stations have a $Q$ value lower than 5.6 due to the lack of humidity/sunshine data, implies that at the moment this is not a strong candidate region of interest.
 %light blue
+
 \subsection{Valle de Ubat\'e}
 
+This cluster is located at ($5.30^\circ$N, $73.80^\circ$W) and covers a radius of 11.4 km$^2$.
 This region, located to the NNE of Bogot\'a s a hilly region along a valley at an elevation above 2600 masl.
-The ratio of number of stations that do not meet our criteria to the stations in this predicted variance region is low (0.3), and almost half of these stations (\%46) have a $Q$ value higher than 6.
+The ratio of number of stations that do not meet our criteria to the stations in this predicted variance region is low (0.3), and almost half of these stations (46\%) have a $Q_j$ value higher than 6.
 This indicates a potential candidate region, although more  sunshine and humidity data are needed.
-In our next paper we will study the amount of atmospheric water vapor near this region.
 %cyan
@@
 
-The region near the town of Villa de Leyva is known for its dry weather, and the stations located within this region seem to support this belief.
+This cluster is located at ($5.61^\circ$N, $73.55^\circ$W) and covers a radius of 7.11 km$^2$.
+This region is located near the town of Villa de Leyva, and it is known for its dry weather.
 The ratio of stations not in our shortlist vs.
@@
 However, Figure \ref{villadeley} seems to indicate that the actual region of interest is narrower than the predicted variance region, as the stations that do not satisfy our criteria are on the NNW and SSW fringes of said region.
-The presence of one very high-$Q$ station is very suggestive of this region being a strong candidate.
+The presence of one very high-$Q_j$ station and the narrow GMM-predicted variance is indicative of this region being a strong candidate.
 However, its relatively low elevation (near 2200 masl) can signify the presence of too much atmospheric water vapor above the surface.
-This will be covered in our next paper.
 %red
 
-\subsection{Tunja Canton}
+\subsection{Cant\'on de Tunja}
 
-The variance predicted region west of Tunja is located in the historic Tunja Canton, and even though the ratio of stations not in our shortlist to stations in this predicted region is low (0.31), there are simply not enough humidity and sunshine data to make any strong conclusion regarding this region.
-However, the high-elevation location of some of these stations is tantalizing, which is why we will keep this as a region of interest for our next paper.
+This cluster is located at ($5.59^\circ$N, $73.24^\circ$W) and covers a radius of 8.29 km$^2$.
+The GMM-predicted variance region west of Tunja is located in the historic Tunja Canton, and even though the ratio of stations not in our shortlist to stations in this predicted region is low (0.31), there are simply not enough humidity and sunshine data to make any strong conclusion about this region.
+However, the high altitude of some of these stations is tantalizing, which is why we will keep this as a region of interest for our next paper.
 
@@
 
+This cluster is located at ($5.72^\circ$N, $72.96^\circ$W) and covers a radius of 8.06 km$^2$.
 This is one of the most promising regions in our sample.
 Located in a wide, sunny valley (hence the name) at an elevation of 2600 masl, it is surrounded by mountains, and rains are not as common as elsewhere in the country.
-This region has the lowest ratio of not in our shortlist stations to stations in this predicted variance region (0.2), $\%40$ of the stations have a $Q$ value higher than 6, and one of the stations has a very high-$Q$ value.
+This region has the lowest ratio of not in our shortlist stations to stations in this predicted variance region (0.2), $\%40$ of the stations have a $Q_j$ value higher than 6, and one of the stations has a very high-$Q_j$ value.
 This will be one of the regions of interest for our next paper, and we think that it warrants radiometer measurements to be carried out.
@@
 
+This cluster is located at ($6.04^\circ$N, $72.71^\circ$W) and covers a radius of 10.52 km$^2$.
 This is a \emph{p\'aramo} region, characterized by high-mountain tundra weather.
@@
 The lack of a high-$Q$ station means that more sunshine-humidity data are needed, but this is possibly not a region of interest.\\
-
-%figures for all these
 
@@
 
+  % vdu In our next paper we will study the amount of atmospheric water vapor near this region.
+
+% vdl  stations located within this region seem to support this belief. This will be covered in our next paper.
+
  \section*{References}

### 2017-04-26 pasp/paperclimate.tex
listo menos el abstract

@@
 \usepackage{epstopdf}
+\usepackage{booktabs}
 %Uncomment next line if AMS fonts required
@@
 
-\title[Site Location Study for a Millimeter Observatory in the Colombian Andes I: In Situ]{Site location study for a high-mountain millimeter observatory in the Colombian Andes I: In situ data}
+%\title[Site Location Study for a Millimeter Observatory in the Colombian Andes I: In Situ]{Site location study for a high-mountain millimeter observatory in the Colombian Andes I: In situ data}
 
+\title[Low Dimensional Embedding of Climate Data for Radio Astronomical Site Testing]{ Low Dimensional Embedding of Climate Data for Radio Astronomical Site Testing in the Colombian Andes}
 % Low Dimensional Embedding of Climate Data for Radio Astronomical Site Testing in the Colombian Andes
@@
 This paper is organized as follows.
+We describe our dataset and the distribution and types of weather stations in Section 2.
+In Section 3 we describe how we applied low-dimensional embedding algorithms to our data.
+This is followed by a discussion in Section 4, where we include an overview of climate patterns in Colombia, and describe our Bayesian probabilistic quality index to assess the suitability of a given weather station.
+We summarize our results in Section 5, where we identify candidate regions of interest for a mm-wave astronomical observatory site.
+Finally, we discuss our main conclusions and future perspectives in Section 6 .
 
@@
 \includegraphics[scale=0.5]{gmml2.pdf}
-\includegraphics[scale=0.5]{gmml3.pdf}
-\includegraphics[scale=0.5]{gmml4.pdf}
-\includegraphics[scale=0.5]{gmml5.pdf}
-\includegraphics[scale=0.5]{gmml6.pdf}
-\includegraphics[scale=0.5]{gmml7.pdf}
-\includegraphics[scale=0.5]{gmml8.pdf}
-\includegraphics[scale=0.5]{gmml9.pdf}
-\includegraphics[scale=0.5]{gmml10.pdf}
 \caption{Climate clusters as predicted by the lowest-BIC Gaussian Mixture Model for precipitation.
@@
 \begin{center}
+\includegraphics[scale=0.5]{gmml3.pdf}
+\includegraphics[scale=0.5]{gmml4.pdf}
+\includegraphics[scale=0.5]{gmml5.pdf}
+\caption{Climate clusters as predicted by the lowest-BIC Gaussian Mixture Model for precipitation.
+Left column: Average monthly precipitation and standard deviation (error bars) for each climate cluster.
+Right column: Geographical location and altitude of weather stations in each precipitation cluster.}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+
+\includegraphics[scale=0.5]{gmml6.pdf}
+\includegraphics[scale=0.5]{gmml7.pdf}
+\includegraphics[scale=0.5]{gmml8.pdf}
+\caption{Climate clusters as predicted by the lowest-BIC Gaussian Mixture Model for precipitation.
+Left column: Average monthly precipitation and standard deviation (error bars) for each climate cluster.
+Right column: Geographical location and altitude of weather stations in each precipitation cluster.}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[scale=0.5]{gmml9.pdf}
+\includegraphics[scale=0.5]{gmml10.pdf}
+\caption{Climate clusters as predicted by the lowest-BIC Gaussian Mixture Model for precipitation.
+Left column: Average monthly precipitation and standard deviation (error bars) for each climate cluster.
+Right column: Geographical location and altitude of weather stations in each precipitation cluster.}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
 \includegraphics[scale=0.5]{gmmd0.pdf}
@@
 \includegraphics[scale=0.5]{gmmd2.pdf}
+\caption{Climate clusters as predicted by the lowest-BIC Gaussian Mixture Model for rain days.
+Left column: Average monthly rain days and standard deviation (error bars) for each climate cluster.
+Right column: Geographical location and altitude of weather stations in each rain days cluster.}\label{clustd}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
 \includegraphics[scale=0.5]{gmmd3.pdf}
@@
 Left column: Average monthly rain days and standard deviation (error bars) for each climate cluster.
-Right column: Geographical location and altitude of weather stations in each rain days cluster.}\label{clustl}
+Right column: Geographical location and altitude of weather stations in each rain days cluster.}
 \end{center}
@@
 \includegraphics[scale=0.5]{gmmb2.pdf}
+\caption{Climate clusters as predicted by the lowest-BIC Gaussian Mixture Model for sunshine duration.
+Left column: Average monthly sunshine duration and standard deviation (error bars) for each climate cluster.
+Right column: Geographical location and altitude of weather stations in each sunshine duration cluster.}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
 \includegraphics[scale=0.5]{gmmb3.pdf}
@@
 Left column: Average monthly sunshine duration and standard deviation (error bars) for each climate cluster.
-Right column: Geographical location and altitude of weather stations in each sunshine duration cluster.}\label{clustl}
+Right column: Geographical location and altitude of weather stations in each sunshine duration cluster.}\label{clustb}
 \end{center}
@@
 \mr
-Precipitation&$P$&11&Full&2,3,7,11\\
+Precipitation &$R$&11&Full&2,3,7,11\\
 Rain Days&$D$&6&Full&1,4,6\\
@@
 \begin{itemize}
-\item $C$ is actually the conjunction of all four of the $P,D,H,S$ events ``The station belongs to a preferred cluster for variable $X$'', i.e. $C=P\cap R\cap H \cap S$.
+\item $C$ is actually the conjunction of all four of the $R,D,H,S$ events ``The station belongs to a preferred cluster for variable $X$'', i.e. $C=R\cap D\cap H \cap S$.
 \end{itemize}
@@
 \begin{itemize}
-\item For example, if $M_X$ is the event ``the station measured variable $X$'', then the event $T_7$ is equivalent to $M_P\cap \neg M_R\cap M_H \cap M_S$
+\item For example, if $M_X$ is the event ``the station measured variable $X$'', then the event $T_7$ is equivalent to $M_R\cap \neg M_D\cap M_H \cap M_S$
 \end{itemize}
@@
 Here $P(h\ |\ C\cap A\cap T_i$ is the distribution of elevations for a station type ($T_i$) on our shortlist ($C\cap A$), $P(A\ |\ T_i)$ is the probability that a station of type $T_i$ is in our shortlist, $P(T_i)$ is the probability that a station type is $T_i$, and $P(C\ |\ A\cap T_i)$ is the probability that a station meets all four criteria ($C$) given that it is in our shortlist ($A$) and is of type $T_i$.
-All of these probabilities except for the last one can be computed directly from the data.
+All of these probabilities except for the last one can be computed directly from the data\footnote{Distributions of elevation-related probabilities were approximated discretely using a bin size estimated from the mean elevation difference between stations when sorted by elevation.
+}.
 The probability $P(C\ |\ A\cap T_i)$ has to be estimated indirectly, as we cannot presume to know under which conditions a station can satisfy all our criteria.
@@
 \begin{eqnarray}
-P(C\ |\ A\cap T_i)&=P(P\cap D\cap H\cap S\ |\ A\cap T_i)\ , \\
+P(C\ |\ A\cap T_i)&=P(R\cap D\cap H\cap S\ |\ A\cap T_i)\ , \\
 \label{pclanti}
-&=\frac{P(P\cap D\cap H\cap S\cap A\cap T_i)}{P(A\cap T_i)}\ . 
+&=\frac{P(R\cap D\cap H\cap S\cap A\cap T_i)}{P(A\cap T_i)}\ . 
 \end{eqnarray}
@@
 \begin{equation}
-T_1=M_P\cap M_D\cap\neg M_H\cap\neg M_S\ .
+T_1=M_R\cap M_D\cap\neg M_H\cap\neg M_S\ .
 \end{equation}
-If this station of type $T_1$ is in our shortlist ($A$), the $P$ and $D$ criteria are met. Thus,
+If this station of type $T_1$ is in our shortlist ($A$), the $R$ and $D$ criteria are met. Thus,
 \begin{equation}
-P(P\cap D\cap H\cap S\cap A\cap T_1)=P( H\cap S\ |\ A\cap T_1)\ .
+P(R\cap D\cap H\cap S\cap A\cap T_1)=P( H\cap S\ |\ A\cap T_1)\ .
 \end{equation}
@@
 
-Since the probabilities for stations in our shortlist vary by orders of magnitude, for clarity's sake we decided to use a logarithmic quality index for a station in our shortlist.
+Since the probabilities for stations in our shortlist vary by orders of magnitude, for clarity's sake we decided to use a probability-based logarithmic quality index for all stations in our shortlist.
 For $j=(0,1,...,N_\mathrm{stations})$,
@@
 \begin{center}
-\includegraphics[scale=0.5]{slist.pdf}
-\includegraphics[scale=0.5]{slistz.pdf}
-\caption{Map of stations in our final shortlist.
+\includegraphics[scale=0.55]{slist.pdf}
+\includegraphics[scale=0.55]{slistz.pdf}
+\caption{Map of stations in our shortlist.
 The highlighted region on (a) is seen in more detail in (b), which corresponds approximately to the Colombian \emph{altiplano} region.
@@
 We rejected 9 of the stations in Figure \ref{shortlist}a, i.e.
-those located outside of the region located within the latitude, longitude range $[(4.3,6.2),(-72.4,-74.5)]$.
+those located outside of the region located within the latitude, longitude range $[(4.3,6.2)^\circ,(-72.4,-74.5)^\circ]$.
 Even tough they could be indicative of unusually dry, sunny regions, there are simply not enough stations nearby to say anything conclusively, as evidenced in Figure \ref{outliers}.
@@
 We selected clusters of climatologically similar stations across Colombia using an unsupervised learning low-dimensional embedding algorithm for each measured variable.
-From these clusters, 70 weather stations came up as potential candidates for identifying high-mountain regions (at an elevation greater that 2000 masl) with an unusually dry, sunny weather.\\
+From these clusters, 70 weather stations came up as potential candidates for identifying high-mountain regions (at an elevation greater that 2000 masl) with an unusually dry, sunny weather.
+Seasonally, the driest months of the year for all the candidate regions of interest are December-February and June-August.
+
+\subsection{Candidate regions of interest}
 
@@
 
-\subsection{La Sabana}
+\subsubsection{La Sabana}
 
@@
 This, compounded with the fact that most (70\%) of these stations have a $Q$ value lower than 5.6 due to the lack of humidity/sunshine data, implies that at the moment this is not a strong candidate region of interest.
-%light blue
 
-\subsection{Valle de Ubat\'e}
+\subsubsection{Valle de Ubat\'e}
 
@@
 This indicates a potential candidate region, although more  sunshine and humidity data are needed.
-%cyan
 
-\subsection{Villa de Leyva}
+\subsubsection{Villa de Leyva}
 
@@
 the stations in this predicted variance region is not very low (0.4).
-However, Figure \ref{villadeley} seems to indicate that the actual region of interest is narrower than the predicted variance region, as the stations that do not satisfy our criteria are on the NNW and SSW fringes of said region.
+However, Figure \ref{unif} seems to indicate that the actual region of interest is narrower than the predicted variance region, as the stations that do not satisfy our criteria are on the NNW and SSW fringes of said region.
 The presence of one very high-$Q_j$ station and the narrow GMM-predicted variance is indicative of this region being a strong candidate.
 However, its relatively low elevation (near 2200 masl) can signify the presence of too much atmospheric water vapor above the surface.
-%red
 
-\subsection{Cant\'on de Tunja}
+\subsubsection{Cant\'on de Tunja}
 
@@
 
-% dark blue
-
-\subsection{Valle del Sol}
+\subsubsection{Valle del Sol}
 
@@
 Located in a wide, sunny valley (hence the name) at an elevation of 2600 masl, it is surrounded by mountains, and rains are not as common as elsewhere in the country.
-This region has the lowest ratio of not in our shortlist stations to stations in this predicted variance region (0.2), $\%40$ of the stations have a $Q_j$ value higher than 6, and one of the stations has a very high-$Q_j$ value.
+This region has the lowest ratio of not in our shortlist stations to stations in this predicted variance region (0.2), $40\%$ of the stations have a $Q_j$ value higher than 6, and one of the stations has a very high-$Q_j$ value.
 This will be one of the regions of interest for our next paper, and we think that it warrants radiometer measurements to be carried out.
 
-\subsection{Pisba National Park}
+\subsubsection{Parque Nacional Natural Pisba}
 
 This cluster is located at ($6.04^\circ$N, $72.71^\circ$W) and covers a radius of 10.52 km$^2$.
-This is a \emph{p\'aramo} region, characterized by high-mountain tundra weather.
-The ratio of stations not in our shortlist to station in this predicted variance region is high (0.5), and it is unlikely that the mist allows for a low-atmospheric water vapor region to be located here, even if rain is sporadic.
-The lack of a high-$Q$ station means that more sunshine-humidity data are needed, but this is possibly not a region of interest.\\
+This is a \emph{p\'aramo} region, characterized by high-mountain tundra weather, with some mountains reaching an elevation of up to 3800 masl.
+The ratio of stations rejected stations to stations in this predicted variance region is high (0.5), and it is unlikely that the mist allows for a low-atmospheric water vapor region to be located here, even if rain is sporadic.
+The lack of a high-$Q$ station means that more sunshine-humidity data are needed, but this is possibly not a region of interest.
 
-% tables for our three favorites, Ubat\'e, Villa, Sol.
+\subsection{Comparison to other works}
 
-One, to the NWW of Bogot\'a, another to the N of Bogot\'a, in the region known as Valle de Ubat\'e, \\
-
-4000 masl might be interesting as well, but there are not enough stations.\\
-
-Seasonality plots\\
+We compared our results with the work of  \citeasnoun{suen2014} and \citeasnoun{suen2016}, where they obtained global precipitable water vapor maps for 2012 using data from the MODIS instrument on board the Aqua and Terra satellites.
+The lowest mean pwv ($\sim15$ mm) regions in the \citeasnoun{suen2014} map for Colombia are near the Ubat\'e, Tunja, Valle del Sol, and Pisba regions.
+The \citeasnoun{suen2016} global median pwv map has an upper bound of 10 mm for stations located in the clusters corresponding to those same regions.
+Unfortunately, the resolution of the maps is not fine enough to pinpoint a site location at a precision higher than $\sim20$ km.
 
@@
 
-%The points of interest outside this region are too sparsely located to suggest a candidate region of interest, \\
+\begin{table}
+\caption{\label{tabfin}Regions of interest with a high-altitude daytime dry, clear-sky climate, organized by geographical cluster number ($N_C$), station location, Bayesian probability quality measure ($Q_j$), distance to cluster center ($r_c$), mean precipitation  ($R$), mean rain days ($D$), mean relative humidity  ($h$), mean sunshine duration  ($S$), and altitude ($h$).
+Detailed table (including all clusters) available at \texttt{https://github.com/saint-germain/ideam} \\}
 
-  % vdu In our next paper we will study the amount of atmospheric water vapor near this region.
+\begin{tabular}{rlrrrrrrr}
+\toprule
+ $N_C$  &                 Municipality  &  $Q_j$ &  $r_c$  &  Mean $R$ &  Mean $D$  &  Mean $H$  &  Mean $S$ &  $h$ \\
+ &                   &    &    (km) &   (mm/mo) &  (d/mo) &   (\%) &   (h/d) &  (masl) \\
+\midrule
+              2 &                      Nemocon &            2.2 &                        28 &                        66 &                   ND &                        ND &                          ND &    2600 \\
+              2 &                         Susa &            4.6 &                        18 &                        83 &                   12 &                        ND &                          ND &    2600 \\
+              2 &                     Cucunuba &            4.7 &                         6 &                        57 &                    9 &                        ND &                          ND &    2620 \\
+              2 &                    Sutatausa &            4.8 &                         6 &                        57 &                   10 &                        ND &                          ND &    2700 \\
+              2 &                     Guacheta &            4.9 &                        16 &                        73 &                   10 &                        ND &                          ND &    2690 \\
+              2 &                  Lenguazaque &            5.2 &                        10 &                        63 &                   12 &                        ND &                          ND &    2650 \\
+              2 &                     Simijaca &            5.2 &                        23 &                        67 &                   11 &                        ND &                          ND &    2572 \\
+              2 &                       Suesca &            5.6 &                        20 &                        59 &                   10 &                        ND &                          ND &    2575 \\
+              2 &             Carmen de Carupa &            6.1 &                        12 &                        54 &                   12 &                        ND &                          ND &    2970 \\
+              2 &                     Cucunuba &            6.1 &                         7 &                        51 &                    5 &                        ND &                          ND &    2562 \\
+              2 &                  Lenguazaque &            6.1 &                         9 &                        57 &                   10 &                        ND &                          ND &    2585 \\
+              2 &                         Susa &            6.1 &                        11 &                        60 &                    8 &                        ND &                          ND &    3130 \\
+              2 &  Ubate &            6.1 &                         4 &                        62 &                   13 &                        ND &                          ND &    2555 \\
+              2 &  Ubate &            6.1 &                         5 &                        62 &                   12 &                        ND &                          ND &    2610 \\
+              2 &                      Nemocon &            7.6 &                        20 &                        52 &                   10 &                        77 &                           4 &    2580 \\
+              \midrule
+              4 &                      Duitama &            4.0 &                        17 &                        69 &                   10 &                        ND &                          ND &    2590 \\
+              4 &                        Paipa &            4.0 &                        12 &                        74 &                    9 &                        ND &                          ND &    2900 \\
+              4 &                        Nobsa &            5.0 &                         6 &                        66 &                   10 &                        ND &                          ND &    2500 \\
+              4 &        Santa Rosa &            5.0 &                        10 &                        70 &                    9 &                        ND &                          ND &    2500 \\
+              4 &                     Sogamoso &            5.0 &                         4 &                        59 &                   10 &                        ND &                          ND &    2500 \\
+              4 &                     Sogamoso &            5.0 &                         7 &                        60 &                   12 &                        ND &                          ND &    2500 \\
+              4 &                     Tibasosa &            5.0 &                         5 &                        62 &                    7 &                        ND &                          ND &    2500 \\
+              4 &                          Iza &            5.6 &                        12 &                        52 &                    9 &                        ND &                          ND &    2470 \\
+              4 &                   Firavitoba &            6.1 &                         6 &                        57 &                    9 &                        ND &                          ND &    2486 \\
+              4 &                       Mongui &            6.1 &                        12 &                        65 &                    9 &                        ND &                          ND &    2970 \\
+              4 &                     Sogamoso &            6.1 &                        11 &                        65 &                   10 &                        ND &                          ND &    3225 \\
+              4 &                     Sogamoso &            6.7 &                         5 &                        61 &                   12 &                        75 &                           5 &    2500 \\
+              4 &                        Nobsa &           10.0 &                        10 &                        68 &                   13 &                        74 &                           4 &    2530 \\
 
-% vdl  stations located within this region seem to support this belief. This will be covered in our next paper.
+\bottomrule
+\end{tabular}
+
+\end{table}
+
+In this paper we seek to find plausible locations for a high-mountain mm-wave astronomical observatory in the northern Andes of Colombia, by analizing 30 years of climate data from 2046 weather stations.
+By low-dimensionally embedding the data, we are able to group together and correlate climate behavior indicative of daytime dry, clear-sky weather conditions at elevations higher than 2000 masl.
+A repository with all our code and the full dataset is available at \texttt{https://github.com/saint-germain/ideam}.\\
+
+From our shortlist of 79 stations that met our criteria, we selected 70 stations located  in the Cundinamarca-Boyaca \emph{altiplano} region of Colombia, which is a high-mountain basin with an average altitude of 2600 masl.
+Weather stations in our shortlist but outside this region are too sparsely located to suggest a candidate region of interest.
+\\
+
+From those 70 stations we obtained 6 geographically correlated candidate regions of interest: La Sabana, Valle de Ubat\'e, Cant\'on de Tunja, Villa de Leyva, Valle del Sol, and Parque Nacional Natural Pisba.
+Seasonally, there appear to be two times of the year (Dec-Feb and Jun-Aug) where the weather conditions indicate the best daytime dry, clear-sky conditions.
+These months could be the best for mm-wave astronomical observations to be carried out.\\
+
+We summarize the information about two regions of interest (Valle de Ubat\'e and Valle del Sol) in Table \ref{tabfin}, selected based on a Bayesian probabilistic quality index.
+Most of the stations in this table are located at elevations above 2600 masl and below 3000 masl.
+The upper bound for precipitable water vapor given by  \citeasnoun{suen2014} and \citeasnoun{suen2016} for these regions is $\sim15$ mm, which means that it is not implausible to find a site with an even lower pwv column in one of the regions of interest identified here.
+For instance, there are some mountains nearby Valle de Ubat\'e and Valle del Sol which have altitudes of up to 3400 masl, which are sites that warrant further measurements using radio sondes and radiometers.
+\\
+
+The Villa de Leyva region, despite having a dry, sunny climate, is located at a comparatively low altitude (2200 masl), which means that the precipitable water vapor column above this region is too high ($>20$ mm according to the \citeasnoun{suen2016} maps).\\
+
+There is an additional issue with the Valle del Sol region.
+The high industrial activity in this region could fill the air with particulate material, which would nullify the suitability of this region.
+Besides obtaining seasonal high-resolution precipitable water vapor maps, in our next paper we will study the amount of particulate material in this region using MODIS (Aqua and Terra) satellite data.
+
+\section*{Acknowledgments}
+
+This work was made possible by the Vicerrector\'ia de Investigaci\'on 01-2015 research grant from Universidad ECCI.
+We would also like to thank the Instituto de Hidrolog\'ia, Meteorolog\'ia y Estudios Ambientales (IDEAM) for providing us with the dataset.
 

### 2017-05-06 pasp/paperclimate.tex
oscar and my corrections

@@
 
-\title[Low Dimensional Embedding of Climate Data for Radio Astronomical Site Testing]{ Low Dimensional Embedding of Climate Data for Radio Astronomical Site Testing in the Colombian Andes}
+\title[Low Dimensional Embedding of Climate Data for Radio Astronomical Site Testing]{Low Dimensional Embedding of Climate Data for Radio Astronomical Site Testing in the Colombian Andes}
 % Low Dimensional Embedding of Climate Data for Radio Astronomical Site Testing in the Colombian Andes
@@
 \ead{oramirezs@ecci.edu.co}
-\author{Oscar Restrepo$^{1}$}
+\author{Oscar Restrepo$^{1,2}$}
 \address{$^1$Grupo de Simulaci\'on, An\'alisis y Modelado, Vicerrector\'ia de Investigaci\'on, Universidad ECCI, Bogot\'a, Colombia}
+\address{$^2$Radio Astronomy Instrumentation Group, Universidad de Chile, Santiago de Chile, Chile}
 \ead{orestrepog@ecci.edu.co}
-\author{Alexander Mart\'inez$^{1,2}$}
+\author{Alexander Mart\'inez$^{1,3}$}
 \address{$^1$Grupo de Simulaci\'on, An\'alisis y Modelado, Vicerrector\'ia de Investigaci\'on, Universidad ECCI, Bogot\'a, Colombia}
-\address{$^2$Instituto de Hidrolog\'ia, Meteorolog\'ia y Estudios Ambientales, Bogot\'a, Colombia}
+\address{$^3$Instituto de Hidrolog\'ia, Meteorolog\'ia y Estudios Ambientales, Bogot\'a, Colombia}
 \vspace{10pt}
@@
 The development of astronomical instrumentation technology in the 0.2-2 THz range has been rapidly growing in recent years.
-Also, interest in THz technology, specially applied to satellite telecommunications has also grown due to the saturation of the ``classical radio window'' in telecommunications  \cite{newera}, and atmospheric models for absorption of THz photons \cite{rosenkranz,lababs} and artifact removal models \cite{removal} have been developed recently due to this interest.
+Another recent surge in interest in telecommunications THz technology has appeared recently due to the saturation of the ``classical radio window'' in telecommunications  \cite{newera}.
+For this reason, there has been development of atmospheric models for absorption of THz photons \cite{rosenkranz,lababs} and artifact removal models \cite{removal}.
 Near-equatorial countries face a challenge in using this frequency band for astronomical observations and telecommunications using terabit satellite links  \cite{suen2016} due to the presence of a tropical belt of dense water vapor which efficiently absorbs THz radiation \cite{tropicalbelt}.
-In the Northern Andes, previous studies of astronomical site testing in the visible range have shown that high nighttime humidity conditions make this region suitable only for educational observatories \cite{pinzon}.
+In the northern Andes, previous studies of astronomical site testing in the visible range have shown that high nighttime humidity conditions make this region suitable only for educational observatories \cite{pinzon}.
 However, considering that millimeter/sub-millimeter observations need not be done during the night, unusually dry, clear-daytime-sky, high altitude regions in the Northern Andes could be suitable candidates for a world-class millimeter-wave observatory.
@@
 This is why medium-to-long term monitoring of the atmosphere above a site is required before building an expensive mm/sub-mm wave radio telescope.
-However, even in best-case scenarios, earth-based telescopes will always be limited by atmospheric absorption of light specially at THz frequencies, where windows of observations are few and in some cases, narrow \cite{arrlim,limits}.
+However, even in best-case scenarios, earth-based telescopes will always be limited by atmospheric absorption of light specially at THz frequencies, where frequency windows of observations are few and in some cases, narrow \cite{arrlim,limits}.
 \\
@@
 For this reason we screened historic climate data for evidence of local long-term clear sky, low humidity conditions as proxies for a locally low precipitable water vapor column.
-By low-dimensionally embedding the data, we are able to reduce the dimensionality of multi-annual monthly data from $N_\mathrm{dim}=12$ to $N_\mathrm{dim}\le3$.
+By low-dimensionally embedding the data, we are able to reduce the dimensionality of multi-annual monthly data from $N_\mathrm{dim}=12$ to $N_\mathrm{dim}\le3$ while preserving $>95$\% of the variance in the data.
 Thus, we can correlate unusually dry regions regardless of their geographical location.
@@
 Finally, we geographically clustered our candidate locations in order to identify regions of interest.
-We compared our regions of interest with regions with unusually low precipitable water vapor in low-resolution water vapor satellite maps \cite{suen2016} and find a possibly significative geographical correlation.
-In an upcoming paper we will analize satellite data in order to retrieve high-resolution seasonal precipitable water vapor maps for our regions of interest.\\
+We compared these regions of interest with regions with unusually low precipitable water vapor in low-resolution ($\sim20$ km) water vapor satellite maps \cite{suen2016} and find a possibly significative geographical correlation.
+In an upcoming paper we will analize satellite data in order to retrieve high-resolution seasonal precipitable water vapor maps for the regions of interest reported here.\\
 
@@
 The meteorological data used in this study were gathered over 30 years (1981-2010) in 2046 weather stations monitored by the Instituto de Hidrolog\'ia, Meteorolog\'ia y Estudios Ambientales (IDEAM) in Colombia (Figure \ref{totmap}).
-The  variables that we considered relevant to our work were: Precipitation (mm/mo), Rain Days (d/mo),  Relative Humidity (\%), and Sunshine Duration (h/d).
-These variables are reported as multi-annual monthly data, i.e.
+The  variables that we considered relevant to our work were: Elevation (in meters above sea level, or masl), Precipitation (mm/mo), Rain Days (d/mo),  Relative Humidity (\%), and Sunshine Duration (h/d).
+The climatological variables are reported as multi-annual monthly data, i.e.
 each variable is reported monthly from January to December averaged over the 30 year range.
@@
 \begin{table}
-\caption{\label{tabstype}Types of stations classified according to its measured variables.}
+\caption{\label{tabstype}Types of stations classified according to its measured variables. Y = measured, N = not measured.}
 \begin{indented}
@@
 We classified stations using a low-dimensional embedding of the data \cite{embed} for each variable.
-This ensures that a classification algorithm can uncover features in the data even even when the data are projected across a number of dimensions that is lower than the native dimensionality of the data (which in this case is monthly, i.e.
+This ensures that a classification algorithm can uncover features in the data even when the data are projected across a number of dimensions that is lower than the native dimensionality of the data (which in this case is monthly, i.e.
 $N_\mathrm{dim}=12$).
 This involves applying a dimensionality reduction algorithm followed by a clustering algorithm.
-Given that we wish to make a grouping of stations showing similar climatological patterns without making too many assumptions on the data, we decided to use unsupervised learning techniques.
+Given that we wish to make a grouping of stations showing similar climatological patterns without making assumptions on the data, we decided to use unsupervised learning techniques.
 In our case, we chose Principal Component Analysis (PCA) covering 95\% of the variance followed by a Gaussian Mixture Model selected by a low/locally minimum Bayesian Information Criterion (BIC).
@@
 To do this we used the Python \texttt{sklearn.decomposition.PCA} module \cite{sklearn}.
-Figure \ref{pcah} shows the relative humidity data projected across 2 principal components explaining 95\% of the variance, thus preserving most of the original structure of the data.
+To illustrate, Figure \ref{pcah} shows the relative humidity data projected across 2 principal components explaining 95\% of the variance, thus preserving most of the original structure of the data.
+Dimensionality-reduced precipitation, rain days, and sunshine duration data are shown later (Figure \ref{lde}).
 
@@
 Depending on the restrictions put on the covariance of each Gaussian distribution across the data space, the number of free parameters can go from $N_\mathrm{G-C}(N_\mathrm{dim}+1)$ (spherical covariance) to  $N_\mathrm{G-C}(3N_\mathrm{dim}+1)$ (full covariance).
-Since the number of components and covariance restriction is given by the user and not \emph{a posteriori} by the algorithm, me need to make sure that the Gaussian components do not over-fit the data, e.g.
+Since the number of components and covariance restriction is given by the user and not \emph{a posteriori} by the algorithm, we need to make sure that the Gaussian components do not over-fit the data, e.g.
 a model which yields one cluster per datum should be disallowed.\\
@@
 \includegraphics[scale=0.5]{bicb.pdf}
-\caption{Bayesian Information Criterion (BIC) results for Gaussian Mixture Models using a different number of clusters and using different covariance restriction methods applied on each measured variable dataset.}\label{bic}
+\caption{Bayesian Information Criterion (BIC) results for Gaussian Mixture Models using a different number of clusters and using different covariance restriction methods \cite{sklearn} applied on each measured variable dataset.}\label{bic}
 \end{center}
@@
 \includegraphics[scale=0.5]{gmmh1.pdf}
-\caption{Climate clusters as predicted by the lowest-BIC Gaussian Mixture Model for relative humidity.
+\caption{Climate clusters (1-2 of 2) as predicted by the lowest-BIC Gaussian Mixture Model for relative humidity.
 Left column: Average monthly relative humidity and standard deviation (error bars) for each climate cluster.
@@
 \includegraphics[scale=0.5]{gmml2.pdf}
-\caption{Climate clusters as predicted by the lowest-BIC Gaussian Mixture Model for precipitation.
+\caption{Climate clusters (1-3 of 11) as predicted by the lowest-BIC Gaussian Mixture Model for precipitation.
 Left column: Average monthly precipitation and standard deviation (error bars) for each climate cluster.
@@
 \includegraphics[scale=0.5]{gmml5.pdf}
-\caption{Climate clusters as predicted by the lowest-BIC Gaussian Mixture Model for precipitation.
+\caption{Climate clusters  (4-6 of 11) as predicted by the lowest-BIC Gaussian Mixture Model for precipitation.
 Left column: Average monthly precipitation and standard deviation (error bars) for each climate cluster.
-Right column: Geographical location and altitude of weather stations in each precipitation cluster.}
+Right column: Geographical location and altitude of weather stations in each precipitation cluster.
+Continuation of Figure \ref{clustl}.}\label{clustl2}
 \end{center}
@@
 \includegraphics[scale=0.5]{gmml8.pdf}
-\caption{Climate clusters as predicted by the lowest-BIC Gaussian Mixture Model for precipitation.
+\caption{Climate clusters (7-9 of 11) as predicted by the lowest-BIC Gaussian Mixture Model for precipitation.
 Left column: Average monthly precipitation and standard deviation (error bars) for each climate cluster.
-Right column: Geographical location and altitude of weather stations in each precipitation cluster.}
+Right column: Geographical location and altitude of weather stations in each precipitation cluster.
+Continuation of Figure \ref{clustl2}.}\label{clustl3}
 \end{center}
@@
 \includegraphics[scale=0.5]{gmml10.pdf}
-\caption{Climate clusters as predicted by the lowest-BIC Gaussian Mixture Model for precipitation.
+\caption{Climate clusters (10-11 of 11) as predicted by the lowest-BIC Gaussian Mixture Model for precipitation.
 Left column: Average monthly precipitation and standard deviation (error bars) for each climate cluster.
-Right column: Geographical location and altitude of weather stations in each precipitation cluster.}
+Right column: Geographical location and altitude of weather stations in each precipitation cluster.
+Continuation of Figure \ref{clustl3}.}\label{clustl4}
 \end{center}
@@
 \includegraphics[scale=0.5]{gmmd2.pdf}
-\caption{Climate clusters as predicted by the lowest-BIC Gaussian Mixture Model for rain days.
+\caption{Climate clusters (1-3 of 6) as predicted by the lowest-BIC Gaussian Mixture Model for rain days.
 Left column: Average monthly rain days and standard deviation (error bars) for each climate cluster.
@@
 \includegraphics[scale=0.5]{gmmd5.pdf}
-\caption{Climate clusters as predicted by the lowest-BIC Gaussian Mixture Model for rain days.
+\caption{Climate clusters (4-6 of 6) as predicted by the lowest-BIC Gaussian Mixture Model for rain days.
 Left column: Average monthly rain days and standard deviation (error bars) for each climate cluster.
-Right column: Geographical location and altitude of weather stations in each rain days cluster.}
+Right column: Geographical location and altitude of weather stations in each rain days cluster.
+Continuation of Figure \ref{clustd}.}
 \end{center}
@@
 \includegraphics[scale=0.5]{gmmb2.pdf}
-\caption{Climate clusters as predicted by the lowest-BIC Gaussian Mixture Model for sunshine duration.
+\caption{Climate clusters (1-3 of 5) as predicted by the lowest-BIC Gaussian Mixture Model for sunshine duration.
 Left column: Average monthly sunshine duration and standard deviation (error bars) for each climate cluster.
-Right column: Geographical location and altitude of weather stations in each sunshine duration cluster.}
+Right column: Geographical location and altitude of weather stations in each sunshine duration cluster.}\label{clustb1}
 \end{center}
@@
 \includegraphics[scale=0.5]{gmmb4.pdf}
-\caption{Climate clusters as predicted by the lowest-BIC Gaussian Mixture Model for sunshine duration.
+\caption{Climate clusters (4-5 of 5) as predicted by the lowest-BIC Gaussian Mixture Model for sunshine duration.
 Left column: Average monthly sunshine duration and standard deviation (error bars) for each climate cluster.
-Right column: Geographical location and altitude of weather stations in each sunshine duration cluster.}\label{clustb}
+Right column: Geographical location and altitude of weather stations in each sunshine duration cluster.
+Continuation of Figure \ref{clustb1}.}\label{clustb}
 \end{center}
@@
 \begin{table}
-\caption{\label{tabclu}Climate clusters according to a lowest BIC-based selection of Gaussian Mixture Models. See Figure \ref{bic}.}
+\caption{\label{tabclu}Climate clusters according to a lowest BIC-based selection of Gaussian Mixture Models following Figure \ref{bic}.
+The selected clusters column refers to the climate clusters shown in Figures \ref{clusth} to \ref{clustd}.
+}
 \begin{indented}
@@
 \br
-Variable&Symbol&Lowest BIC&Lowest BIC&Preferred Clusters\\
+Variable&Symbol&Lowest BIC&Lowest BIC&Selected Clusters\\
 &&No. of Clusters&Cov. Method &\\
 \mr
+Relative Humidity&$H$&2&Full&2\\
 Precipitation &$R$&11&Full&2,3,7,11\\
 Rain Days&$D$&6&Full&1,4,6\\
-Relative Humidity&$H$&2&Full&2\\
 Sunshine Duration&$S$&5&Tied&1,4,5\\
@@
 We wish to identify all stations for whom all four criteria (humidity, precipitation, rain days, sunshine duration) for identifying a region as having a stable sunny, dry climate can be met.
-However, since only a few stations measured all variables, it is not straightforward to accept or reject stations based on their satisfing a criterion, i.e.
+However, since only a few stations measured all variables, it is not straightforward to accept or reject stations based on their satisfying a criterion, i.e.
 belonging to one of our preferred clusters.
@@
 
-We quantify the quality of a given site using the probability for the hypothesis that a station which appears in our shortlist meets all four criteria, given that it is located at a given elevation.
-We can write down this probability in terms of the following events:
+We quantify the quality of a given site using the probability for the hypothesis that a station $j$ which appears in our shortlist meets all four criteria, given that it is located at a given elevation.
+We can write down this probability $P_j$ in terms of the following events:
 
@@
 \begin{itemize}
-\item $C$ is actually the conjunction of all four of the $R,D,H,S$ events ``The station belongs to a preferred cluster for variable $X$'', i.e. $C=R\cap D\cap H \cap S$.
+\item $C$ is actually the conjunction of all four of the $R,D,H,S$ events ``the station belongs to a selected cluster for variable $X$'', i.e. $C=R\cap D\cap H \cap S$ (See Table \ref{tabclu}).
 \end{itemize}
@@
 We cannot  calculate exactly this probability, as it depends on unknown unknowns, i.e.
-whether a station that did not measure a given variable in reality belongs to one of our preferred climatological clusters for that variable.
+whether a station that did not measure a given variable in reality belongs to one of our preferred climatological clusters for that variable (Table \ref{tabclu}).
 Thus we will use Bayes' theorem to help estimate this probability using simpler conditional probabilities.
@@
 \begin{equation}
-P(C\cap A\cap T_i\ |\ h)=\frac{P(C\cap A\cap T_i\cap h)}{P(h)}\ .
+P_j=P(C\cap A\cap T_i\ |\ h)=\frac{P(C\cap A\cap T_i\cap h)}{P(h)}\ .
 \end{equation}
@@
 \begin{equation}
-P(C\cap A\cap T_i\ |\ h)=\frac{P(h\ |\ C\cap A\cap T_i)}{P(h)}P(C\cap A\cap T_i)\ .
+P_j=\frac{P(h\ |\ C\cap A\cap T_i)}{P(h)}P(C\cap A\cap T_i)\ .
 \end{equation}
 If a station satisfies all four criteria, $P(C\cap A\cap T_i)=1$. This probability can be expressed in terms of simpler conditional probabilities,
-\begin{equation}
-P(C\cap A\cap T_i\ |\ h)=\frac{P(h\ |\ C\cap A\cap T_i)}{P(h)}P(C\ |\ A\cap T_i)P(A\ |\ T_i)P(T_i)\ .
+\begin{equation}\label{peejay}
+P_j=\frac{P(h\ |\ C\cap A\cap T_i)}{P(h)}P(C\ |\ A\cap T_i)P(A\ |\ T_i)P(T_i)\ ,
 \end{equation}
-Here $P(h\ |\ C\cap A\cap T_i$ is the distribution of elevations for a station type ($T_i$) on our shortlist ($C\cap A$), $P(A\ |\ T_i)$ is the probability that a station of type $T_i$ is in our shortlist, $P(T_i)$ is the probability that a station type is $T_i$, and $P(C\ |\ A\cap T_i)$ is the probability that a station meets all four criteria ($C$) given that it is in our shortlist ($A$) and is of type $T_i$.
+where 
+\begin{itemize}
+\item $P(h\ |\ C\cap A\cap T_i)$ is the distribution of elevations for a station type ($T_i$) on our shortlist ($C\cap A$), 
+\item $P(A\ |\ T_i)$ is the probability that a station of type $T_i$ is in our shortlist, 
+\item $P(T_i)$ is the probability that a station type is $T_i$, and 
+\item $P(C\ |\ A\cap T_i)$ is the probability that a station meets all four criteria ($C$) given that it is in our shortlist ($A$) and is of type $T_i$. 
+\end{itemize}
 All of these probabilities except for the last one can be computed directly from the data\footnote{Distributions of elevation-related probabilities were approximated discretely using a bin size estimated from the mean elevation difference between stations when sorted by elevation.
-}.
+}.\\
+
  The probability $P(C\ |\ A\cap T_i)$ has to be estimated indirectly, as we cannot presume to know under which conditions a station can satisfy all our criteria.
-We can rewrite this probability as,
+However, we can rewrite this probability as,
 \begin{eqnarray}
@@
 \begin{equation}
-P(C\ |\ A\cap T_1)=\frac{P( H\cap S\ |\ A\cap T_1)}{P(A\cap T_1)}=P(H\cap S\ |\ A\cap T_1)
+P(C\ |\ A\cap T_1)=\frac{P( H\cap S\ |\ A\cap T_1)}{P(A\cap T_1)}=P(H\cap S\ |\ A\cap T_1) \ .
 \end{equation}
@@
 \begin{equation}
-P(C\ |\ A\cap T_i)\simeq P(H\cap S)\ .
+P(C\ |\ A\cap T_)\simeq P(H\cap S)\ .
 \end{equation}
@@
 
-Since the probabilities for stations in our shortlist vary by orders of magnitude, for clarity's sake we decided to use a probability-based logarithmic quality index for all stations in our shortlist.
-For $j=(0,1,...,N_\mathrm{stations})$,
+The probabilities for stations in our shortlist ($P_j$ in Equation \ref{peejay}) can be used to evaluate the quality of a site, but since they vary by orders of magnitude, for the sake of clarity we decided to use instead a probability-derived logarithmic quality index.
+Thus, we define the quality index $Q_j$ for a station $j$ in our shortlist as,
 \begin{equation}
@@
 \caption{Map of stations in our shortlist.
-The highlighted region on (a) is seen in more detail in (b), which corresponds approximately to the Colombian \emph{altiplano} region.
+The highlighted region in (a) is seen in more detail in (b), which corresponds approximately to the Colombian \emph{altiplano} region.
 Marker size is proportional to the probability $P_j$, where the largest size corresponds to $Q_j=10$.}\label{shortlist}
@@
 those located outside of the region located within the latitude, longitude range $[(4.3,6.2)^\circ,(-72.4,-74.5)^\circ]$.
-Even tough they could be indicative of unusually dry, sunny regions, there are simply not enough stations nearby to say anything conclusively, as evidenced in Figure \ref{outliers}.
+Even though they could be indicative of unusually dry, sunny regions, there are simply not enough stations nearby to say anything conclusively, as evidenced in Figure \ref{outliers}.
 Adding to this, the only variable they measured was precipitation ($Q_j=2.2$), which leads us to reject them from our shortlist.
@@
 From these clusters, 70 weather stations came up as potential candidates for identifying high-mountain regions (at an elevation greater that 2000 masl) with an unusually dry, sunny weather.
-Seasonally, the driest months of the year for all the candidate regions of interest are December-February and June-August.
+Seasonally, the driest months of the year for all the candidate regions of interest are December-February and June-August (Figures \ref{clusth}-\ref{clustb}).
 
@@
 
-Figure \ref{unif} shows 6 candidate regions of interest, where weather stations indicating dry, sunny climate are regionally correlated.
+Figure \ref{unif} shows 6 candidate regions of interest, where weather stations indicating dry, sunny climate are regionally correlated within a $10-20$ km radius.
 In order to visualize the extent of those regions we plotted the predicted variance regions for each cluster along with other (rejected) stations in our sample in Figure \ref{unif}.
-It should be noted that the spherical variance GMM geographical clustering overestimates the extent of each region, and tells us nothing of their actual shape.
-We can now reject regions where the number of rejected stations within the predicted variance regions is too similar to the number of stations in our shortlist in Figure \ref{unif}.
+It should be noted that the spherical variance GMM geographical clustering overestimates the extent of each region, and tells us nothing of their actual geographical shape.
+\\
+
+We can now reject regions where the number of rejected stations within the predicted variance regions is similar or greater than the number of stations in our shortlist in Figure \ref{unif}.
 We discuss the suitability of each of these regions below, from South-West to North-East based on our Bayesian probabilistic quality index (Section \ref{bpa}).
@@
 
-The first cluster is located at ($4.77^\circ$N, $74.18^\circ$W) and covers an radius of 15.3 km$^2$.
+The first cluster is located at ($4.77^\circ$N, $74.18^\circ$W) and covers an radius of 15.3 km.
 This is a region located to the west of Bogot\'a at an elevation above 2600 masl, with a large variety of microclimates.
@@
 
-This cluster is located at ($5.30^\circ$N, $73.80^\circ$W) and covers a radius of 11.4 km$^2$.
+This cluster is located at ($5.30^\circ$N, $73.80^\circ$W) and covers a radius of 11.4 km.
 This region, located to the NNE of Bogot\'a s a hilly region along a valley at an elevation above 2600 masl.
@@
 
-This cluster is located at ($5.61^\circ$N, $73.55^\circ$W) and covers a radius of 7.11 km$^2$.
+This cluster is located at ($5.61^\circ$N, $73.55^\circ$W) and covers a radius of 7.11 km.
 This region is located near the town of Villa de Leyva, and it is known for its dry weather.
@@
 
-This cluster is located at ($5.59^\circ$N, $73.24^\circ$W) and covers a radius of 8.29 km$^2$.
+This cluster is located at ($5.59^\circ$N, $73.24^\circ$W) and covers a radius of 8.29 km.
 The GMM-predicted variance region west of Tunja is located in the historic Tunja Canton, and even though the ratio of stations not in our shortlist to stations in this predicted region is low (0.31), there are simply not enough humidity and sunshine data to make any strong conclusion about this region.
@@
 
-This cluster is located at ($5.72^\circ$N, $72.96^\circ$W) and covers a radius of 8.06 km$^2$.
+This cluster is located at ($5.72^\circ$N, $72.96^\circ$W) and covers a radius of 8.06 km.
 This is one of the most promising regions in our sample.
@@
 
-This cluster is located at ($6.04^\circ$N, $72.71^\circ$W) and covers a radius of 10.52 km$^2$.
+This cluster is located at ($6.04^\circ$N, $72.71^\circ$W) and covers a radius of 10.52 km.
 This is a \emph{p\'aramo} region, characterized by high-mountain tundra weather, with some mountains reaching an elevation of up to 3800 masl.
@@
 
-We compared our results with the work of  \citeasnoun{suen2014} and \citeasnoun{suen2016}, where they obtained global precipitable water vapor maps for 2012 using data from the MODIS instrument on board the Aqua and Terra satellites.
+We compared our results to the work of  \citeasnoun{suen2014} and \citeasnoun{suen2016}, where they obtained global precipitable water vapor maps for 2012 using data from the MODIS instrument on board the Aqua and Terra satellites.
 The lowest mean pwv ($\sim15$ mm) regions in the \citeasnoun{suen2014} map for Colombia are near the Ubat\'e, Tunja, Valle del Sol, and Pisba regions.
@@
 Most of the stations in this table are located at elevations above 2600 masl and below 3000 masl.
-The upper bound for precipitable water vapor given by  \citeasnoun{suen2014} and \citeasnoun{suen2016} for these regions is $\sim15$ mm, which means that it is not implausible to find a site with an even lower pwv column in one of the regions of interest identified here.
-For instance, there are some mountains nearby Valle de Ubat\'e and Valle del Sol which have altitudes of up to 3400 masl, which are sites that warrant further measurements using radio sondes and radiometers.
+We validated our results using satellite measurements of the upper bound for precipitable water vapor \cite{suen2014,suen2016} for those two regions, whis is reported to be $\sim15$ mm.
+This means that it is not implausible to find a site with an even lower water vapor column in one of the regions of interest identified here.
+Some mountains nearby Valle de Ubat\'e and Valle del Sol  have altitudes of up to 3400 masl, and therefore warrant further measurements using radio sondes and radiometers.
 \\
@@
 The high industrial activity in this region could fill the air with particulate material, which would nullify the suitability of this region.
-Besides obtaining seasonal high-resolution precipitable water vapor maps, in our next paper we will study the amount of particulate material in this region using MODIS (Aqua and Terra) satellite data.
+Besides obtaining seasonal high-resolution precipitable water vapor maps, in our next paper we will study the amount of particulate material in this region using MODIS (Aqua and Terra) satellite data.\\
+
+The methods used here can be extended and adapted to similar climatological datasets for other countries. This can be a first step in radio astronomy site testing for developing nations.
 

### 2017-05-09 pasp/paperclimate.tex
correcciones oscres

@@
 Near-equatorial countries face a challenge in using this frequency band for astronomical observations and telecommunications using terabit satellite links  \cite{suen2016} due to the presence of a tropical belt of dense water vapor which efficiently absorbs THz radiation \cite{tropicalbelt}.
-In the northern Andes, previous studies of astronomical site testing in the visible range have shown that high nighttime humidity conditions make this region suitable only for educational observatories \cite{pinzon}.
-However, considering that millimeter/sub-millimeter observations need not be done during the night, unusually dry, clear-daytime-sky, high altitude regions in the Northern Andes could be suitable candidates for a world-class millimeter-wave observatory.
+In northern South-America, previous studies of astronomical site testing in the visible range have shown that high nighttime humidity conditions make this region suitable only for educational observatories \cite{pinzon}.
+However, considering that millimeter/sub-millimeter observations need not be done during the night, unusually dry, clear-daytime-sky, high altitude regions in the northern Andes could be suitable candidates for a world-class millimeter-wave observatory.
 \\
@@
 (a) 1$\sigma$ (solid) and 2$\sigma$ (dashed) countours of a gaussian kernel density estimation of the geographical distribution of weather stations (small green points) from the IDEAM 1981-2010 database.
-The estimation was done using the Python \texttt{scipy.stats.gaussian\_kde}  function.
+The estimation was done using the Python \texttt{scipy.stats.gaussian\_kde}  function \cite{scipy}.
 (b) Altitude histogram for all stations in the dataset.}\label{totmap}
@@
 \begin{table}
-\caption{\label{tabstype}Types of stations classified according to its measured variables. Y = measured, N = not measured.}
+\caption{\label{tabstype}Types of stations classified according to its reported variables. Y = measured, N = not measured.}
 \begin{indented}
@@
 
-We classified stations using a low-dimensional embedding of the data \cite{embed} for each variable.
-This ensures that a classification algorithm can uncover features in the data even when the data are projected across a number of dimensions that is lower than the native dimensionality of the data (which in this case is monthly, i.e.
+We classified stations using a low-dimensional embedding \cite{embed} of the available climate data.
+This ensures that a classification algorithm can uncover features in the data even when the data are projected across a number of dimensions that is lower than its native dimensionality (which in this case is monthly, i.e.
 $N_\mathrm{dim}=12$).
-This involves applying a dimensionality reduction algorithm followed by a clustering algorithm.
+This involves performing a dimensionality reduction algorithm followed by a clustering algorithm.
 Given that we wish to make a grouping of stations showing similar climatological patterns without making assumptions on the data, we decided to use unsupervised learning techniques.
@@
 \caption{Map of stations in our shortlist.
-The highlighted region in (a) is seen in more detail in (b), which corresponds approximately to the Colombian \emph{altiplano} region.
+The highlighted region in (a) is seen in more detail in (b), which corresponds approximately to the Colombian \emph{altiplano cundiboyacense} region.
 Marker size is proportional to the probability $P_j$, where the largest size corresponds to $Q_j=10$.}\label{shortlist}
@@
 
-Given that many of our 83 candidate stations were lacking in relative humidity and sunshine duration data, we extrapolated using data from nearby stations (at less than 4 km away and at an elevation difference of less than 100 m).
-We made sure that in general, for stations less than 5 km away, the difference in relative humidity and sunshine duration data is less than 15\%.
+Many of our 83 candidate stations, selected using the method described at the beginning of this section, were lacking in relative humidity and sunshine duration data.
+Given that for stations less than 5 km away the difference in relative humidity and sunshine duration data is less than 15\%, we extrapolated using data from nearby stations (at less than 4 km away and at an elevation difference of less than 100 m).
 The extrapolated data evidenced that 4 of the stations in the shortlist did not satisfy the relative humidity and sunshine duration criteria.
@@
 This figure shows two candidate locations in the northern region (Boyac\'a department), where the quality index $Q_j$ (proportional to the point size) indicates that all four criteria are satisfied.
-In order to see if those regions are geographically correlated, we grouped together these stations in a simplified manner using a lowest-BIC spherical covariance Gaussian Mixture Model on their coordinates \ref{unif}.
+In order to see if those regions are geographically correlated, we grouped together these stations in a simplified manner using a lowest-BIC spherical covariance Gaussian Mixture Model on their coordinates (Figure \ref{unif}).
 
@@
 This cluster is located at ($5.30^\circ$N, $73.80^\circ$W) and covers a radius of 11.4 km.
-This region, located to the NNE of Bogot\'a s a hilly region along a valley at an elevation above 2600 masl.
+This region, located to the NNE of Bogot\'a is a hilly region along a valley at an elevation above 2600 masl.
 The ratio of number of stations that do not meet our criteria to the stations in this predicted variance region is low (0.3), and almost half of these stations (46\%) have a $Q_j$ value higher than 6.
@@
 
-From our shortlist of 79 stations that met our criteria, we selected 70 stations located  in the Cundinamarca-Boyaca \emph{altiplano} region of Colombia, which is a high-mountain basin with an average altitude of 2600 masl.
+From our shortlist of 79 stations that met our criteria, we selected 70 stations located  in the Cundinamarca-Boyac\'a \emph{altiplano} region of Colombia, which is a high-mountain basin with an average altitude of 2600 masl.
 Weather stations in our shortlist but outside this region are too sparsely located to suggest a candidate region of interest.

### 2017-05-09 pasp/paperclimate.tex
penultimos cambios de oleonardo

@@
 Near-equatorial countries face a challenge in using this frequency band for astronomical observations and telecommunications using terabit satellite links  \cite{suen2016} due to the presence of a tropical belt of dense water vapor which efficiently absorbs THz radiation \cite{tropicalbelt}.
-In northern South-America, previous studies of astronomical site testing in the visible range have shown that high nighttime humidity conditions make this region suitable only for educational observatories \cite{pinzon}.
+In northern South America, previous studies of astronomical site testing in the visible range have shown that high nighttime humidity conditions make this region suitable only for educational observatories \cite{pinzon}.
 However, considering that millimeter/sub-millimeter observations need not be done during the night, unusually dry, clear-daytime-sky, high altitude regions in the northern Andes could be suitable candidates for a world-class millimeter-wave observatory.
@@
 
-The first cluster is located at ($4.77^\circ$N, $74.18^\circ$W) and covers an radius of 15.3 km.
+The first cluster is located at ($4.77^\circ$N, $74.18^\circ$W) and covers an radius of 15.3 km (Figure \ref{unif}, purple circle).
 This is a region located to the west of Bogot\'a at an elevation above 2600 masl, with a large variety of microclimates.
@@
 
-This cluster is located at ($5.30^\circ$N, $73.80^\circ$W) and covers a radius of 11.4 km.
+This cluster is located at ($5.30^\circ$N, $73.80^\circ$W) and covers a radius of 11.4 km (Figure \ref{unif}, blue circle).
 This region, located to the NNE of Bogot\'a is a hilly region along a valley at an elevation above 2600 masl.
@@
 
-This cluster is located at ($5.61^\circ$N, $73.55^\circ$W) and covers a radius of 7.11 km.
+This cluster is located at ($5.61^\circ$N, $73.55^\circ$W) and covers a radius of 7.11 km (Figure \ref{unif}, yellow circle).
 This region is located near the town of Villa de Leyva, and it is known for its dry weather.
@@
 
-This cluster is located at ($5.59^\circ$N, $73.24^\circ$W) and covers a radius of 8.29 km.
+This cluster is located at ($5.59^\circ$N, $73.24^\circ$W) and covers a radius of 8.29 km (Figure \ref{unif}, red circle).
 The GMM-predicted variance region west of Tunja is located in the historic Tunja Canton, and even though the ratio of stations not in our shortlist to stations in this predicted region is low (0.31), there are simply not enough humidity and sunshine data to make any strong conclusion about this region.
@@
 
-This cluster is located at ($5.72^\circ$N, $72.96^\circ$W) and covers a radius of 8.06 km.
+This cluster is located at ($5.72^\circ$N, $72.96^\circ$W) and covers a radius of 8.06 km (Figure \ref{unif}, green circle).
 This is one of the most promising regions in our sample.
@@
 
-This cluster is located at ($6.04^\circ$N, $72.71^\circ$W) and covers a radius of 10.52 km.
+This cluster is located at ($6.04^\circ$N, $72.71^\circ$W) and covers a radius of 10.52 km (Figure \ref{unif}, dark green circle).
 This is a \emph{p\'aramo} region, characterized by high-mountain tundra weather, with some mountains reaching an elevation of up to 3800 masl.
@@
 We compared our results to the work of  \citeasnoun{suen2014} and \citeasnoun{suen2016}, where they obtained global precipitable water vapor maps for 2012 using data from the MODIS instrument on board the Aqua and Terra satellites.
-The lowest mean pwv ($\sim15$ mm) regions in the \citeasnoun{suen2014} map for Colombia are near the Ubat\'e, Tunja, Valle del Sol, and Pisba regions.
-The \citeasnoun{suen2016} global median pwv map has an upper bound of 10 mm for stations located in the clusters corresponding to those same regions.
+The lowest mean precipitable water vapor ($\sim15$ mm) regions in the \citeasnoun{suen2014} map for Colombia are near the Ubat\'e, Tunja, Valle del Sol, and Pisba regions.
+The \citeasnoun{suen2016} global median precipitable water vapor map has an upper bound of 10 mm for stations located in the clusters corresponding to those same regions.
 Unfortunately, the resolution of the maps is not fine enough to pinpoint a site location at a precision higher than $\sim20$ km.
@@
 \begin{table}
-\caption{\label{tabfin}Regions of interest with a high-altitude daytime dry, clear-sky climate, organized by geographical cluster number ($N_C$), station location, Bayesian probability quality measure ($Q_j$), distance to cluster center ($r_c$), mean precipitation  ($R$), mean rain days ($D$), mean relative humidity  ($h$), mean sunshine duration  ($S$), and altitude ($h$).
+\caption{\label{tabfin}Regions of interest with a high-altitude daytime dry, clear-sky climate, organized by geographical cluster number ($N_C$), station location, Bayesian probability quality measure ($Q_j$), distance to cluster center ($r_c$), mean precipitation  ($\bar{R}$), mean rain days ($\bar{D}$), mean relative humidity  ($\bar{H}$), mean sunshine duration  ($\bar{S}$), and altitude ($h$).
 Detailed table (including all clusters) available at \texttt{https://github.com/saint-germain/ideam} \\}
+%https://tex.stackexchange.com/questions/98388/how-to-make-table-with-rotated-table-headers-in-latex
 
@@
 \toprule
- $N_C$  &                 Municipality  &  $Q_j$ &  $r_c$  &  Mean $R$ &  Mean $D$  &  Mean $H$  &  Mean $S$ &  $h$ \\
+ $N_C$  &                 Municipality  &  $Q_j$ &  $r_c$  &   $\bar{R}$ &   $\bar{D}$  &   $\bar{H}$  &  $\bar{S}$ &  $h$ \\
  &                   &    &    (km) &   (mm/mo) &  (d/mo) &   (\%) &   (h/d) &  (masl) \\

### 2017-05-09 pasp/paperclimate.tex
penultimos cambios de oleonardo

@@
 \begin{table}
-\caption{\label{tabfin}Regions of interest with a high-altitude daytime dry, clear-sky climate, organized by geographical cluster number ($N_C$), station location, Bayesian probability quality measure ($Q_j$), distance to cluster center ($r_c$), mean precipitation  ($\bar{R}$), mean rain days ($\bar{D}$), mean relative humidity  ($\bar{H}$), mean sunshine duration  ($\bar{S}$), and altitude ($h$).
+\caption{\label{tabfin}Stations in two regions of interest (RI) showing a high-altitude daytime dry, clear-sky climate, organized by location, Bayesian probability quality measure ($Q_j$), distance to cluster center ($r_c$), mean precipitation  ($\bar{R}$), mean rain days ($\bar{D}$), mean relative humidity  ($\bar{H}$), mean sunshine duration  ($\bar{S}$), and altitude ($h$).
 Detailed table (including all clusters) available at \texttt{https://github.com/saint-germain/ideam} \\}
@@
 
-\begin{tabular}{rlrrrrrrr}
+\begin{tabular}{clrrrrrrr}
 \toprule
- $N_C$  &                 Municipality  &  $Q_j$ &  $r_c$  &   $\bar{R}$ &   $\bar{D}$  &   $\bar{H}$  &  $\bar{S}$ &  $h$ \\
+RI&                   Municipality  &  $Q_j$ &  $r_c$  &   $\bar{R}$ &   $\bar{D}$  &   $\bar{H}$  &  $\bar{S}$ &  $h$ \\
  &                   &    &    (km) &   (mm/mo) &  (d/mo) &   (\%) &   (h/d) &  (masl) \\
 \midrule
-              2 &                      Nemocon &            2.2 &                        28 &                        66 &                   ND &                        ND &                          ND &    2600 \\
-              2 &                         Susa &            4.6 &                        18 &                        83 &                   12 &                        ND &                          ND &    2600 \\
-              2 &                     Cucunuba &            4.7 &                         6 &                        57 &                    9 &                        ND &                          ND &    2620 \\
-              2 &                    Sutatausa &            4.8 &                         6 &                        57 &                   10 &                        ND &                          ND &    2700 \\
-              2 &                     Guacheta &            4.9 &                        16 &                        73 &                   10 &                        ND &                          ND &    2690 \\
-              2 &                  Lenguazaque &            5.2 &                        10 &                        63 &                   12 &                        ND &                          ND &    2650 \\
-              2 &                     Simijaca &            5.2 &                        23 &                        67 &                   11 &                        ND &                          ND &    2572 \\
-              2 &                       Suesca &            5.6 &                        20 &                        59 &                   10 &                        ND &                          ND &    2575 \\
-              2 &             Carmen de Carupa &            6.1 &                        12 &                        54 &                   12 &                        ND &                          ND &    2970 \\
-              2 &                     Cucunuba &            6.1 &                         7 &                        51 &                    5 &                        ND &                          ND &    2562 \\
-              2 &                  Lenguazaque &            6.1 &                         9 &                        57 &                   10 &                        ND &                          ND &    2585 \\
-              2 &                         Susa &            6.1 &                        11 &                        60 &                    8 &                        ND &                          ND &    3130 \\
-              2 &  Ubate &            6.1 &                         4 &                        62 &                   13 &                        ND &                          ND &    2555 \\
-              2 &  Ubate &            6.1 &                         5 &                        62 &                   12 &                        ND &                          ND &    2610 \\
-              2 &                      Nemocon &            7.6 &                        20 &                        52 &                   10 &                        77 &                           4 &    2580 \\
+            &                       Nemocon &            2.2 &                        28 &                        66 &                   ND &                        ND &                          ND &    2600 \\
+            &                          Susa &            4.6 &                        18 &                        83 &                   12 &                        ND &                          ND &    2600 \\
+            &                      Cucunuba &            4.7 &                         6 &                        57 &                    9 &                        ND &                          ND &    2620 \\
+            &                     Sutatausa &            4.8 &                         6 &                        57 &                   10 &                        ND &                          ND &    2700 \\
+            &                      Guacheta &            4.9 &                        16 &                        73 &                   10 &                        ND &                          ND &    2690 \\
+            &                   Lenguazaque &            5.2 &                        10 &                        63 &                   12 &                        ND &                          ND &    2650 \\
+            &                      Simijaca &            5.2 &                        23 &                        67 &                   11 &                        ND &                          ND &    2572 \\
+            &                        Suesca &            5.6 &                        20 &                        59 &                   10 &                        ND &                          ND &    2575 \\
+                      \rotatebox{90}{\rlap{Valle de Ubat\'e}}
+            &              Carmen de Carupa &            6.1 &                        12 &                        54 &                   12 &                        ND &                          ND &    2970 \\
+            &                      Cucunuba &            6.1 &                         7 &                        51 &                    5 &                        ND &                          ND &    2562 \\
+            &                   Lenguazaque &            6.1 &                         9 &                        57 &                   10 &                        ND &                          ND &    2585 \\
+            &                          Susa &            6.1 &                        11 &                        60 &                    8 &                        ND &                          ND &    3130 \\
+            &   Ubate &            6.1 &                         4 &                        62 &                   13 &                        ND &                          ND &    2555 \\
+            &   Ubate &            6.1 &                         5 &                        62 &                   12 &                        ND &                          ND &    2610 \\
+            &                       Nemocon &            7.6 &                        20 &                        52 &                   10 &                        77 &                           4 &    2580 \\
                 \midrule
-              4 &                      Duitama &            4.0 &                        17 &                        69 &                   10 &                        ND &                          ND &    2590 \\
-              4 &                        Paipa &            4.0 &                        12 &                        74 &                    9 &                        ND &                          ND &    2900 \\
-              4 &                        Nobsa &            5.0 &                         6 &                        66 &                   10 &                        ND &                          ND &    2500 \\
-              4 &        Santa Rosa &            5.0 &                        10 &                        70 &                    9 &                        ND &                          ND &    2500 \\
-              4 &                     Sogamoso &            5.0 &                         4 &                        59 &                   10 &                        ND &                          ND &    2500 \\
-              4 &                     Sogamoso &            5.0 &                         7 &                        60 &                   12 &                        ND &                          ND &    2500 \\
-              4 &                     Tibasosa &            5.0 &                         5 &                        62 &                    7 &                        ND &                          ND &    2500 \\
-              4 &                          Iza &            5.6 &                        12 &                        52 &                    9 &                        ND &                          ND &    2470 \\
-              4 &                   Firavitoba &            6.1 &                         6 &                        57 &                    9 &                        ND &                          ND &    2486 \\
-              4 &                       Mongui &            6.1 &                        12 &                        65 &                    9 &                        ND &                          ND &    2970 \\
-              4 &                     Sogamoso &            6.1 &                        11 &                        65 &                   10 &                        ND &                          ND &    3225 \\
-              4 &                     Sogamoso &            6.7 &                         5 &                        61 &                   12 &                        75 &                           5 &    2500 \\
-              4 &                        Nobsa &           10.0 &                        10 &                        68 &                   13 &                        74 &                           4 &    2530 \\
+            &                       Duitama &            4.0 &                        17 &                        69 &                   10 &                        ND &                          ND &    2590 \\
+            &                         Paipa &            4.0 &                        12 &                        74 &                    9 &                        ND &                          ND &    2900 \\
+            &                         Nobsa &            5.0 &                         6 &                        66 &                   10 &                        ND &                          ND &    2500 \\
+            &         Santa Rosa &            5.0 &                        10 &                        70 &                    9 &                        ND &                          ND &    2500 \\
+            &                      Sogamoso &            5.0 &                         4 &                        59 &                   10 &                        ND &                          ND &    2500 \\
+            &                      Sogamoso &            5.0 &                         7 &                        60 &                   12 &                        ND &                          ND &    2500 \\
+            &                      Tibasosa &            5.0 &                         5 &                        62 &                    7 &                        ND &                          ND &    2500 \\
+                                              \rotatebox{90}{\rlap{~Valle del Sol}}
+            &                           Iza &            5.6 &                        12 &                        52 &                    9 &                        ND &                          ND &    2470 \\
+            &                    Firavitoba &            6.1 &                         6 &                        57 &                    9 &                        ND &                          ND &    2486 \\
+            &                        Mongui &            6.1 &                        12 &                        65 &                    9 &                        ND &                          ND &    2970 \\
+            &                      Sogamoso &            6.1 &                        11 &                        65 &                   10 &                        ND &                          ND &    3225 \\
+            &                      Sogamoso &            6.7 &                         5 &                        61 &                   12 &                        75 &                           5 &    2500 \\
+            &                         Nobsa &           10.0 &                        10 &                        68 &                   13 &                        74 &                           4 &    2530 \\
 

### 2017-05-09 pasp/paperclimate.tex
penultimos cambios de oleonardo

@@
 There is an additional issue with the Valle del Sol region.
-The high industrial activity in this region could fill the air with particulate material, which would nullify the suitability of this region.
-Besides obtaining seasonal high-resolution precipitable water vapor maps, in our next paper we will study the amount of particulate material in this region using MODIS (Aqua and Terra) satellite data.\\
+The high industrial activity in this region fills the air with particulate material \cite{sogphd,sogamoso}, which would nullify the suitability of this region.
+Besides obtaining seasonal high-resolution precipitable water vapor maps, in our next paper we will study the amount of particulate material in the reported regions of interest using MODIS (Aqua and Terra) satellite data.\\
 
-The methods used here can be extended and adapted to similar climatological datasets for other countries. This can be a first step in radio astronomy site testing for developing nations.
+The methods used here can be extended and adapted to similar climatological datasets in other regions. This can be a preliminary step in radio astronomy site testing in developing countries.
 

### 2017-05-11 mnras/mnras_template.tex
mnras

--- 2017-05-09 pasp/paperclimate.tex
@@
-\documentclass[12pt]{iopart}
-\newcommand{\gguide}{{\it Preparing graphics for IOP Publishing journals}}
-\usepackage{harvard} 
+% mnras_template.tex
+%
+% LaTeX template for creating an MNRAS paper
+%
+% v3.0 released 14 May 2015
+% (version numbers match those of mnras.cls)
+%
+% Copyright (C) Royal Astronomical Society 2015
+% Authors:
+% Keith T. Smith (Royal Astronomical Society)
+
+% Change log
+%
+% v3.0 May 2015
+%    Renamed to match the new package name
+%    Version number matches mnras.cls
+%    A few minor tweaks to wording
+% v1.0 September 2013
+%    Beta testing only - never publicly released
+%    First version: a simple (ish) template for creating an MNRAS paper
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Basic setup. Most papers should leave these options alone.
+\documentclass[a4paper,fleqn,usenatbib]{mnras}
+
+% MNRAS is set in Times font. If you don't have this installed (most LaTeX
+% installations will be fine) or prefer the old Computer Modern fonts, comment
+% out the following line
+%\usepackage{newtxtext,newtxmath}
+% Depending on your LaTeX fonts installation, you might get better results with one of these:
+%\usepackage{mathptmx}
+%\usepackage{txfonts}
+
+% Use vector fonts, so it zooms properly in on-screen viewing software
+% Don't change these lines unless you know what you are doing
+\usepackage[T1]{fontenc}
+\usepackage{ae,aecompl}
+
+%%%%% AUTHORS - PLACE YOUR OWN PACKAGES HERE %%%%%
+
+% Only include extra packages if you really need them. Common packages are:
+\usepackage{graphicx}	% Including figure files
+\usepackage{amsmath}	% Advanced maths commands
+\usepackage{amssymb}	% Extra maths symbols
 \usepackage{nicefrac} 
-\usepackage{graphicx}
 \usepackage{epstopdf}
 \usepackage{booktabs}
-%Uncomment next line if AMS fonts required
-%\usepackage{iopams}  
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%%%% AUTHORS - PLACE YOUR OWN COMMANDS HERE %%%%%
+
+% Please keep new commands to a minimum, and use \newcommand not \def to avoid
+% overwriting existing commands. Example:
+%\newcommand{\pcm}{\,cm$^{-2}$}	% per cm-squared
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%%%%% TITLE PAGE %%%%%%%%%%%%%%%%%%%
+
+% Title of the paper, and the short title which is used in the headers.
+% Keep the title short and informative.
+\title[Low Dimensional Embedding of Climate Data for Radio Astronomical Site Testing]{Low Dimensional Embedding of Climate Data for Radio Astronomical Site Testing in the Colombian Andes}
+
+% The list of authors, and the short list which is used in the headers.
+% If you need two or more lines of authors, add an extra line using \newauthor
+\author[G. Chaparro-Molano et al.]{Germ\'an Chaparro-Molano$^{1}$\thanks{E-mail: gchaparrom@ecci.edu.co}
+Oscar Leonardo Ram\'irez-Su\'arez$^{1}$
+Oscar Restrepo$^{1,2}$\newauthor
+and Alexander Mart\'inez$^{1,3}$\\
+% List of institutions
+$^{1}$Grupo de Simulaci\'on, An\'alisis y Modelado, Vicerrector\'ia de Investigaci\'on, Universidad ECCI, Bogot\'a, Colombia\\
+$^{2}$Radio Astronomy Instrumentation Group, Universidad de Chile, Santiago de Chile, Chile\\
+$^{3}$Instituto de Hidrolog\'ia, Meteorolog\'ia y Estudios Ambientales, Bogot\'a, Colombia
+}
+
+% These dates will be filled out by the publisher
+\date{Accepted XXX. Received YYY; in original form ZZZ}
+
+% Enter the current year, for the copyright statements etc.
+\pubyear{2017}
+
+% Don't change these lines
 \begin{document}
+\label{firstpage}
+\pagerange{\pageref{firstpage}--\pageref{lastpage}}
+\maketitle
 
-%\title[Site Location Study for a Millimeter Observatory in the Colombian Andes I: In Situ]{Site location study for a high-mountain millimeter observatory in the Colombian Andes I: In situ data}
-
-\title[Low Dimensional Embedding of Climate Data for Radio Astronomical Site Testing]{Low Dimensional Embedding of Climate Data for Radio Astronomical Site Testing in the Colombian Andes}
-% Low Dimensional Embedding of Climate Data for Radio Astronomical Site Testing in the Colombian Andes
-
-\author{Germ¬\'an Chaparro Molano$^{1}$}
-\address{$^1$Grupo de Simulaci\'on, An\'alisis y Modelado, Vicerrector\'ia de Investigaci\'on, Universidad ECCI, Bogot\'a, Colombia}
-\ead{gchaparrom@ecci.edu.co}
-\author{Oscar Leonardo Ram\'irez Su\'arez$^{1}$}
-\address{$^1$Grupo de Simulaci\'on, An\'alisis y Modelado, Vicerrector\'ia de Investigaci\'on, Universidad ECCI, Bogot\'a, Colombia}
-\ead{oramirezs@ecci.edu.co}
-\author{Oscar Restrepo$^{1,2}$}
-\address{$^1$Grupo de Simulaci\'on, An\'alisis y Modelado, Vicerrector\'ia de Investigaci\'on, Universidad ECCI, Bogot\'a, Colombia}
-\address{$^2$Radio Astronomy Instrumentation Group, Universidad de Chile, Santiago de Chile, Chile}
-\ead{orestrepog@ecci.edu.co}
-\author{Alexander Mart\'inez$^{1,3}$}
-\address{$^1$Grupo de Simulaci\'on, An\'alisis y Modelado, Vicerrector\'ia de Investigaci\'on, Universidad ECCI, Bogot\'a, Colombia}
-\address{$^3$Instituto de Hidrolog\'ia, Meteorolog\'ia y Estudios Ambientales, Bogot\'a, Colombia}
-\vspace{10pt}
-\begin{indented}
-\item[]December 2016
-\end{indented}
-
+% Abstract of the paper
 \begin{abstract}
-This document describes the  preparation of an article using \LaTeXe\ and 
-\verb"iopart.cls" (the IOP Publishing \LaTeXe\ preprint class file).
-This class file is designed to help 
-authors produce preprints in a form suitable for submission to any of the
-journals listed in table~\ref{jlab1} on the next page.  You are not obliged to use this class file---we accept
-submissions using all common \LaTeX\ class and style files.  The \verb"iopart.cls"
-class file is supplied merely as a convenience for those authors who find it useful.
-This document gives both general advice that applies whatever class file you use, and specific advice
-that applies if you choose to use \verb"iopart.cls".
-
-We also accept submissions in Word format.  Instructions for Word submissions are available via the `Author Guidelines' link at \verb"http://authors.iop.org".
-
-If you have any queries about this document or any aspect of preparing your article for submission please contact us at the e-mail address given above.
+This is a simple template for authors to write new MNRAS papers.
+The abstract should briefly describe the aims, methods, and main results of the paper.
+It should be a single paragraph not more than 250 words (200 words for Letters).
+No references should appear in the abstract.
 \end{abstract}
 
-% Uncomment for PACS numbers
-%\pacs{00.00, 20.00, 42.10}
-%
-% Uncomment for keywords
-%\vspace{2pc}
-\noindent{\it Keywords}: Water Vapor, Atmospheric Effects, Site Testing, Radio Astronomy, Satellite Communication, \\
-%
-% Uncomment for Submitted to journal title message
-\submitto{PASP}
-%
-% Uncomment if a separate title page is required
-%\maketitle
-% 
-% For two-column output uncomment the next line and choose [10pt] rather than [12pt] in the \documentclass declaration
-%\ioptwocol
-%
+% Select between one and six entries from the list of approved keywords.
+% Don't make up new ones.
+\begin{keywords}
+keyword1 -- keyword2 -- keyword3
+\end{keywords}
 
-% kw:
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%%% BODY OF PAPER %%%%%%%%%%%%%%%%%%
 
@@
 The development of astronomical instrumentation technology in the 0.2-2 THz range has been rapidly growing in recent years.
-Another recent surge in interest in telecommunications THz technology has appeared recently due to the saturation of the ``classical radio window'' in telecommunications  \cite{newera}.
-For this reason, there has been development of atmospheric models for absorption of THz photons \cite{rosenkranz,lababs} and artifact removal models \cite{removal}.
-Near-equatorial countries face a challenge in using this frequency band for astronomical observations and telecommunications using terabit satellite links  \cite{suen2016} due to the presence of a tropical belt of dense water vapor which efficiently absorbs THz radiation \cite{tropicalbelt}.
-In northern South America, previous studies of astronomical site testing in the visible range have shown that high nighttime humidity conditions make this region suitable only for educational observatories \cite{pinzon}.
+Another recent surge in interest in telecommunications THz technology has appeared recently due to the saturation of the ``classical radio window'' in telecommunications  \citep{newera}.
+For this reason, there has been development of atmospheric models for absorption of THz photons \citep{rosenkranz,lababs} and artifact removal models \citep{removal}.
+Near-equatorial countries face a challenge in using this frequency band for astronomical observations and telecommunications using terabit satellite links  \citep{suen2016} due to the presence of a tropical belt of dense water vapor which efficiently absorbs THz radiation \citep{tropicalbelt}.
+In northern South America, previous studies of astronomical site testing in the visible range have shown that high nighttime humidity conditions make this region suitable only for educational observatories \citep{pinzon}.
 However, considering that millimeter/sub-millimeter observations need not be done during the night, unusually dry, clear-daytime-sky, high altitude regions in the northern Andes could be suitable candidates for a world-class millimeter-wave observatory.
@@
 
-Astronomical observations in the millimeter and sub-millimeter wavelength range require that atmospheric effects affecting absorption at these wavelengths are kept to a minimum \cite{southpole,radford2016,chajpwv}.
-The main factor contributing to the atmospheric opacity is water vapor, which very efficiently absorbs light in the THz range \cite{pardocerni,pardocerni2,cont350} due to a continuum absorption spectrum formed by collisionally broadened absorption lines of water vapor in this frequency range  \cite{linecont,submmlines,turner2009}.
-In order to characterize a site according to its atmospheric transparency to THz radiation, it is necessary to retrieve the precipitable water vapor profile either via remote sensing techniques such as microwave radiometry \cite{radiometer,paine2000fourier,southpole2}, satellite measurements \cite{aqua,spacemols,suen2014,spaceradio}, radiosonde humidity measurements \cite{radiosonde,radiosonde2} or indirectly via models of in situ climatological measurements \cite{climatology} or GPS-delay studies \cite{gpsorig,gpsmap}.
+Astronomical observations in the millimeter and sub-millimeter wavelength range require that atmospheric effects affecting absorption at these wavelengths are kept to a minimum \citep{southpole,radford2016,chajpwv}.
+The main factor contributing to the atmospheric opacity is water vapor, which very efficiently absorbs light in the THz range \citep{pardocerni,pardocerni2,cont350} due to a continuum absorption spectrum formed by collisionally broadened absorption lines of water vapor in this frequency range  \citep{linecont,submmlines,turner2009}.
+In order to characterize a site according to its atmospheric transparency to THz radiation, it is necessary to retrieve the precipitable water vapor profile either via remote sensing techniques such as microwave radiometry \citep{radiometer,paine2000fourier,southpole2}, satellite measurements \citep{aqua,spacemols,suen2014,spaceradio}, radiosonde humidity measurements \citep{radiosonde,radiosonde2} or indirectly via models of in situ climatological measurements \citep{climatology} or GPS-delay studies \citep{gpsorig,gpsmap}.
 \\
@@
 Most of the precipitable water vapor that can potentially affect the path of a THz photon hoping to get through the atmosphere actually exists in the troposphere.
-The lower the elevation of a potential site, the longer the path becomes for that photon, increasing its chances of being absorbed by a water vapor molecule \cite{liebe1989mpm,linecont,lababs}.
-For this reason, the quality of a site is determined by the atmospheric water vapor column above the potential site \cite{mkradio,southpole3,he2012,atacama}.
+The lower the elevation of a potential site, the longer the path becomes for that photon, increasing its chances of being absorbed by a water vapor molecule \citep{liebe1989mpm,linecont,lababs}.
+For this reason, the quality of a site is determined by the atmospheric water vapor column above the potential site \citep{mkradio,southpole3,he2012,atacama}.
 This water vapor column is usually expressed as the amount of precipitable water vapor, given in mm.\\
 
-The water vapor column above a potential site can change rapidly (during the course of the day) or it can change seasonally \cite{arm2013,caumont2016}.
+The water vapor column above a potential site can change rapidly (during the course of the day) or it can change seasonally \citep{arm2013,caumont2016}.
 This is why medium-to-long term monitoring of the atmosphere above a site is required before building an expensive mm/sub-mm wave radio telescope.
-However, even in best-case scenarios, earth-based telescopes will always be limited by atmospheric absorption of light specially at THz frequencies, where frequency windows of observations are few and in some cases, narrow \cite{arrlim,limits}.
+However, even in best-case scenarios, earth-based telescopes will always be limited by atmospheric absorption of light specially at THz frequencies, where frequency windows of observations are few and in some cases, narrow \citep{arrlim,limits}.
 \\
 
-Although radiometer measurements are desirable for directly measuring the atmospheric water vapor column, their development and/or deployment can be complex and expensive \cite{radiopro,receiver}.
+Although radiometer measurements are desirable for directly measuring the atmospheric water vapor column, their development and/or deployment can be complex and expensive \citep{radiopro,receiver}.
 For this reason we screened historic climate data for evidence of local long-term clear sky, low humidity conditions as proxies for a locally low precipitable water vapor column.
@@
 Finally, we geographically clustered our candidate locations in order to identify regions of interest.
-We compared these regions of interest with regions with unusually low precipitable water vapor in low-resolution ($\sim20$ km) water vapor satellite maps \cite{suen2016} and find a possibly significative geographical correlation.
+We compared these regions of interest with regions with unusually low precipitable water vapor in low-resolution ($\sim20$ km) water vapor satellite maps \citep{suen2016} and find a possibly significative geographical correlation.
 In an upcoming paper we will analize satellite data in order to retrieve high-resolution seasonal precipitable water vapor maps for the regions of interest reported here.\\
@@
 
-\begin{figure}
+\begin{figure*}
 \begin{center}
@@
 (a) 1$\sigma$ (solid) and 2$\sigma$ (dashed) countours of a gaussian kernel density estimation of the geographical distribution of weather stations (small green points) from the IDEAM 1981-2010 database.
-The estimation was done using the Python \texttt{scipy.stats.gaussian\_kde}  function \cite{scipy}.
+The estimation was done using the Python \texttt{scipy.stats.gaussian\_kde}  function \citep{scipy}.
 (b) Altitude histogram for all stations in the dataset.}\label{totmap}
 \end{center}
-\end{figure}
+\end{figure*}
 
@@
 \caption{\label{tabstype}Types of stations classified according to its reported variables. Y = measured, N = not measured.}
-\begin{indented}
-\item[]\begin{tabular}{@{}cccccc}
-\br
+\begin{tabular}{@{}cccccc}
+\midrule
 Type&Precip.&Rain &Relative&Sunshine&No. of Stations\\
 &&Days&Humidity&Duration&\\
-\mr
+\midrule
 $T_1$&Y&N&N&N&29\\
@@
 $T_8$&Y&Y&Y&Y&314\\
-\br
+\bottomrule
 \end{tabular}
-\end{indented}
 \end{table}
@@
 
-We classified stations using a low-dimensional embedding \cite{embed} of the available climate data.
+We classified stations using a low-dimensional embedding \citep{embed} of the available climate data.
 This ensures that a classification algorithm can uncover features in the data even when the data are projected across a number of dimensions that is lower than its native dimensionality (which in this case is monthly, i.e.
@@
 For precipitation, rain days and sunshine duration, $N_\mathrm{dim}^\mathrm{PCA}(95\%)=3$ and for relative humidity $N_\mathrm{dim}^\mathrm{PCA}(95\%)=2$.
-To do this we used the Python \texttt{sklearn.decomposition.PCA} module \cite{sklearn}.
+To do this we used the Python \texttt{sklearn.decomposition.PCA} module \citep{sklearn}.
 To illustrate, Figure \ref{pcah} shows the relative humidity data projected across 2 principal components explaining 95\% of the variance, thus preserving most of the original structure of the data.
@@
 \begin{center}
-\includegraphics[scale=0.8]{pcah.pdf}
+\includegraphics[width=\columnwidth]{pcah.pdf}
 \caption{Relative humidity data projected across 2 principal components. The dimensionality of the data has been reduced from 12 to 2 while covering 95\% of the variance of the data. }\label{pcah}
@@
 
-After projecting the data along the components found by the PCA algorithm, we clustered the dimensionally-reduced data using a Gaussian Mixture Model, which uses an Expectation-Maximization algorithm to find a Maximum Likelihood model composed of a number $N_\mathrm{G-C}$ of Gaussian distributions, each of which represents a cluster in the data \cite{gmm}.
+After projecting the data along the components found by the PCA algorithm, we clustered the dimensionally-reduced data using a Gaussian Mixture Model, which uses an Expectation-Maximization algorithm to find a Maximum Likelihood model composed of a number $N_\mathrm{G-C}$ of Gaussian distributions, each of which represents a cluster in the data \citep{gmm}.
 Depending on the restrictions put on the covariance of each Gaussian distribution across the data space, the number of free parameters can go from $N_\mathrm{G-C}(N_\mathrm{dim}+1)$ (spherical covariance) to  $N_\mathrm{G-C}(3N_\mathrm{dim}+1)$ (full covariance).
@@
 
-In order to achieve this, we compare the Bayesian Information Criterion \cite{bicref} for a grid of Gaussian clusters ($N_\mathrm{G-C}<20$) and covariance restriction methods for each variable (Figure \ref{bic}).
-To do this we used the Python \texttt{sklearn.mixture.GMM} module \cite{sklearn}.
+In order to achieve this, we compare the Bayesian Information Criterion \citep{bicref} for a grid of Gaussian clusters ($N_\mathrm{G-C}<20$) and covariance restriction methods for each variable (Figure \ref{bic}).
+To do this we used the Python \texttt{sklearn.mixture.GMM} module \citep{sklearn}.
 We selected the models which produced the lowest BIC value, and plotted the GMM-classified, PCA-projected climate data in Figure \ref{lde}.
@@
 \includegraphics[scale=0.5]{bicb.pdf}
-\caption{Bayesian Information Criterion (BIC) results for Gaussian Mixture Models using a different number of clusters and using different covariance restriction methods \cite{sklearn} applied on each measured variable dataset.}\label{bic}
+\caption{Bayesian Information Criterion (BIC) results for Gaussian Mixture Models using a different number of clusters and using different covariance restriction methods \citep{sklearn} applied on each measured variable dataset.}\label{bic}
 \end{center}
-\end{figure}
+\end{figure*}
 
-\begin{figure}
+\begin{figure*}
 \begin{center}
@@
 \end{center}
-\end{figure}
+\end{figure*}
 
-\begin{figure}
+\begin{figure*}
 \begin{center}
-\includegraphics[scale=0.5]{gmmh0.pdf}
-\includegraphics[scale=0.5]{gmmh1.pdf}
+\includegraphics[scale=0.5,trim={0 15 0 18},clip]{gmmh0.pdf}
+\includegraphics[scale=0.5,trim={0 15 0 18},clip]{gmmh1.pdf}
 \caption{Climate clusters (1-2 of 2) as predicted by the lowest-BIC Gaussian Mixture Model for relative humidity.
@@
 \end{center}
-\end{figure}
+\end{figure*}
 
-\begin{figure}
+\begin{figure*}
 \begin{center}
-\includegraphics[scale=0.5]{gmml0.pdf}
-\includegraphics[scale=0.5]{gmml1.pdf}
-\includegraphics[scale=0.5]{gmml2.pdf}
-\caption{Climate clusters (1-3 of 11) as predicted by the lowest-BIC Gaussian Mixture Model for precipitation.
+\includegraphics[scale=0.5,trim={0 17 0 18},clip]{gmml0.pdf}
+\includegraphics[scale=0.5,trim={0 17 0 18},clip]{gmml1.pdf}
+\includegraphics[scale=0.5,trim={0 17 0 18},clip]{gmml2.pdf}
+\includegraphics[scale=0.5,trim={0 17 0 18},clip]{gmml3.pdf}
+\caption{Climate clusters (1-4 of 11) as predicted by the lowest-BIC Gaussian Mixture Model for precipitation.
 Left column: Average monthly precipitation and standard deviation (error bars) for each climate cluster.
@@
 \end{center}
-\end{figure}
+\end{figure*}
 
-\begin{figure}
+\begin{figure*}
 \begin{center}
-\includegraphics[scale=0.5]{gmml3.pdf}
-\includegraphics[scale=0.5]{gmml4.pdf}
-\includegraphics[scale=0.5]{gmml5.pdf}
-\caption{Climate clusters  (4-6 of 11) as predicted by the lowest-BIC Gaussian Mixture Model for precipitation.
+\includegraphics[scale=0.5,trim={0 17 0 18},clip]{gmml4.pdf}
+\includegraphics[scale=0.5,trim={0 17 0 18},clip]{gmml5.pdf}
+\includegraphics[scale=0.5,trim={0 17 0 18},clip]{gmml6.pdf}
+\includegraphics[scale=0.5,trim={0 17 0 18},clip]{gmml7.pdf}
+\caption{Climate clusters  (5-8 of 11) as predicted by the lowest-BIC Gaussian Mixture Model for precipitation.
 Left column: Average monthly precipitation and standard deviation (error bars) for each climate cluster.
@@
 \end{center}
-\end{figure}
+\end{figure*}
 
-\begin{figure}
+\begin{figure*}
 \begin{center}
 
-\includegraphics[scale=0.5]{gmml6.pdf}
-\includegraphics[scale=0.5]{gmml7.pdf}
-\includegraphics[scale=0.5]{gmml8.pdf}
-\caption{Climate clusters (7-9 of 11) as predicted by the lowest-BIC Gaussian Mixture Model for precipitation.
+\includegraphics[scale=0.5,trim={0 17 0 18},clip]{gmml8.pdf}
+\includegraphics[scale=0.5,trim={0 17 0 18},clip]{gmml9.pdf}
+\includegraphics[scale=0.5,trim={0 17 0 18},clip]{gmml10.pdf}
+\caption{Climate clusters (9-11 of 11) as predicted by the lowest-BIC Gaussian Mixture Model for precipitation.
 Left column: Average monthly precipitation and standard deviation (error bars) for each climate cluster.
@@
 \end{center}
-\end{figure}
+\end{figure*}
 
-\begin{figure}
+\begin{figure*}
 \begin{center}
-\includegraphics[scale=0.5]{gmml9.pdf}
-\includegraphics[scale=0.5]{gmml10.pdf}
-\caption{Climate clusters (10-11 of 11) as predicted by the lowest-BIC Gaussian Mixture Model for precipitation.
-Left column: Average monthly precipitation and standard deviation (error bars) for each climate cluster.
-Right column: Geographical location and altitude of weather stations in each precipitation cluster.
-Continuation of Figure \ref{clustl3}.}\label{clustl4}
-\end{center}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-\includegraphics[scale=0.5]{gmmd0.pdf}
-\includegraphics[scale=0.5]{gmmd1.pdf}
-\includegraphics[scale=0.5]{gmmd2.pdf}
-\caption{Climate clusters (1-3 of 6) as predicted by the lowest-BIC Gaussian Mixture Model for rain days.
+\includegraphics[scale=0.5,trim={0 17 0 18},clip]{gmmd0.pdf}
+\includegraphics[scale=0.5,trim={0 17 0 18},clip]{gmmd1.pdf}
+\includegraphics[scale=0.5,trim={0 17 0 18},clip]{gmmd2.pdf}
+\includegraphics[scale=0.5,trim={0 17 0 18},clip]{gmmd3.pdf}
+\caption{Climate clusters (1-4 of 6) as predicted by the lowest-BIC Gaussian Mixture Model for rain days.
 Left column: Average monthly rain days and standard deviation (error bars) for each climate cluster.
@@
 \end{center}
-\end{figure}
+\end{figure*}
 
-\begin{figure}
+\begin{figure*}
 \begin{center}
-\includegraphics[scale=0.5]{gmmd3.pdf}
-\includegraphics[scale=0.5]{gmmd4.pdf}
-\includegraphics[scale=0.5]{gmmd5.pdf}
-\caption{Climate clusters (4-6 of 6) as predicted by the lowest-BIC Gaussian Mixture Model for rain days.
+\includegraphics[scale=0.5,trim={0 17 0 18},clip]{gmmd4.pdf}
+\includegraphics[scale=0.5,trim={0 17 0 18},clip]{gmmd5.pdf}
+\caption{Climate clusters (5-6 of 6) as predicted by the lowest-BIC Gaussian Mixture Model for rain days.
 Left column: Average monthly rain days and standard deviation (error bars) for each climate cluster.
@@
 \end{center}
-\end{figure}
+\end{figure*}
 
-\begin{figure}
+\begin{figure*}
 \begin{center}
-\includegraphics[scale=0.5]{gmmb0.pdf}
-\includegraphics[scale=0.5]{gmmb1.pdf}
-\includegraphics[scale=0.5]{gmmb2.pdf}
-\caption{Climate clusters (1-3 of 5) as predicted by the lowest-BIC Gaussian Mixture Model for sunshine duration.
+\includegraphics[scale=0.5,trim={0 17 0 18},clip]{gmmb0.pdf}
+\includegraphics[scale=0.5,trim={0 17 0 18},clip]{gmmb1.pdf}
+\includegraphics[scale=0.5,trim={0 17 0 18},clip]{gmmb2.pdf}
+\includegraphics[scale=0.5,trim={0 17 0 18},clip]{gmmb3.pdf}
+\caption{Climate clusters (1-4 of 5) as predicted by the lowest-BIC Gaussian Mixture Model for sunshine duration.
 Left column: Average monthly sunshine duration and standard deviation (error bars) for each climate cluster.
@@
 \end{center}
-\end{figure}
+\end{figure*}
 
-\begin{figure}
+\begin{figure*}
 \begin{center}
-\includegraphics[scale=0.5]{gmmb3.pdf}
-\includegraphics[scale=0.5]{gmmb4.pdf}
-\caption{Climate clusters (4-5 of 5) as predicted by the lowest-BIC Gaussian Mixture Model for sunshine duration.
+
+\includegraphics[scale=0.5,trim={0 17 0 18},clip]{gmmb4.pdf}
+\caption{Climate clusters (5 of 5) as predicted by the lowest-BIC Gaussian Mixture Model for sunshine duration.
 Left column: Average monthly sunshine duration and standard deviation (error bars) for each climate cluster.
@@
 \end{center}
-\end{figure}
+\end{figure*}
 
@@
 }
-\begin{indented}
-\item[]\begin{tabular}{@{}lclll}
-\br
-Variable&Symbol&Lowest BIC&Lowest BIC&Selected Clusters\\
-&&No. of Clusters&Cov. Method &\\
-\mr
-Relative Humidity&$H$&2&Full&2\\
+\begin{tabular}{@{}lclll}
+\midrule
+Variable&Symbol& No. of&Covariance&Selected \\
+&&clusters&Method &Clusters\\
+\midrule
+Relative &$H$&2&Full&2\\
+Humidity&&&&\\
+\midrule
 Precipitation &$R$&11&Full&2,3,7,11\\
+\midrule
 Rain Days&$D$&6&Full&1,4,6\\
-Sunshine Duration&$S$&5&Tied&1,4,5\\
-\br
+\midrule
+Sunshine &$S$&5&Tied&1,4,5\\
+Duration&&&&\\
+\bottomrule
 \end{tabular}
-\end{indented}
 \end{table}
@@
 \end{itemize}
-
-Thus, the probability that a station $j=(0,1,...,N_\mathrm{stations})$ of type $T_i$, meets all four criteria ($C$) and appears in our shortlist ($A$) given that it is located at an elevation $h$, can be written as $P_j=P(C\cap A\cap T_i\ |\ h)$.
+Thus, the probability that a station $j=\{0,1,...,N_\mathrm{shortlist}\}$ of type $T_i$, meets all four criteria ($C$) and appears in our shortlist ($A$) given that it is located at an elevation $h$, can be written as $P_j=P(C\cap A\cap T_i\ |\ h)$.
 We cannot  calculate exactly this probability, as it depends on unknown unknowns, i.e.
-whether a station that did not measure a given variable in reality belongs to one of our preferred climatological clusters for that variable (Table \ref{tabclu}).
+on whether a station that did not measure a given variable in reality belongs to one of our preferred climatological clusters for that variable (Table \ref{tabclu}).
 Thus we will use Bayes' theorem to help estimate this probability using simpler conditional probabilities.
@@
 However, we can rewrite this probability as,
-\begin{eqnarray}
-P(C\ |\ A\cap T_i)&=P(R\cap D\cap H\cap S\ |\ A\cap T_i)\ , \\
-\label{pclanti}
-&=\frac{P(R\cap D\cap H\cap S\cap A\cap T_i)}{P(A\cap T_i)}\ . 
-\end{eqnarray}
+\begin{equation}\label{pclanti}
+P(C\ |\ A\cap T_i)=\frac{P(C\cap A\cap T_i)}{P(A\cap T_i)}\  .
+\end{equation}
 In order to estimate $P(C\ |\ A\cap T_i)$ we assume that the probability of a station meeting our criteria for a given number of variables is independent of the station type.
@@
 \end{equation}
-If this station of type $T_1$ is in our shortlist ($A$), the $R$ and $D$ criteria are met. Thus,
+If this station of type $T_1$ is in our shortlist ($A$), the $R$ and $D$ criteria are already met, so $A\cap R \cap D=A$. Thus,
 \begin{equation}
-P(R\cap D\cap H\cap S\cap A\cap T_1)=P( H\cap S\ |\ A\cap T_1)\ .
+P(C\cap A\cap T_1)=P( H\cap S\ |\ A\cap T_1)\ .
 \end{equation}
@@
 \end{equation}
-The last term in the previous expression is unknown, but we can approximate it assuming, as mentioned above, that it is independent of the station type. Therefore, 
+The last term in the previous expression is unknown, but we can approximate it by assuming that it is independent of the station type. Therefore, 
 \begin{equation}
-P(C\ |\ A\cap T_)\simeq P(H\cap S)\ .
+P(C\ |\ A\cap T_1)\simeq P(H\cap S)\ .
 \end{equation}
@@
 
-\begin{figure}
+\begin{figure*}
 \begin{center}
-\includegraphics[scale=0.55]{outliers.pdf}
+\includegraphics[scale=0.55,trim={0 17 0 18},clip]{outliers.pdf}
 \caption{Map of stations in our final shortlist but outside of the Colombian \emph{altiplano} region highlighted in Figure \ref{shortlist}.
@@
 
-\begin{figure}
+\begin{figure*}
 \begin{center}
-\includegraphics[scale=0.65]{unif.pdf}
+\includegraphics[scale=0.7,trim={0 0 50 0},clip]{unif.pdf}
 \caption{Geographical clustering results  for a lowest-BIC spherical covariance Gaussian Mixture Model of the stations located inside the red circle in Figure \ref{shortlist}b, indicating the predicted variance regions for each cluster (red circles), and plotted along rejected stations in our original dataset (black points).
@@
 
-We compared our results to the work of  \citeasnoun{suen2014} and \citeasnoun{suen2016}, where they obtained global precipitable water vapor maps for 2012 using data from the MODIS instrument on board the Aqua and Terra satellites.
-The lowest mean precipitable water vapor ($\sim15$ mm) regions in the \citeasnoun{suen2014} map for Colombia are near the Ubat\'e, Tunja, Valle del Sol, and Pisba regions.
-The \citeasnoun{suen2016} global median precipitable water vapor map has an upper bound of 10 mm for stations located in the clusters corresponding to those same regions.
+We compared our results to the work of  \citet{suen2014} and \citet{suen2016}, where they obtained global precipitable water vapor maps for 2012 using data from the MODIS instrument on board the Aqua and Terra satellites.
+The lowest mean precipitable water vapor ($\sim15$ mm) regions in the \citet{suen2014} map for Colombia are near the Ubat\'e, Tunja, Valle del Sol, and Pisba regions.
+The \citet{suen2016} global median precipitable water vapor map has an upper bound of 10 mm for stations located in the clusters corresponding to those same regions.
 Unfortunately, the resolution of the maps is not fine enough to pinpoint a site location at a precision higher than $\sim20$ km.
@@
 
-\begin{table}
-\caption{\label{tabfin}Stations in two regions of interest (RI) showing a high-altitude daytime dry, clear-sky climate, organized by location, Bayesian probability quality measure ($Q_j$), distance to cluster center ($r_c$), mean precipitation  ($\bar{R}$), mean rain days ($\bar{D}$), mean relative humidity  ($\bar{H}$), mean sunshine duration  ($\bar{S}$), and altitude ($h$).
-Detailed table (including all clusters) available at \texttt{https://github.com/saint-germain/ideam} \\}
-%https://tex.stackexchange.com/questions/98388/how-to-make-table-with-rotated-table-headers-in-latex
+\begin{table*}
+\caption{\label{tabfin}Stations in two regions of interest showing a high-altitude daytime dry, clear-sky climate, organized by location, Bayesian probability quality measure ($Q_j$), distance to cluster center ($r_c$), mean precipitation  ($\bar{R}$), mean rain days ($\bar{D}$), mean relative humidity  ($\bar{H}$), mean sunshine duration  ($\bar{S}$), and altitude ($h$).
+Detailed table (including all clusters) available at \texttt{https://github.com/saint-germain/ideam} .}
 
-\begin{tabular}{clrrrrrrr}
+\begin{tabular}{c|llcc|cccc|c}
 \toprule
-RI&                   Municipality  &  $Q_j$ &  $r_c$  &   $\bar{R}$ &   $\bar{D}$  &   $\bar{H}$  &  $\bar{S}$ &  $h$ \\
- &                   &    &    (km) &   (mm/mo) &  (d/mo) &   (\%) &   (h/d) &  (masl) \\
+Region of&   Station&                Municipality  &  $Q_j$ &  $r_c$  &   $\bar{R}$ &   $\bar{D}$  &   $\bar{H}$  &  $\bar{S}$ &  $h$ \\
+interest &    code   &            &    &    (km) &   (mm/mo) &  (d/mo) &   (\%) &   (h/d) &  (masl) \\
 \midrule
-            &                       Nemocon &            2.2 &                        28 &                        66 &                   ND &                        ND &                          ND &    2600 \\
-            &                          Susa &            4.6 &                        18 &                        83 &                   12 &                        ND &                          ND &    2600 \\
-            &                      Cucunuba &            4.7 &                         6 &                        57 &                    9 &                        ND &                          ND &    2620 \\
-            &                     Sutatausa &            4.8 &                         6 &                        57 &                   10 &                        ND &                          ND &    2700 \\
-            &                      Guacheta &            4.9 &                        16 &                        73 &                   10 &                        ND &                          ND &    2690 \\
-            &                   Lenguazaque &            5.2 &                        10 &                        63 &                   12 &                        ND &                          ND &    2650 \\
-            &                      Simijaca &            5.2 &                        23 &                        67 &                   11 &                        ND &                          ND &    2572 \\
-            &                        Suesca &            5.6 &                        20 &                        59 &                   10 &                        ND &                          ND &    2575 \\
+  &   21201410 &           Nemocon &  2.2 &            28 &       66 &       ND &    ND &  ND &    2600 \\
+  &   24011060 &              Susa &  4.6 &            18 &       83 &       12 &    ND &  ND &    2600 \\
+  &   24010140 &          Cucunuba &  4.7 &             6 &       57 &        9 &    ND &  ND &    2620 \\
+  &   24015210 &         Sutatausa &  4.8 &             6 &       57 &       10 &    ND &  ND &    2700 \\
+  &   24010170 &          Guacheta &  4.9 &            16 &       73 &       10 &    ND &  ND &    2690 \\
+  &   24010070 &       Lenguazaque &  5.2 &            10 &       63 &       12 &    ND &  ND &    2650 \\
+  &   24015130 &          Simijaca &  5.2 &            23 &       67 &       11 &    ND &  ND &    2572 \\
+  &   21201620 &            Suesca &  5.6 &            20 &       59 &       10 &    ND &  ND &    2575 \\
                                        \rotatebox{90}{\rlap{Valle de Ubat\'e}}
-            &              Carmen de Carupa &            6.1 &                        12 &                        54 &                   12 &                        ND &                          ND &    2970 \\
-            &                      Cucunuba &            6.1 &                         7 &                        51 &                    5 &                        ND &                          ND &    2562 \\
-            &                   Lenguazaque &            6.1 &                         9 &                        57 &                   10 &                        ND &                          ND &    2585 \\
-            &                          Susa &            6.1 &                        11 &                        60 &                    8 &                        ND &                          ND &    3130 \\
-            &   Ubate &            6.1 &                         4 &                        62 &                   13 &                        ND &                          ND &    2555 \\
-            &   Ubate &            6.1 &                         5 &                        62 &                   12 &                        ND &                          ND &    2610 \\
-            &                       Nemocon &            7.6 &                        20 &                        52 &                   10 &                        77 &                           4 &    2580 \\
+  &   24010610 &  Carmen de Carupa &  6.1 &            12 &       54 &       12 &    ND &  ND &    2970 \\
+  &   24011080 &          Cucunuba &  6.1 &             7 &       51 &        5 &    ND &  ND &    2562 \\
+  &   24010280 &       Lenguazaque &  6.1 &             9 &       57 &       10 &    ND &  ND &    2585 \\
+  &   24010440 &              Susa &  6.1 &            11 &       60 &        8 &    ND &  ND &    3130 \\
+  &   24011090 &             Ubate &  6.1 &             4 &       62 &       13 &    ND &  ND &    2555 \\
+  &   24015110 &             Ubate &  6.1 &             5 &       62 &       12 &    ND &  ND &    2610 \\
+  &   21205400 &           Nemocon &  7.6 &            20 &       52 &       10 &    77 &   4 &    2580 \\
       \midrule
-            &                       Duitama &            4.0 &                        17 &                        69 &                   10 &                        ND &                          ND &    2590 \\
-            &                         Paipa &            4.0 &                        12 &                        74 &                    9 &                        ND &                          ND &    2900 \\
-            &                         Nobsa &            5.0 &                         6 &                        66 &                   10 &                        ND &                          ND &    2500 \\
-            &         Santa Rosa &            5.0 &                        10 &                        70 &                    9 &                        ND &                          ND &    2500 \\
-            &                      Sogamoso &            5.0 &                         4 &                        59 &                   10 &                        ND &                          ND &    2500 \\
-            &                      Sogamoso &            5.0 &                         7 &                        60 &                   12 &                        ND &                          ND &    2500 \\
-            &                      Tibasosa &            5.0 &                         5 &                        62 &                    7 &                        ND &                          ND &    2500 \\
+  &   24030760 &           Duitama &  4.0 &            17 &       69 &       10 &    ND &  ND &    2590 \\
+  &   24030510 &             Paipa &  4.0 &            12 &       74 &        9 &    ND &  ND &    2900 \\
+  &   24030790 &             Nobsa &  5.0 &             6 &       66 &       10 &    ND &  ND &    2500 \\
+  &   24031040 &        Santa Rosa &  5.0 &            10 &       70 &        9 &    ND &  ND &    2500 \\
+  &   24035140 &          Sogamoso &  5.0 &             4 &       59 &       10 &    ND &  ND &    2500 \\
+  &   24030940 &          Sogamoso &  5.0 &             7 &       60 &       12 &    ND &  ND &    2500 \\
+  &   24030410 &          Tibasosa &  5.0 &             5 &       62 &        7 &    ND &  ND &    2500 \\
                                       \rotatebox{90}{\rlap{~Valle del Sol}}
-            &                           Iza &            5.6 &                        12 &                        52 &                    9 &                        ND &                          ND &    2470 \\
-            &                    Firavitoba &            6.1 &                         6 &                        57 &                    9 &                        ND &                          ND &    2486 \\
-            &                        Mongui &            6.1 &                        12 &                        65 &                    9 &                        ND &                          ND &    2970 \\
-            &                      Sogamoso &            6.1 &                        11 &                        65 &                   10 &                        ND &                          ND &    3225 \\
-            &                      Sogamoso &            6.7 &                         5 &                        61 &                   12 &                        75 &                           5 &    2500 \\
-            &                         Nobsa &           10.0 &                        10 &                        68 &                   13 &                        74 &                           4 &    2530 \\
+  &   24030410 &               Iza &  5.6 &            12 &       52 &        9 &    ND &  ND &    2470 \\
+  &   24030540 &        Firavitoba &  6.1 &             6 &       57 &        9 &    ND &  ND &    2486 \\
+  &   24030190 &            Mongui &  6.1 &            12 &       65 &        9 &    ND &  ND &    2970 \\
+  &   24030760 &          Sogamoso &  6.1 &            11 &       65 &       10 &    ND &  ND &    3225 \\
+  &   24035340 &          Sogamoso &  6.7 &             5 &       61 &       12 &    75 &   5 &    2500 \\
+  &   24035150 &             Nobsa & 10.0 &            10 &       68 &       13 &    74 &   4 &    2530 \\
 
@@
 
-\end{table}
+\end{table*}
 
@@
 Most of the stations in this table are located at elevations above 2600 masl and below 3000 masl.
-We validated our results using satellite measurements of the upper bound for precipitable water vapor \cite{suen2014,suen2016} for those two regions, whis is reported to be $\sim15$ mm.
+We validated our results using satellite measurements of the upper bound for precipitable water vapor \citep{suen2014,suen2016} for those two regions, whis is reported to be $\sim15$ mm.
 This means that it is not implausible to find a site with an even lower water vapor column in one of the regions of interest identified here.
@@
 
-The Villa de Leyva region, despite having a dry, sunny climate, is located at a comparatively low altitude (2200 masl), which means that the precipitable water vapor column above this region is too high ($>20$ mm according to the \citeasnoun{suen2016} maps).\\
+The Villa de Leyva region, despite having a dry, sunny climate, is located at a comparatively low altitude (2200 masl), which means that the precipitable water vapor column above this region is too high ($>20$ mm according to the \citet{suen2016} maps).\\
 
 There is an additional issue with the Valle del Sol region.
-The high industrial activity in this region fills the air with particulate material \cite{sogphd,sogamoso}, which would nullify the suitability of this region.
+The high industrial activity in this region fills the air with particulate material \citep{sogphd,sogamoso}, which would nullify the suitability of this region.
 Besides obtaining seasonal high-resolution precipitable water vapor maps, in our next paper we will study the amount of particulate material in the reported regions of interest using MODIS (Aqua and Terra) satellite data.\\
@@
 
- \section*{References}
+% \section*{References}
+ \bibliographystyle{mnras}
  \bibliography{paperclimate}
-\bibliographystyle{jphysicsB}
 
+%\bibliographystyle{mnras}
+%\bibliography{example} % if your bibtex file is called example.bib
+
+% Alternatively you could enter them by hand, like this:
+% This method is tedious and prone to error if you have lots of references
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%%% APPENDICES %%%%%%%%%%%%%%%%%%%%%
+
+%\appendix
+
+%\section{Some extra material}
+
+%If you want to present additional material which would interrupt the flow of the main paper,
+%it can be placed in an Appendix which appears after the list of references.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+% Don't change these lines
+\bsp	% typesetting comment
+\label{lastpage}
 \end{document}
+
+% End of mnras_template.tex

### 2017-05-11 mnras/mnras_template.tex
mnras

@@
 \usepackage{nicefrac} 
-\usepackage{epstopdf}
 \usepackage{booktabs}
@@
 \begin{keywords}
-keyword1 -- keyword2 -- keyword3
+atmospheric effects -- methods: data analysis -- methods: statistical -- site testing
 \end{keywords}
@@
 
-In this paper we seek to find plausible locations for a high-mountain mm-wave astronomical observatory in the northern Andes of Colombia, by analizing 30 years of climate data from 2046 weather stations.
-By low-dimensionally embedding the data, we are able to group together and correlate climate behavior indicative of daytime dry, clear-sky weather conditions at elevations higher than 2000 masl.
+In this paper we find plausible locations for a high-mountain mm-wave astronomical observatory in the northern Andes of Colombia, by analizing 30 years of climate data from 2046 weather stations.
+By low-dimensionally embedding the data, we are able to group together and correlate climate behavior indicative of daytime dry, clear-sky weather conditions.
 A repository with all our code and the full dataset is available at \texttt{https://github.com/saint-germain/ideam}.\\
 
-From our shortlist of 79 stations that met our criteria, we selected 70 stations located  in the Cundinamarca-Boyac\'a \emph{altiplano} region of Colombia, which is a high-mountain basin with an average altitude of 2600 masl.
+From our shortlist of 79 stations at elevations higher than 2000 masl that met our criteria, we selected 70 stations located  in the Cundinamarca-Boyac\'a \emph{altiplano} region of Colombia, which is a high-mountain basin with an average altitude of 2600 masl.
 Weather stations in our shortlist but outside this region are too sparsely located to suggest a candidate region of interest.
@@
 Most of the stations in this table are located at elevations above 2600 masl and below 3000 masl.
-We validated our results using satellite measurements of the upper bound for precipitable water vapor \citep{suen2014,suen2016} for those two regions, whis is reported to be $\sim15$ mm.
+We validated our results using satellite measurements of the upper bound for mean precipitable water vapor \citep{suen2014,suen2016} for those two regions, which is reported to be $\sim15$ mm in $\sim20$ km resolution maps.
 This means that it is not implausible to find a site with an even lower water vapor column in one of the regions of interest identified here.

### 2017-05-16 mnras/mnras_template.tex
before submission

@@
 \usepackage{ae,aecompl}
+\pdfminorversion=5
 
@@
 % Keep the title short and informative.
-\title[Low Dimensional Embedding of Climate Data for Radio Astronomical Site Testing]{Low Dimensional Embedding of Climate Data for Radio Astronomical Site Testing in the Colombian Andes}
+\title[Embedding of Climate Data for Site Testing]{Low Dimensional Embedding of Climate Data for Radio Astronomical Site Testing in the Colombian Andes}
 
@@
 % If you need two or more lines of authors, add an extra line using \newauthor
-\author[G. Chaparro-Molano et al.]{Germ\'an Chaparro-Molano$^{1}$\thanks{E-mail: gchaparrom@ecci.edu.co}
-Oscar Leonardo Ram\'irez-Su\'arez$^{1}$
-Oscar Restrepo$^{1,2}$\newauthor
-and Alexander Mart\'inez$^{1,3}$\\
+\author[G. Chaparro-Molano et al.]{Germ\'an Chaparro-Molano$^{1}$\thanks{E-mail: gchaparrom@ecci.edu.co},
+Oscar Leonardo Ram\'irez-Su\'arez$^{1}$,
+Oscar Restrepo$^{1,2},$\newauthor
+and Alexander Mart\'inez$^{1,3}$\newauthor \\
 % List of institutions
@@
 \begin{abstract}
-This is a simple template for authors to write new MNRAS papers.
-The abstract should briefly describe the aims, methods, and main results of the paper.
-It should be a single paragraph not more than 250 words (200 words for Letters).
-No references should appear in the abstract.
+We set out to evaluate the potential of the Colombian Andes for millimeter-wave astronomical observations.
+Previous studies for astronomical site testing in this region have suggested that nighttime humidity and cloud cover conditions make most sites unsuitable for professional visible-light observations.
+Millimeter observations can be done during the day, but require that the precipitable water vapor column above a site stays below $\sim$10 mm.
+Due to a lack of direct radiometric or radiosonde measurements, we present a method for correlating climate data from weather stations to sites with a low precipitable water vapor column.
+We use unsupervised learning techniques to low-dimensionally embed climate data (precipitation, rain days, relative humidity, and sunshine duration) in order to group together stations with similar long-term climate behavior.
+The data were taken over a period of 30 years by 2046 weather stations across the Colombian territory.
+We find 6 regions with unusually dry, clear-sky conditions, ranging in elevations from 2200 to 3800 masl.
+We evaluate the suitability of each region using a quality index derived from a Bayesian probabilistic analysis of the station type and elevation distributions.
+Two of these regions show a high probability of having an exceptionally low precipitable water vapor column.
+We compared our results with global precipitable water vapor maps and find a plausible geographical correlation with regions with low water vapor columns ($\sim15$ mm) at an accuracy of $\sim20$ km.
+Our methods can be applied to similar datasets taken in other countries as a first step toward astronomical site evaluation.
+
 \end{abstract}
@@
 Another recent surge in interest in telecommunications THz technology has appeared recently due to the saturation of the ``classical radio window'' in telecommunications  \citep{newera}.
-For this reason, there has been development of atmospheric models for absorption of THz photons \citep{rosenkranz,lababs} and artifact removal models \citep{removal}.
-Near-equatorial countries face a challenge in using this frequency band for astronomical observations and telecommunications using terabit satellite links  \citep{suen2016} due to the presence of a tropical belt of dense water vapor which efficiently absorbs THz radiation \citep{tropicalbelt}.
+For this reason, atmospheric models for absorption of THz photons \citep{rosenkranz,lababs} and artifact removal models \citep{removal} have been developed.
+Countries near the Equator face a challenge in using this frequency band for astronomical observations and telecommunications using terabit satellite links  \citep{suen2016} due to the presence of a tropical belt of dense water vapor which efficiently absorbs THz radiation \citep{tropicalbelt}.
 In northern South America, previous studies of astronomical site testing in the visible range have shown that high nighttime humidity conditions make this region suitable only for educational observatories \citep{pinzon}.
-However, considering that millimeter/sub-millimeter observations need not be done during the night, unusually dry, clear-daytime-sky, high altitude regions in the northern Andes could be suitable candidates for a world-class millimeter-wave observatory.
+However, considering that millimeter/sub-millimeter observations need not be done during the night, unusually dry, clear-daytime-sky, high elevation regions in the northern Andes could be suitable candidates for a millimeter-wave observatory.
 \\
@@
 The estimation was done using the Python \texttt{scipy.stats.gaussian\_kde}  function \citep{scipy}.
-(b) Altitude histogram for all stations in the dataset.}\label{totmap}
+(b) Elevation histogram for all stations in the dataset.}\label{totmap}
 \end{center}
@@
 From the climatological clusters in Figures \ref{clusth}-\ref{clustb} we selected clusters that indicate a dry, sunny climate.
-Table \ref{tabclu} shows the selected number of clusters and covariance method for each variable along with our preferred clusters .
+Table \ref{tabclu} shows the lowest-BIC number of clusters and covariance method for each variable along with our selected clusters.
 From here on, if a station appears in one of our preferred clusters, we will say that it satisfies our criterion for that specific variable.\\
@@
 Left column: Average monthly relative humidity and standard deviation (error bars) for each climate cluster.
-Right column: Geographical location and altitude of weather stations in each relative humidity cluster.}\label{clusth}
+Right column: Geographical location and elevation of weather stations in each relative humidity cluster.}\label{clusth}
 \end{center}
@@
 Left column: Average monthly precipitation and standard deviation (error bars) for each climate cluster.
-Right column: Geographical location and altitude of weather stations in each precipitation cluster.}\label{clustl}
+Right column: Geographical location and elevation of weather stations in each precipitation cluster.}\label{clustl}
 \end{center}
@@
 Left column: Average monthly precipitation and standard deviation (error bars) for each climate cluster.
-Right column: Geographical location and altitude of weather stations in each precipitation cluster.
+Right column: Geographical location and elevation of weather stations in each precipitation cluster.
 Continuation of Figure \ref{clustl}.}\label{clustl2}
@@
 Left column: Average monthly precipitation and standard deviation (error bars) for each climate cluster.
-Right column: Geographical location and altitude of weather stations in each precipitation cluster.
+Right column: Geographical location and elevation of weather stations in each precipitation cluster.
 Continuation of Figure \ref{clustl2}.}\label{clustl3}
@@
 Left column: Average monthly rain days and standard deviation (error bars) for each climate cluster.
-Right column: Geographical location and altitude of weather stations in each rain days cluster.}\label{clustd}
+Right column: Geographical location and elevation of weather stations in each rain days cluster.}\label{clustd}
 \end{center}
@@
 Left column: Average monthly rain days and standard deviation (error bars) for each climate cluster.
-Right column: Geographical location and altitude of weather stations in each rain days cluster.
+Right column: Geographical location and elevation of weather stations in each rain days cluster.
 Continuation of Figure \ref{clustd}.}
@@
 Left column: Average monthly sunshine duration and standard deviation (error bars) for each climate cluster.
-Right column: Geographical location and altitude of weather stations in each sunshine duration cluster.}\label{clustb1}
+Right column: Geographical location and elevation of weather stations in each sunshine duration cluster.}\label{clustb1}
 \end{center}
@@
 Left column: Average monthly sunshine duration and standard deviation (error bars) for each climate cluster.
-Right column: Geographical location and altitude of weather stations in each sunshine duration cluster.
+Right column: Geographical location and elevation of weather stations in each sunshine duration cluster.
 Continuation of Figure \ref{clustb1}.}\label{clustb}
@@
 The GMM-predicted variance region west of Tunja is located in the historic Tunja Canton, and even though the ratio of stations not in our shortlist to stations in this predicted region is low (0.31), there are simply not enough humidity and sunshine data to make any strong conclusion about this region.
-However, the high altitude of some of these stations is tantalizing, which is why we will keep this as a region of interest for our next paper.
+However, the high elevation of some of these stations is tantalizing, which is why we will keep this as a region of interest for our next paper.
 
@@
 \begin{table*}
-\caption{\label{tabfin}Stations in two regions of interest showing a high-altitude daytime dry, clear-sky climate, organized by location, Bayesian probability quality measure ($Q_j$), distance to cluster center ($r_c$), mean precipitation  ($\bar{R}$), mean rain days ($\bar{D}$), mean relative humidity  ($\bar{H}$), mean sunshine duration  ($\bar{S}$), and altitude ($h$).
+\caption{\label{tabfin}Stations in two regions of interest showing a high-elevation daytime dry, clear-sky climate, organized by location, Bayesian probability quality measure ($Q_j$), distance to cluster center ($r_c$), mean precipitation  ($\bar{R}$), mean rain days ($\bar{D}$), mean relative humidity  ($\bar{H}$), mean sunshine duration  ($\bar{S}$), and elevation ($h$).
 Detailed table (including all clusters) available at \texttt{https://github.com/saint-germain/ideam} .}
@@
 
-From our shortlist of 79 stations at elevations higher than 2000 masl that met our criteria, we selected 70 stations located  in the Cundinamarca-Boyac\'a \emph{altiplano} region of Colombia, which is a high-mountain basin with an average altitude of 2600 masl.
+From our shortlist of 79 stations at elevations higher than 2000 masl that met our criteria, we selected 70 stations located  in the Cundinamarca-Boyac\'a \emph{altiplano} region of Colombia, which is a high-mountain basin with an average elevation of 2600 masl.
 Weather stations in our shortlist but outside this region are too sparsely located to suggest a candidate region of interest.
@@
 This means that it is not implausible to find a site with an even lower water vapor column in one of the regions of interest identified here.
-Some mountains nearby Valle de Ubat\'e and Valle del Sol  have altitudes of up to 3400 masl, and therefore warrant further measurements using radio sondes and radiometers.
+Some mountains nearby Valle de Ubat\'e and Valle del Sol  have elevations of up to 3400 masl, and therefore warrant further measurements using radio sondes and radiometers.
 \\
 
-The Villa de Leyva region, despite having a dry, sunny climate, is located at a comparatively low altitude (2200 masl), which means that the precipitable water vapor column above this region is too high ($>20$ mm according to the \citet{suen2016} maps).\\
+The Villa de Leyva region, despite having a dry, sunny climate, is located at a comparatively low elevation (2200 masl), which means that the precipitable water vapor column above this region is too high ($>20$ mm according to the \citet{suen2016} maps).\\
 

### 2017-06-02 pasp/paperclimate.tex
ultimas correciones oleo
matpart

@@
 \begin{indented}
-\item[]December 2016
+\item[]
 \end{indented}
@@
 \begin{abstract}
-This document describes the  preparation of an article using \LaTeXe\ and 
-\verb"iopart.cls" (the IOP Publishing \LaTeXe\ preprint class file).
-This class file is designed to help 
-authors produce preprints in a form suitable for submission to any of the
-journals listed in table~\ref{jlab1} on the next page.  You are not obliged to use this class file---we accept
-submissions using all common \LaTeX\ class and style files.  The \verb"iopart.cls"
-class file is supplied merely as a convenience for those authors who find it useful.
-This document gives both general advice that applies whatever class file you use, and specific advice
-that applies if you choose to use \verb"iopart.cls".
-
-We also accept submissions in Word format.  Instructions for Word submissions are available via the `Author Guidelines' link at \verb"http://authors.iop.org".
-
-If you have any queries about this document or any aspect of preparing your article for submission please contact us at the e-mail address given above.
+We set out to evaluate the potential of the Colombian Andes for millimeter-wave astronomical observations.
+Previous studies for astronomical site testing in this region have suggested that nighttime humidity and cloud cover conditions make most sites unsuitable for professional visible-light observations.
+Millimeter observations can be done during the day, but require that the precipitable water vapor column above a site stays below $\sim$10 mm.
+Due to a lack of direct radiometric or radiosonde measurements, we present a method for correlating climate data from weather stations to sites with a low precipitable water vapor column.
+We use unsupervised learning techniques to low-dimensionally embed climate data (precipitation, rain days, relative humidity, and sunshine duration) in order to group together stations with similar long-term climate behavior.
+The data were taken over a period of 30 years by 2046 weather stations across the Colombian territory.
+We find 6 regions with unusually dry, clear-sky conditions, ranging in elevations from 2200 to 3800 masl.
+We evaluate the suitability of each region using a quality index derived from a Bayesian probabilistic analysis of the station type and elevation distributions.
+Two of these regions show a high probability of having an exceptionally low precipitable water vapor column.
+We compared our results with global precipitable water vapor maps and find a plausible geographical correlation with regions with low water vapor columns ($\sim15$ mm) at an accuracy of $\sim20$ km.
+Our methods can be applied to similar datasets taken in other countries as a first step toward astronomical site evaluation.
 \end{abstract}
@@
 %\vspace{2pc}
-\noindent{\it Keywords}: Water Vapor, Atmospheric Effects, Site Testing, Radio Astronomy, Satellite Communication, \\
+\noindent{\it Keywords}: atmospheric effects -- methods: data analysis -- methods: statistical -- site testing \\
 %
@@
 
-The development of astronomical instrumentation technology in the 0.2-2 THz range has been rapidly growing in recent years.
+The development of astronomical instrumentation technology in the 0.1-2 THz range has been rapidly growing in recent years.
 Another recent surge in interest in telecommunications THz technology has appeared recently due to the saturation of the ``classical radio window'' in telecommunications  \cite{newera}.
-For this reason, there has been development of atmospheric models for absorption of THz photons \cite{rosenkranz,lababs} and artifact removal models \cite{removal}.
-Near-equatorial countries face a challenge in using this frequency band for astronomical observations and telecommunications using terabit satellite links  \cite{suen2016} due to the presence of a tropical belt of dense water vapor which efficiently absorbs THz radiation \cite{tropicalbelt}.
+For this reason, atmospheric models for absorption of THz photons \cite{rosenkranz,lababs} and artifact removal models \cite{removal} have been developed.
+Countries near the Equator face a challenge in using this frequency band for astronomical observations and telecommunications using terabit satellite links  \cite{suen2016} due to the presence of a tropical belt of dense water vapor which efficiently absorbs THz radiation \cite{tropicalbelt}.
 In northern South America, previous studies of astronomical site testing in the visible range have shown that high nighttime humidity conditions make this region suitable only for educational observatories \cite{pinzon}.
@@
 This means that not all stations registered all the meteorological variables relevant to this work.
-Thus, we classify stations according to which variables they measured (Table \ref{tabstype}).\\
+Thus, we classify stations according to which variables they measured (Table \ref{tabstype}).
+The data is public and can be found in the repository for this paper at \texttt{https://github.com/saint-germain/ideam} .\\
 
@@
 \br
-Variable&Symbol&Lowest BIC&Lowest BIC&Selected Clusters\\
-&&No. of Clusters&Cov. Method &\\
+Variable&Symbol& No. of&Covariance&Selected \\
+&&clusters&Method &Clusters\\
 \mr
@@
 
-Thus, the probability that a station $j=(0,1,...,N_\mathrm{stations})$ of type $T_i$, meets all four criteria ($C$) and appears in our shortlist ($A$) given that it is located at an elevation $h$, can be written as $P_j=P(C\cap A\cap T_i\ |\ h)$.
+Thus, the probability that a station $j=\{0,1,...,N_\mathrm{shortlist}\}$ of type $T_i$, meets all four criteria ($C$) and appears in our shortlist ($A$) given that it is located at an elevation $h$, can be written as $P_j=P(C\cap A\cap T_i\ |\ h)$.
 We cannot  calculate exactly this probability, as it depends on unknown unknowns, i.e.
-whether a station that did not measure a given variable in reality belongs to one of our preferred climatological clusters for that variable (Table \ref{tabclu}).
+on whether a station that did not measure a given variable in reality belongs to one of our preferred climatological clusters for that variable (Table \ref{tabclu}).
 Thus we will use Bayes' theorem to help estimate this probability using simpler conditional probabilities.
@@
 
- The probability $P(C\ |\ A\cap T_i)$ has to be estimated indirectly, as we cannot presume to know under which conditions a station can satisfy all our criteria.
-However, we can rewrite this probability as,
-\begin{eqnarray}
-P(C\ |\ A\cap T_i)&=P(R\cap D\cap H\cap S\ |\ A\cap T_i)\ , \\
-\label{pclanti}
-&=\frac{P(R\cap D\cap H\cap S\cap A\cap T_i)}{P(A\cap T_i)}\ . 
-\end{eqnarray}
+\begin{equation}\label{pclanti}
+P(C\ |\ A\cap T_i)=\frac{P(C\cap A\cap T_i)}{P(A\cap T_i)}\  .
+\end{equation}
 In order to estimate $P(C\ |\ A\cap T_i)$ we assume that the probability of a station meeting our criteria for a given number of variables is independent of the station type.
@@
 \begin{table}
-\caption{\label{tabfin}Stations in two regions of interest (RI) showing a high-altitude daytime dry, clear-sky climate, organized by location, Bayesian probability quality measure ($Q_j$), distance to cluster center ($r_c$), mean precipitation  ($\bar{R}$), mean rain days ($\bar{D}$), mean relative humidity  ($\bar{H}$), mean sunshine duration  ($\bar{S}$), and altitude ($h$).
-Detailed table (including all clusters) available at \texttt{https://github.com/saint-germain/ideam} \\}
-%https://tex.stackexchange.com/questions/98388/how-to-make-table-with-rotated-table-headers-in-latex
+\caption{\label{tabfin}Stations in two regions of interest showing a high-elevation daytime dry, clear-sky climate, organized by location, Bayesian probability quality measure ($Q_j$), distance to cluster center ($r_c$), mean precipitation  ($\bar{R}$), mean rain days ($\bar{D}$), mean relative humidity  ($\bar{H}$), mean sunshine duration  ($\bar{S}$), and elevation ($h$).
+Detailed table (including all clusters) available at \texttt{https://github.com/saint-germain/ideam} .}
 
-\begin{tabular}{clrrrrrrr}
+\begin{tabular}{cllccccccc}
 \toprule
-RI&                   Municipality  &  $Q_j$ &  $r_c$  &   $\bar{R}$ &   $\bar{D}$  &   $\bar{H}$  &  $\bar{S}$ &  $h$ \\
- &                   &    &    (km) &   (mm/mo) &  (d/mo) &   (\%) &   (h/d) &  (masl) \\
+Region of&   Station&                Municipality  &  $Q_j$ &  $r_c$  &   $\bar{R}$ &   $\bar{D}$  &   $\bar{H}$  &  $\bar{S}$ &  $h$ \\
+interest &    code   &            &    &    (km) &   (mm/mo) &  (d/mo) &   (\%) &   (h/d) &  (masl) \\
 \midrule
-            &                       Nemocon &            2.2 &                        28 &                        66 &                   ND &                        ND &                          ND &    2600 \\
-            &                          Susa &            4.6 &                        18 &                        83 &                   12 &                        ND &                          ND &    2600 \\
-            &                      Cucunuba &            4.7 &                         6 &                        57 &                    9 &                        ND &                          ND &    2620 \\
-            &                     Sutatausa &            4.8 &                         6 &                        57 &                   10 &                        ND &                          ND &    2700 \\
-            &                      Guacheta &            4.9 &                        16 &                        73 &                   10 &                        ND &                          ND &    2690 \\
-            &                   Lenguazaque &            5.2 &                        10 &                        63 &                   12 &                        ND &                          ND &    2650 \\
-            &                      Simijaca &            5.2 &                        23 &                        67 &                   11 &                        ND &                          ND &    2572 \\
-            &                        Suesca &            5.6 &                        20 &                        59 &                   10 &                        ND &                          ND &    2575 \\
+  &   21201410 &           Nemocon &  2.2 &            28 &       66 &       ND &    ND &  ND &    2600 \\
+  &   24011060 &              Susa &  4.6 &            18 &       83 &       12 &    ND &  ND &    2600 \\
+  &   24010140 &          Cucunuba &  4.7 &             6 &       57 &        9 &    ND &  ND &    2620 \\
+  &   24015210 &         Sutatausa &  4.8 &             6 &       57 &       10 &    ND &  ND &    2700 \\
+  &   24010170 &          Guacheta &  4.9 &            16 &       73 &       10 &    ND &  ND &    2690 \\
+  &   24010070 &       Lenguazaque &  5.2 &            10 &       63 &       12 &    ND &  ND &    2650 \\
+  &   24015130 &          Simijaca &  5.2 &            23 &       67 &       11 &    ND &  ND &    2572 \\
+  &   21201620 &            Suesca &  5.6 &            20 &       59 &       10 &    ND &  ND &    2575 \\
                                        \rotatebox{90}{\rlap{Valle de Ubat\'e}}
-            &              Carmen de Carupa &            6.1 &                        12 &                        54 &                   12 &                        ND &                          ND &    2970 \\
-            &                      Cucunuba &            6.1 &                         7 &                        51 &                    5 &                        ND &                          ND &    2562 \\
-            &                   Lenguazaque &            6.1 &                         9 &                        57 &                   10 &                        ND &                          ND &    2585 \\
-            &                          Susa &            6.1 &                        11 &                        60 &                    8 &                        ND &                          ND &    3130 \\
-            &   Ubate &            6.1 &                         4 &                        62 &                   13 &                        ND &                          ND &    2555 \\
-            &   Ubate &            6.1 &                         5 &                        62 &                   12 &                        ND &                          ND &    2610 \\
-            &                       Nemocon &            7.6 &                        20 &                        52 &                   10 &                        77 &                           4 &    2580 \\
+  &   24010610 &  Carmen de Carupa &  6.1 &            12 &       54 &       12 &    ND &  ND &    2970 \\
+  &   24011080 &          Cucunuba &  6.1 &             7 &       51 &        5 &    ND &  ND &    2562 \\
+  &   24010280 &       Lenguazaque &  6.1 &             9 &       57 &       10 &    ND &  ND &    2585 \\
+  &   24010440 &              Susa &  6.1 &            11 &       60 &        8 &    ND &  ND &    3130 \\
+  &   24011090 &             Ubate &  6.1 &             4 &       62 &       13 &    ND &  ND &    2555 \\
+  &   24015110 &             Ubate &  6.1 &             5 &       62 &       12 &    ND &  ND &    2610 \\
+  &   21205400 &           Nemocon &  7.6 &            20 &       52 &       10 &    77 &   4 &    2580 \\
       \midrule
-            &                       Duitama &            4.0 &                        17 &                        69 &                   10 &                        ND &                          ND &    2590 \\
-            &                         Paipa &            4.0 &                        12 &                        74 &                    9 &                        ND &                          ND &    2900 \\
-            &                         Nobsa &            5.0 &                         6 &                        66 &                   10 &                        ND &                          ND &    2500 \\
-            &         Santa Rosa &            5.0 &                        10 &                        70 &                    9 &                        ND &                          ND &    2500 \\
-            &                      Sogamoso &            5.0 &                         4 &                        59 &                   10 &                        ND &                          ND &    2500 \\
-            &                      Sogamoso &            5.0 &                         7 &                        60 &                   12 &                        ND &                          ND &    2500 \\
-            &                      Tibasosa &            5.0 &                         5 &                        62 &                    7 &                        ND &                          ND &    2500 \\
+  &   24030760 &           Duitama &  4.0 &            17 &       69 &       10 &    ND &  ND &    2590 \\
+  &   24030510 &             Paipa &  4.0 &            12 &       74 &        9 &    ND &  ND &    2900 \\
+  &   24030790 &             Nobsa &  5.0 &             6 &       66 &       10 &    ND &  ND &    2500 \\
+  &   24031040 &        Santa Rosa &  5.0 &            10 &       70 &        9 &    ND &  ND &    2500 \\
+  &   24035140 &          Sogamoso &  5.0 &             4 &       59 &       10 &    ND &  ND &    2500 \\
+  &   24030940 &          Sogamoso &  5.0 &             7 &       60 &       12 &    ND &  ND &    2500 \\
+  &   24030410 &          Tibasosa &  5.0 &             5 &       62 &        7 &    ND &  ND &    2500 \\
                                       \rotatebox{90}{\rlap{~Valle del Sol}}
-            &                           Iza &            5.6 &                        12 &                        52 &                    9 &                        ND &                          ND &    2470 \\
-            &                    Firavitoba &            6.1 &                         6 &                        57 &                    9 &                        ND &                          ND &    2486 \\
-            &                        Mongui &            6.1 &                        12 &                        65 &                    9 &                        ND &                          ND &    2970 \\
-            &                      Sogamoso &            6.1 &                        11 &                        65 &                   10 &                        ND &                          ND &    3225 \\
-            &                      Sogamoso &            6.7 &                         5 &                        61 &                   12 &                        75 &                           5 &    2500 \\
-            &                         Nobsa &           10.0 &                        10 &                        68 &                   13 &                        74 &                           4 &    2530 \\
+  &   24030410 &               Iza &  5.6 &            12 &       52 &        9 &    ND &  ND &    2470 \\
+  &   24030540 &        Firavitoba &  6.1 &             6 &       57 &        9 &    ND &  ND &    2486 \\
+  &   24030190 &            Mongui &  6.1 &            12 &       65 &        9 &    ND &  ND &    2970 \\
+  &   24030760 &          Sogamoso &  6.1 &            11 &       65 &       10 &    ND &  ND &    3225 \\
+  &   24035340 &          Sogamoso &  6.7 &             5 &       61 &       12 &    75 &   5 &    2500 \\
+  &   24035150 &             Nobsa & 10.0 &            10 &       68 &       13 &    74 &   4 &    2530 \\
 
@@
 
-In this paper we seek to find plausible locations for a high-mountain mm-wave astronomical observatory in the northern Andes of Colombia, by analizing 30 years of climate data from 2046 weather stations.
-By low-dimensionally embedding the data, we are able to group together and correlate climate behavior indicative of daytime dry, clear-sky weather conditions at elevations higher than 2000 masl.
+In this paper we find plausible locations for a high-mountain mm-wave astronomical observatory in the northern Andes of Colombia, by analizing 30 years of climate data from 2046 weather stations.
+By low-dimensionally embedding the data, we are able to group together and correlate climate behavior indicative of daytime dry, clear-sky weather conditions.
 A repository with all our code and the full dataset is available at \texttt{https://github.com/saint-germain/ideam}.\\
 
-From our shortlist of 79 stations that met our criteria, we selected 70 stations located  in the Cundinamarca-Boyac\'a \emph{altiplano} region of Colombia, which is a high-mountain basin with an average altitude of 2600 masl.
+From our shortlist of 79 stations at elevations higher than 2000 masl that met our criteria, we selected 70 stations located  in the Cundinamarca-Boyac\'a \emph{altiplano} region of Colombia, which is a high-mountain basin with an average elevation of 2600 masl.
 Weather stations in our shortlist but outside this region are too sparsely located to suggest a candidate region of interest.
@@
 Most of the stations in this table are located at elevations above 2600 masl and below 3000 masl.
-We validated our results using satellite measurements of the upper bound for precipitable water vapor \cite{suen2014,suen2016} for those two regions, whis is reported to be $\sim15$ mm.
+We validated our results using satellite measurements of the upper bound for mean precipitable water vapor \cite{suen2014,suen2016} for those two regions, which is reported to be $\sim15$ mm in $\sim20$ km resolution maps.
 This means that it is not implausible to find a site with an even lower water vapor column in one of the regions of interest identified here.
-Some mountains nearby Valle de Ubat\'e and Valle del Sol  have altitudes of up to 3400 masl, and therefore warrant further measurements using radio sondes and radiometers.
+Some mountains nearby Valle de Ubat\'e and Valle del Sol  have elevations of up to 3400 masl, and therefore warrant further measurements using radio sondes and radiometers.
 \\
 
-The Villa de Leyva region, despite having a dry, sunny climate, is located at a comparatively low altitude (2200 masl), which means that the precipitable water vapor column above this region is too high ($>20$ mm according to the \citeasnoun{suen2016} maps).\\
+The Villa de Leyva region, despite having a dry, sunny climate, is located at a comparatively low elevation (2200 masl), which means that the precipitable water vapor column above this region is too high ($>20$ mm according to the \cite{suen2016} maps).\\
 

### 2017-07-06 mnras/mnras_template.tex

@@
 We summarize our results in Section 5, where we identify candidate regions of interest for a mm-wave astronomical observatory site.
-Finally, we discuss our main conclusions and future perspectives in Section 6 .
+Finally, we discuss our main conclusions and future perspectives in Section 6.
 

### 2017-07-11 opacities.py
better tau plot

@@ -1,0 +1,15 @@
+# CASA session
+from asap import opacity
+import numpy as np
+import matplotlib.pyplot as plt
+c=opacity.model(273.,62500.,0.3,3200.)
+d=opacity.model(273.,62500.,0.9,3200.)
+e=opacity.model(285.,72500.,0.9,2600.)
+freq=np.linspace(1e9,400e9,1000)
+opc=np.log10(100*np.array([c.get_opacities(i) for i in freq])*.01)
+ope=np.log10(100*np.array([e.get_opacities(i) for i in freq])*0.01)
+np.savetxt('e.txt',ope)
+np.savetxt('c.txt',opc)
+np.savetxt('freq.txt',freq)
+opd=np.log10(100*np.array([d.get_opacities(i) for i in freq])*.01)
+np.savetxt('d.txt',opd)

### 2017-07-11 pasp/paperclimate.tex
ref version

@@
 Two of these regions show a high probability of having an exceptionally low precipitable water vapor column.
-We compared our results with global precipitable water vapor maps and find a plausible geographical correlation with regions with low water vapor columns ($\sim15$ mm) at an accuracy of $\sim20$ km.
+We compared our results with global precipitable water vapor maps and find a plausible geographical correlation with regions with low water vapor columns ($\sim10$ mm) at an accuracy of $\sim20$ km.
 Our methods can be applied to similar datasets taken in other countries as a first step toward astronomical site evaluation.
@@
 The development of astronomical instrumentation technology in the 0.1-2 THz range has been rapidly growing in recent years.
-Another recent surge in interest in telecommunications THz technology has appeared recently due to the saturation of the ``classical radio window'' in telecommunications  \cite{newera}.
+Another recent surge in interest in technology at this frequency range has appeared recently due to the saturation of the ``classical radio window'' in telecommunications  \cite{newera}.
 For this reason, atmospheric models for absorption of THz photons \cite{rosenkranz,lababs} and artifact removal models \cite{removal} have been developed.
@@
 The main factor contributing to the atmospheric opacity is water vapor, which very efficiently absorbs light in the THz range \cite{pardocerni,pardocerni2,cont350} due to a continuum absorption spectrum formed by collisionally broadened absorption lines of water vapor in this frequency range  \cite{linecont,submmlines,turner2009}.
-In order to characterize a site according to its atmospheric transparency to THz radiation, it is necessary to retrieve the precipitable water vapor profile either via remote sensing techniques such as microwave radiometry \cite{radiometer,paine2000fourier,southpole2}, satellite measurements \cite{aqua,spacemols,suen2014,spaceradio}, radiosonde humidity measurements \cite{radiosonde,radiosonde2} or indirectly via models of in situ climatological measurements \cite{climatology} or GPS-delay studies \cite{gpsorig,gpsmap}.
+In order to characterize a site according to its atmospheric transparency to THz radiation, it is necessary to retrieve the precipitable water vapor (pwv) profile using remote sensing techniques such as microwave radiometry \cite{radiometer,paine2000fourier,southpole2}, satellite measurements \cite{aqua,spacemols,suen2014,spaceradio}, radiosonde humidity measurements \cite{radiosonde,radiosonde2}, or indirectly via models of in situ climatological measurements \cite{climatology} or GPS-delay studies \cite{gpsmet,gpsorig,gpsmap}.
 \\
@@
 This is why medium-to-long term monitoring of the atmosphere above a site is required before building an expensive mm/sub-mm wave radio telescope.
-However, even in best-case scenarios, earth-based telescopes will always be limited by atmospheric absorption of light specially at THz frequencies, where frequency windows of observations are few and in some cases, narrow \cite{arrlim,limits}.
+However, even in best-case scenarios, earth-based telescopes will always be limited by atmospheric absorption of light specially at THz frequencies, where frequency windows of observations are few and in some cases, narrow \cite{arrlim,limits}.\\
+
+We estimated the expected zenith opacity above the Colombian Andes (Figure \ref{opac}) at cm-mm wavelengths considering columns of 2, 10, and 20 mm of precipitable vater vapor and typical weather conditions.
+The elevation for the simulated sites in Figure \ref{opac} were chosen according to existing millimeter telescope sites (Pico Veleta, Plateau de Bure) \cite{mminf}.
+Thus, considering a best-case scenario (below 2 mm pwv), the potential observing frequency bands would be $70-115$ GHz, $120-170$ GHz, and $206-223$ GHz.
+The opacity models were calculated using the CASA \cite{casa} \texttt{asap.opacity.model} class, and deriving the pwv value from the 22 GHz K-band zenith opacity \cite{deuber,sanpedromartir}.
 \\
+
+\begin{figure}
+\begin{center}
+\includegraphics[scale=0.5]{opac.pdf}
+\caption{Atmospheric opacity models for the Colombian Andes for different precipitable water vapor (pwv) values corresponding to three scenarios: \textit{Red line} (20 mm pwv) a 2600 masl site with 90\% relative humidity at 13$^\circ$C, \textit{Blue line} (10 mm pwv) a 3400 masl site with 90\% relative humidity at 0$^\circ$C, \textit{Red line} (2 mm pwv) a 3400 masl site with 30\% relative humidity at 0$^\circ$C.}\label{opac}
+\end{center}
+\end{figure}
 
@@
 
-From the climatological clusters in Figures \ref{clusth}-\ref{clustb} we selected clusters that indicate a dry, sunny climate.
+From the climatological clusters in Figures \ref{clusth}-\ref{clustb} we selected clusters that indicate a clear-sky, potentially dry climate.
 Table \ref{tabclu} shows the lowest-BIC number of clusters and covariance method for each variable along with our selected clusters.
@@
 
-We wish to identify all stations for whom all four criteria (humidity, precipitation, rain days, sunshine duration) for identifying a region as having a stable sunny, dry climate can be met.
-However, since only a few stations measured all variables, it is not straightforward to accept or reject stations based on their satisfying a criterion, i.e.
-belonging to one of our preferred clusters.
-This requires setting up a measure of quality for individual stations depending on their location and the variables they measured.
-We will describe the selection process and probabilistic analysis that allows us to quantify a station's capability for indicating regions with an unusually dry, sunny climate.\\
-
-We start by creating a shortlist of stations that meet criteria for (and only for) the variables they measured.
-This means that if a station has precipitation and humidity data, it should belong to a humidity and precipitation cluster listed in Table \ref{tabclu} or else it is rejected.
+We wish to identify all stations for whom all four criteria (low humidity, precipitation, rain days, and high sunshine duration) for identifying a region as having a clear-sky, potentially dry climate can be met.
+However, since only 6\% of the stations measured all variables (type $T_8$ in Table \ref{tabstype}), it is not straightforward to accept or reject stations based on their belonging to one of our selected clusters.
+In other words, a station that meets less than all four criteria can still be considered in our analysis, but it will not be given the same level of importance as a station that meets all four criteria.
+This requires setting up a method to determine the quality of a site depending on its location and station type.
+Thus we selected stations that meet criteria for (and only for) the variables they measured.
+Thus, if a station measured precipitation and humidity (Type $T_3$), it had to belong to a precipitation and humidity cluster listed in Table \ref{tabclu} or else it was rejected.
 665 stations across Colombia showed such behavior.
-However, since we are interested in high-mountain regions, we included a criterion for elevation limiting our list to 119 stations located above 2000 masl.
-This shortlist was further reduced by requiring that the total average plus 1$\sigma$ for precipitation, rain days, and relative humidity is less than 4.3 mm/d, 17 d/mo, and 81\% respectively, and for sunshine duration, more than 4 h/d for any given station.
-This left us with 83 weather stations.
 \\
 
-However, we want to give a measure of the quality of the station vis-a-vis our criteria, i.e.
-a station that meets less than all four criteria should still be considered in our analysis, but it cannot be given the same importance as a station that meets all four criteria.
-In order to account for this, and controlling for the fact that stations are not distributed evenly in elevation (Figure \ref{totmap}b), we propose a quality measure based on a probabilistic analysis of the data.
+In order to estimate how much consideration we should give to a shortlisted station with an incomplete measurement set, we propose the use of a quality index based on a probabilistic analysis of the entire dataset, accounting for the non-uniform joint distribution of station types (Table \ref{tabstype}) and elevations (Figure \ref{totmap}b).
+Neglecting to include all stations in the calculation of this joint probability leads to very strong selection biases, assigning unfairly high quality indices to substandard stations.
+Therefore, even though we will later limit our selection to high-elevation sites in order to achieve lower pwv regions, we considered all the information provided by the full dataset in our probabilistic analysis.
 
@@
 
-We quantify the quality of a given site using the probability for the hypothesis that a station $j$ which appears in our shortlist meets all four criteria, given that it is located at a given elevation.
+We quantify the quality of a given site with the probability for the hypothesis that a station $j$ which appears in our shortlist meets all four criteria, given that it is located at a given elevation and is of a given type.
 We can write down this probability $P_j$ in terms of the following events:
@@
 
-Many of our 83 candidate stations, selected using the method described at the beginning of this section, were lacking in relative humidity and sunshine duration data.
-Given that for stations less than 5 km away the difference in relative humidity and sunshine duration data is less than 15\%, we extrapolated using data from nearby stations (at less than 4 km away and at an elevation difference of less than 100 m).
-The extrapolated data evidenced that 4 of the stations in the shortlist did not satisfy the relative humidity and sunshine duration criteria.
-The remaining 79 stations are mapped in Figure \ref{shortlist}a.\\
+As mentioned above, as we go higher in elevation, the water vapor column decreases.
+For this reason we included an additional criterion for elevation limiting our list to 119 stations located above 2000 masl.
+This list was further reduced by requiring that for any given station, the total average plus 1$\sigma$ for precipitation, rain days, and relative humidity is respectively less than 4.3 mm/d, 17 d/mo, and 81\%, and for sunshine duration, more than 4 h/d.
+This left us with a shortlist of 83 weather stations.\\
 
-We rejected 9 of the stations in Figure \ref{shortlist}a, i.e.
-those located outside of the region located within the latitude, longitude range $[(4.3,6.2)^\circ,(-72.4,-74.5)^\circ]$.
-Even though they could be indicative of unusually dry, sunny regions, there are simply not enough stations nearby to say anything conclusively, as evidenced in Figure \ref{outliers}.
-Adding to this, the only variable they measured was precipitation ($Q_j=2.2$), which leads us to reject them from our shortlist.
+Many of the stations in our dataset are lacking in relative humidity and sunshine duration data.
+In order to see if we could extrapolate using data from nearby stations, we estimated the difference between relative humidity and sunshine duration data for all stations in the entire dataset.
+For stations at a relative elevation of less than 100 m and located at less than 4 km away from each other, this difference amounts to less than 15\%.
+Thus, we extrapolated data for some of the 83 candidate stations in our shortlist, and were able to reject 4 stations which did not satisfy the relative humidity and sunshine duration criteria.
+The remaining 79 stations are plotted in Figure \ref{shortlist}a.\\
+
+We rejected 9 additional stations located outside of the region located within the latitude, longitude range $[(4.3,6.2)^\circ,(-72.4,-74.5)^\circ]$ (Figure \ref{shortlist}a) due to geographical sparsity.
+Even though they could be indicative of clear-sky, potentially dry regions, there are not enough stations nearby to make any strong conclusions regarding those regions, as shown in Figure \ref{outliers}.
+Adding to this, in all 9 cases the only variable they measured was precipitation ($Q_j=2.2$), which led us to reject them from our shortlist.
 \\
@@
 We selected clusters of climatologically similar stations across Colombia using an unsupervised learning low-dimensional embedding algorithm for each measured variable.
-From these clusters, 70 weather stations came up as potential candidates for identifying high-mountain regions (at an elevation greater that 2000 masl) with an unusually dry, sunny weather.
-Seasonally, the driest months of the year for all the candidate regions of interest are December-February and June-August (Figures \ref{clusth}-\ref{clustb}).
+From these clusters, 70 weather stations came up as potential candidates for identifying high-mountain regions (at an elevation greater that 2000 masl) with clear-sky, potentially dry weather.
+The driest months of the year for all the candidate regions of interest are December-February and June-August (Figures \ref{clusth}-\ref{clustb}).
+It should be noted that  humidity in the Colombian Andes can change rapidly during the course of the day \cite{pinzon}, and daytime relative humidity measurements $<50$\% are often reported in high-mountain plateau stations.
 
@@
 
-Figure \ref{unif} shows 6 candidate regions of interest, where weather stations indicating dry, sunny climate are regionally correlated within a $10-20$ km radius.
+Figure \ref{unif} shows 6 candidate regions of interest, where weather stations indicating clear-sky, potentially dry climate are regionally correlated within a $10-20$ km radius.
 In order to visualize the extent of those regions we plotted the predicted variance regions for each cluster along with other (rejected) stations in our sample in Figure \ref{unif}.
@@
 
-The first cluster is located at ($4.77^\circ$N, $74.18^\circ$W) and covers an radius of 15.3 km (Figure \ref{unif}, purple circle).
+The first cluster is located at ($4.77^\circ$N, $74.18^\circ$W) and covers an radius of 15.3 km (Figure \ref{unif}, magenta circle).
 This is a region located to the west of Bogot\'a at an elevation above 2600 masl, with a large variety of microclimates.
@@
 
-This cluster is located at ($5.30^\circ$N, $73.80^\circ$W) and covers a radius of 11.4 km (Figure \ref{unif}, blue circle).
+This cluster is located at ($5.30^\circ$N, $73.80^\circ$W) and covers a radius of 11.4 km (Figure \ref{unif}, cyan circle).
 This region, located to the NNE of Bogot\'a is a hilly region along a valley at an elevation above 2600 masl.
@@
 
-This cluster is located at ($5.59^\circ$N, $73.24^\circ$W) and covers a radius of 8.29 km (Figure \ref{unif}, red circle).
+This cluster is located at ($5.59^\circ$N, $73.24^\circ$W) and covers a radius of 8.29 km (Figure \ref{unif}, purple circle).
 The GMM-predicted variance region west of Tunja is located in the historic Tunja Canton, and even though the ratio of stations not in our shortlist to stations in this predicted region is low (0.31), there are simply not enough humidity and sunshine data to make any strong conclusion about this region.
@@
 
-This cluster is located at ($6.04^\circ$N, $72.71^\circ$W) and covers a radius of 10.52 km (Figure \ref{unif}, dark green circle).
+This cluster is located at ($6.04^\circ$N, $72.71^\circ$W) and covers a radius of 10.52 km (Figure \ref{unif}, teal circle).
 This is a \emph{p\'aramo} region, characterized by high-mountain tundra weather, with some mountains reaching an elevation of up to 3800 masl.
@@
 We compared our results to the work of  \citeasnoun{suen2014} and \citeasnoun{suen2016}, where they obtained global precipitable water vapor maps for 2012 using data from the MODIS instrument on board the Aqua and Terra satellites.
-The lowest mean precipitable water vapor ($\sim15$ mm) regions in the \citeasnoun{suen2014} map for Colombia are near the Ubat\'e, Tunja, Valle del Sol, and Pisba regions.
-The \citeasnoun{suen2016} global median precipitable water vapor map has an upper bound of 10 mm for stations located in the clusters corresponding to those same regions.
-Unfortunately, the resolution of the maps is not fine enough to pinpoint a site location at a precision higher than $\sim20$ km.
+Even though raw MODIS data has a potentially higher spatial resolution, the resolution of the processed maps is not fine enough to pinpoint a site location at a precision higher than $\sim20$ km.
+The area with the lowest mean precipitable water vapor that MODIS is able to measure ($\sim10-15$ mm) in the \citeasnoun{suen2014} map for Colombia spans the Tunja, Valle del Sol, Pisba, and part of the Valle de Ubat\'e regions.
+However, it also spans nearby regions with rejected stations, which means that low-resolution precipitable water vapor maps can indicate a wide, first order candidate region, but fail to account for highly local weather conditions, which are accounted for in our method.
+For this reason, in our next paper we will correlate the results presented here with existing GPS-delay water vapor measurements from the Colombian Global Navigation Satellite System GeoRED network (\texttt{https://geored.sgc.gov.co}) and seasonal, higher resolution MODIS data.
 
@@
 
-In this paper we find plausible locations for a high-mountain mm-wave astronomical observatory in the northern Andes of Colombia, by analizing 30 years of climate data from 2046 weather stations.
+In this paper we find plausible locations for a high-mountain cm- to mm-wave astronomical observatory in the northern Andes of Colombia, by analizing 30 years of climate data from 2046 weather stations.
 By low-dimensionally embedding the data, we are able to group together and correlate climate behavior indicative of daytime dry, clear-sky weather conditions.
@@
 Seasonally, there appear to be two times of the year (Dec-Feb and Jun-Aug) where the weather conditions indicate the best daytime dry, clear-sky conditions.
-These months could be the best for mm-wave astronomical observations to be carried out.\\
+These months could be the best for mm-wave astronomical observations to be carried out.
+The Villa de Leyva region, despite having a dry, sunny climate, is located at a comparatively low elevation (2200 masl), which means that the precipitable water vapor column above this region is too high ($>20$ mm according to the \cite{suen2016} maps).
+The La Sabana region is also not a good candidate due to a high rejection rate of stations within this region.\\
 
-We summarize the information about two regions of interest (Valle de Ubat\'e and Valle del Sol) in Table \ref{tabfin}, selected based on a Bayesian probabilistic quality index.
+We summarize the information about two regions of interest (Valle de Ubat\'e and Valle del Sol) in Table \ref{tabfin} with the highest Bayesian probabilistic indices, reflecting the quality of the data for stations in those regions.
 Most of the stations in this table are located at elevations above 2600 masl and below 3000 masl.
-We validated our results using satellite measurements of the upper bound for mean precipitable water vapor \cite{suen2014,suen2016} for those two regions, which is reported to be $\sim15$ mm in $\sim20$ km resolution maps.
+We validated our results using satellite measurements of the upper bound for mean precipitable water vapor \cite{suen2014,suen2016} for those two regions, which is reported to be $\sim10-15$ mm in $\sim20$ km resolution maps.
 This means that it is not implausible to find a site with an even lower water vapor column in one of the regions of interest identified here.
-Some mountains nearby Valle de Ubat\'e and Valle del Sol  have elevations of up to 3400 masl, and therefore warrant further measurements using radio sondes and radiometers.
-\\
-
-The Villa de Leyva region, despite having a dry, sunny climate, is located at a comparatively low elevation (2200 masl), which means that the precipitable water vapor column above this region is too high ($>20$ mm according to the \cite{suen2016} maps).\\
+Some mountains nearby Valle de Ubat\'e and Valle del Sol have elevations of up to 3400 masl, and therefore could warrant further measurements using GPS stations, radio sondes and radiometers.
+The same could be said for the Pisba and Tunja regions, but the low quality of the climate data (indicated by low Bayesian probabilistic indices) indicates that more data is needed.
+For this reason, we will correlate the results for these four regions with GPS-delay water vapor measurements in our following paper.\\
 
@@
 The high industrial activity in this region fills the air with particulate material \cite{sogphd,sogamoso}, which would nullify the suitability of this region.
-Besides obtaining seasonal high-resolution precipitable water vapor maps, in our next paper we will study the amount of particulate material in the reported regions of interest using MODIS (Aqua and Terra) satellite data.\\
+Besides obtaining seasonal, higher-resolution precipitable water vapor maps, in our next paper we will study the amount of particulate material in the reported regions of interest using MODIS (Aqua and Terra) satellite data.\\
 
-The methods used here can be extended and adapted to similar climatological datasets in other regions. This can be a preliminary step in radio astronomy site testing in developing countries.
+The methods used here can be extended and adapted to climatological datasets in other countries.
+Given a similar spatial and temporal coverage, our methods can provide a better picture of local climate than global satellite maps, and can help narrow down plausible locations for direct atmospheric opacity measurement campaigns.
+This can be a low-cost, preliminary step in radio astronomy site testing in developing countries.
 
@@
 This work was made possible by the Vicerrector\'ia de Investigaci\'on 01-2015 research grant from Universidad ECCI.
-We would also like to thank the Instituto de Hidrolog\'ia, Meteorolog\'ia y Estudios Ambientales (IDEAM) for providing us with the dataset.
+We would also like to thank the Instituto de Hidrolog\'ia, Meteorolog\'ia y Estudios Ambientales (IDEAM) for providing us with the dataset, and the anonymous referee for helping us convey our ideas more clearly.
 

### 2017-07-12 AOD_PM25/DatosTXT_HDF/extractor.py
cleanup

@@ -1,134 +1,0 @@
-# Debe correrlo en DatosTXT_HDF
-# Esta es la carpeta donde se corrio el script anterior (limpieza) sobre los hdf
-# Debe haber una carpeta paralela que se llama Datos_Insitu, aqui se debe guardar los .csv de los insitu
-# segun los nombres de las estaciones que estan en estaciones.csv (este .csv esta en el archivo anterior ..) con la fecha
-# por ejemplo kennedy2014.csv o minambiente2007.csv
-# el resultado es un archivo que se llama por ejemplo kennedy_data_2013.csv que esta en el ..
-
-
-import numpy as np
-
-import pandas as pd
-import os
-
-from datetime import datetime
-from datetime import timedelta
-estfile="../estaciones.csv"
-edf=pd.read_csv(estfile)
-isfilelist=[]
-for isfile in os.listdir("../Datos_Insitu/"):
-    if isfile.endswith(".csv") and not(isfile.startswith(".")):
-        isfilelist.append(isfile)
-for isfile in isfilelist:
-# Buscamos los datos in situ
-    filename="../Datos_Insitu/"+isfile
-    df=pd.read_csv(filename, names=["No","date","PM25"]) # con el formato alex
-    df=df.drop(df.index[[0]]) # quitamos el encabezado
-# Filtramos los datos invalidos
-    filter=(df.PM25=='Sin Data')
-    df1=df[~filter]
-    filter2=(df1.PM25=='InVld')
-    df1=df1[~filter2]
-    filter3=(df1.PM25=='Apagado')
-    df1=df1[~filter3]
-    filter4=(df1.PM25=='RS232')
-    df1=df1[~filter4]
-    filter5=(df1.PM25=='<Muestra')
-    df1=df1[~filter5]
-# Filtramos los NaN
-    filter6=np.isnan(pd.to_numeric(df1.PM25, errors='coerce'))
-    df1=df1[~filter6]
-# Encontramos el total de datos invalidos por filtro y los sumamos
-    invdata=filter.sum()+filter2.sum()+filter3.sum()+filter4.sum()+filter5.sum()
-    print "Hay %s datos invalidos de %s datos en %s" %(invdata,len(df),isfile)
-# Si todos los datos estan mal, rechazamos ese in situ para esa estacion
-    if(invdata==len(df)):
-        print isfile+" no sera incluido en el analisis."
-# Bloque de busqueda de archivos satelitales (una busqueda por cada in situ)
[…]

### 2017-07-12 AOD_PM25/DatosTXT_HDF/limpiardatos.py

@@ -1,19 +1,0 @@
-# Este script elimina pixeles muertos y filtra la ubicacion hasta un grado alrededor de Bogota
-# Para los datos de MODIS que ya han pasado de hdf a txt a traves de read_mod_aerosol_and_dump_ascii.py
-import pandas as pd
-import os
-filelist=[]
-for file in os.listdir("."):
-    if file.endswith(".txt"):
-        filelist.append(file)
-for filename in filelist:
-    df=pd.read_csv(filename)
-    filter=df.Optical_Depth_Land_And_Ocean!=-9999
-    df1=df[filter]
-    df1=df1[df1.Latitude < 5.21]
-    df1=df1[df1.Latitude > 4.21]
-    df1=df1[df1.Longitude > -74.6]
-    df1=df1[df1.Longitude < -73.6]
-    if len(df1) > 0:
-        df1.to_csv(filename[0:22]+'.filtered.csv')
-    print filename[0:22]+" tiene %s datos" % len(df1)

### 2017-07-12 AOD_PM25/createflist.sh

@@ -1,1 +1,0 @@
-ls DatosTXT_HDF/ | grep .hdf | awk '{print "DatosTXT_HDF/"$0}' > fileList.txt

### 2017-07-12 AOD_PM25/read_mod_aerosol_and_dump_ascii.py

@@ -1,1 +1,0 @@
-#!/usr/bin/python
'''
Module: read_mod_aerosol_and_dump_ascii.py
==========================================================================================
Disclaimer: The code is for demonstration purposes only. Users are responsible to check for accuracy and revise to fit their objective.

Author: Justin Roberts-Pierel, 2015 
Organization: NASA ARSET
Purpose: To save a MODIS HDF4 file (or series of files) in ASCII format, saving time, lat, lon, and other SDS dependent on file type

See the README associated with this module for more information.

 Este script toma la lista de archivos de fileList.txt generado con createflist.sh y automaticamente
 convierte los hdf en txt
 Es distinto al que se encuentra en la pagina de la NASA porque no pregunta nada, solo hace la conversion
 sobre los hdf que esten en fileList.txt
==========================================================================================
'''

#import necessary modules
from pyhdf import SD
import numpy as np
import time
import calendar
import sys

#This uses the file "fileList.txt", containing the list of files, in order to read the files
try:
	fileList=open('fileList.txt','r')
except:
	print('Did not find a text file containing file names (perhaps name does not match)')
	sys.exit()

#loops through all files listed in the text file
for FILE_NAME in fileList:
	FILE_NAME=FILE_NAME.strip()
#	user_input=raw_input('\nWould you like to process\n' + FILE_NAME + '\n\n(Y/N)')
        user_input="Y"
	if(user_input == 'N' or user_input == 'n'):
		continue
	else:
		if '3K' in FILE_NAME: #then this is a 3km MODIS file
			print('This is a 3km MODIS file. Saving... ')
			#saves all the SDS to be outputted to ASCII in a dictionary
			dataFields=dict([(1,'Optical_Depth_Land_And_Ocean'),(2,'Image_Optical_Depth_Land_And_Ocean'),(3,'Land_sea_Flag'),(4,'Land_Ocean_Quality_Flag')])
	 # The name of the SDS to read
		elif 'L2' in FILE_NAME: #Same as above but for 10km MODIS file
			print('This is a 10km MODIS file. Saving... ')
[…]

### 2017-07-12 yearlydata/clean.sh

@@ -1,4 +1,0 @@
-rm *.txt
-rm *.png
-rm *.csv
-rm -r estaciones

### 2017-07-12 yearlydata/create_plots.sh

@@ -1,6 +1,0 @@
-awk '{print "python plot.py " $1}' tags.txt | /bin/sh
-awk '{print "python plotmonth.py " $1}' tags.txt | /bin/sh
-awk '{print "python plotav.py " $1}' tags.txt | /bin/sh
-python plot_all.py
-python plotav_all.py
-python plotmonth_all.py

### 2017-07-12 yearlydata/extraccion.sh

@@ -1,55 +1,0 @@
-# Unzip and remove junk
-unzip IDEAM.txt.zip
-rm -r __MACOSX/
-# Get relevant data, cut to remove ancillary data, remove DOS end-of-line character, remove asterisks, write only T data to temp.txt using 12 columns (one p/ month), write to temp.txt
-# In OSX, use gcut instead of cut, as the OSX version of cut doesn’t like —option-delimiter
-grep -A 45 'MEDIOS  DIARIOS DE TEMPERATURA' IDEAM.txt | grep MEDIA | gcut -c22-25,31-34,40-43,49-52,58-61,67-70,76-79,85-88,94-97,103-106,112-115,121-124 --output-delimiter=' ,' | tr -d '\015' | sed 's/\*/ /g' > temp.txt
-# Delete leading spaces before temps > 0, write to temp
-sed 's/, \([0-9]\)/,\1/g' temp.txt > utemp.txt ; mv utemp.txt temp.txt
-# Add a comma before each line (to properly count nr. of fields <=12). At this point empty fields are not recognized as fields.
-sed 's/^/,/g' temp.txt > utemp.txt ; mv utemp.txt temp.txt
-# Count nr. of fields < 12 for each line in temp, write to fields.txt
-awk '{if (NF<12) print (12-NF)*6-1;else print 0 }' temp.txt > fields.txt
-# Get required number of spaces to even out (from the right) temp.txt
-awk '{print "printf " "'\''%"$1"s\\n'\''"}' fields.txt | /bin/sh > spaces.txt
-# Paste side-by-side temp.txt with required spaces from spaces.txt
-paste -d" " temp.txt spaces.txt > utemp.txt ; mv utemp.txt temp.txt
-# Replace spaces with comma-separated fields (to get 12 fields p/line)
-sed 's/      / ,    /g' temp.txt > utemp.txt ; mv utemp.txt temp.txt
-# Get year+station tag for each line in temp.txt
-grep -A2 'MEDIOS  DIARIOS DE TEMPERATURA' IDEAM.txt | grep ANO | awk '{print $7,$10}' > temptag.txt
-# Join yr+st tags with temp.txt, write to jtemp.txt
-paste temptag.txt temp.txt > jtemp.txt
-# Get unique list of station codes, write to tags.txt
-cut -c6-14 jtemp.txt | uniq > tags.txt
-# This block writes to a file for each station.
-# Create subdirectory
-mkdir estaciones
[…]

### 2017-07-12 yearlydata/humedad.sh

@@ -1,53 +1,0 @@
-# Unzip and remove junk
-unzip IDEAM.txt.zip
-rm -r __MACOSX/
-# Get relevant data, cut to remove ancillary data, remove DOS end-of-line character, remove asterisks, write only T data to temp.txt using 12 columns (one p/ month), write to temp.txt
-# CAMBIO: In OSX, use gcut instead of cut, as the OSX version of cut doesn’t like —option-delimiter
-grep -A 45 'VALORES MEDIOS  DIARIOS DE HUMEDAD' IDEAM.txt | grep MEDIA | gcut -c24-25,33-34,42-43,51-52,60-61,69-70,78-79,87-88,96-97,105-106,114-115,123-124 --output-delimiter=' ,' | tr -d '\015' | sed 's/\*/ /g' > temp.txt
-# Add a comma before each line (to properly count nr. of fields <=12). At this point empty fields are not recognized as fields.
-sed 's/^/,/g' temp.txt > utemp.txt ; mv utemp.txt temp.txt
-# CAMBIO: Count nr. of fields < 12 for each line in temp, write to fields.txt
-awk '{if (NF<12) print (12-NF)*4-1;else print 0 }' temp.txt > fields.txt
-# Get required number of spaces to even out (from the right) temp.txt
-awk '{print "printf " "'\''%"$1"s\\n'\''"}' fields.txt | /bin/sh > spaces.txt
-# Paste side-by-side temp.txt with required spaces from spaces.txt
-paste -d" " temp.txt spaces.txt > utemp.txt ; mv utemp.txt temp.txt
-# Replace spaces with comma-separated fields (to get 12 fields p/line)
-sed 's/    / ,  /g' temp.txt > utemp.txt ; mv utemp.txt temp.txt
-# Get year+station tag for each line in temp.txt
-grep -A2 'VALORES MEDIOS  DIARIOS DE HUMEDAD' IDEAM.txt | grep ANO | awk '{print $7,$10}' > temptag.txt
-# Join yr+st tags with temp.txt, write to jtemp.txt
-paste temptag.txt temp.txt > jtemp.txt
-# Get unique list of station codes, write to tags.txt
-cut -c6-14 jtemp.txt | uniq > tags.txt
-# This block writes to a file for each station.
-# Create subdirectory
-mkdir estaciones
-# Split jtemp into files for each station, write to (stationcode).txt
[…]

### 2017-07-12 yearlydata/hvsalt.py

@@ -1,22 +1,0 @@
-import subprocess
-import sys
-a=sys.argv[1]
-cmd = "grep -A 55 'MEDIOS  DIARIOS DE HUMEDAD' IDEAM.txt | grep -A55 'ANO  {0}' | grep -A2 'VALORES  ANUALES' | grep MEDIA | awk '{{print $2}}' > {0}mean.txt".format(a)
-p = subprocess.Popen(cmd , shell=True, stdout=subprocess.PIPE, stderr=subprocess.PIPE)
-p.wait()
-cmd = "grep -A 60 'MEDIOS  DIARIOS DE HUMEDAD' IDEAM.txt | grep -A4 'ANO  {0}' | grep ELEVACION | awk '{{print $2}}' > {0}alt.txt".format(a)
-p = subprocess.Popen(cmd , shell=True, stdout=subprocess.PIPE, stderr=subprocess.PIPE)
-p.wait()
-cmd = "paste -d',' {0}alt.txt {0}mean.txt > {0}tvsa.txt ; rm {0}alt.txt ;rm {0}mean.txt".format(a)
-p = subprocess.Popen(cmd , shell=True, stdout=subprocess.PIPE, stderr=subprocess.PIPE)
-p.wait()
-import pandas as pd
-import matplotlib as mpl
-import matplotlib.pyplot as plt
-mpl.rcParams['figure.figsize'] = 12,8
-df5 = pd.read_csv('%stvsa.txt' % a,index_col=False,names='ab')
-plt.scatter(df5.a,df5.b)
-plt.xlabel('Altitude')
-plt.ylabel('Rel. Humidity')
-plt.title('Mean Temperature for %s - All stations' % a)
-plt.savefig('%stvsalt' % a)

### 2017-07-12 yearlydata/plot.py

@@ -1,21 +1,0 @@
-import pandas as pd
-import numpy as np
-import matplotlib as mpl
-import matplotlib.pyplot as plt
-import sys
-a=sys.argv[1]
-print a
-mpl.rcParams['figure.figsize'] = 12, 8
-s='estaciones/%sf.txt' % a
-df = pd.read_csv(s)
-syr=float(df.columns[0])
-tot=int(len(df))
-date=np.arange(0,tot)/12.+syr
-plt.plot(date,df)
-plt.xlabel('Year')
-plt.ylabel('Rel. Humidity ($\%$)')
-sn='estaciones/%sn.txt' % a
-with open (sn, "r") as myfile:
-    a=myfile.read().replace('\n', '').replace('LA-','GUAJIRA-').replace('SAN-','S_ANDRES-').replace('NORTE-','N_SANT-').replace('HUIGUAJIRA-','HUILA-')
-plt.title("Monthly Mean Rel. Humidity: "+a)
-plt.savefig(a+"-historic-hum")

### 2017-07-12 yearlydata/plot_all.py

@@ -1,29 +1,0 @@
-#plot_all.py
-import pandas as pd
-import numpy as np
-import matplotlib as mpl
-import matplotlib.pyplot as plt
-mpl.rcParams['figure.figsize'] = 15, 8
-df1 = pd.read_csv('headers.txt')
-for i in range(0,len(df1.iloc[0])):
-    a=df1.iloc[0][i]
-    s='estaciones/%sf.txt' % a
-    df = pd.read_csv(s)
-    syr=float(df.columns[0])
-    tot=int(len(df))
-    date=np.arange(0,tot)/12.+syr
-    plt.plot(date,df,label=list(df1.columns.values)[i])
-    plt.legend(bbox_to_anchor=(0., 1.02, 1., .102), loc=0,
-           ncol=4, mode="expand", borderaxespad=1.)
-    lgd=plt.legend(bbox_to_anchor=(0., 1.02, 1., .102), loc=0,
-           ncol=4, mode="expand", borderaxespad=1.)
-    plt.xlabel('Year')
-    plt.ylabel('Rel. Humidity ($\%$)')
-    sn='estaciones/%sn.txt' % a
-    with open (sn, "r") as myfile:
-        a=myfile.read().replace('\n', '').replace('LA-','GUAJIRA-').replace('SAN-','S_ANDRES-').replace('NORTE-','N_SANT-').replace('HUIGUAJIRA-','HUILA-')  
-plt.title("Monthly Mean Rel. Humidity")
-x1,x2,y1,y2 = plt.axis()
-plt.axis([1975,2015,y1,y2])
-#plt.show()
-plt.savefig("all-monthly-mean-hum", bbox_extra_artists=(lgd,), bbox_inches='tight')

### 2017-07-12 yearlydata/plotav.py

@@ -1,23 +1,0 @@
-import pandas as pd
-import numpy as np
-import matplotlib as mpl
-import matplotlib.pyplot as plt
-import sys
-a=sys.argv[1]
-print a
-mpl.rcParams['figure.figsize'] = 12, 8
-s='estaciones/%s.txt' % a
-df = pd.read_csv(s, index_col=False,names='xabcdefghijkl')
-df=df.convert_objects(convert_numeric=True)
-s1='estaciones/%sf.txt' % a
-df1 = pd.read_csv(s1)
-syr=float(df1.columns[0])
-w=df.mean(axis=1)
-plt.plot(np.arange(syr,len(w)+syr),w)
-plt.xlabel('Year')
-plt.ylabel('Mean Rel. Humidity ($\%$)')
-sn='estaciones/%sn.txt' % a
-with open (sn, "r") as myfile:
-    a=myfile.read().replace('\n', '').replace('LA-','GUAJIRA-').replace('SAN-','S_ANDRES-').replace('NORTE-','N_SANT-').replace('HUIGUAJIRA-','HUILA-')
-plt.title("Yearly Mean Rel. Humidity: "+a)
-plt.savefig(a+"-historic_yr-hum")

### 2017-07-12 yearlydata/plotav_all.py

@@ -1,31 +1,0 @@
-import pandas as pd
-import numpy as np
-import matplotlib as mpl
-import matplotlib.pyplot as plt
-mpl.rcParams['figure.figsize'] = 15, 8
-df1 = pd.read_csv('headers.txt')
-
-for i in range(0,len(df1.iloc[0])):
-    a=df1.iloc[0][i]
-    s='estaciones/%s.txt' % a
-    df = pd.read_csv(s)
-    df=df.convert_objects(convert_numeric=True)
-    s1='estaciones/%sf.txt' % a
-    df2 = pd.read_csv(s1)
-    syr=float(df2.columns[0])
-    w=df.mean(axis=1)
-    plt.plot(np.arange(syr,len(w)+syr),w,label=list(df1.columns.values)[i])
-    plt.xlabel('Year')
-    plt.ylabel('Mean Rel. Humidity ($\%$C)')
-    plt.legend(bbox_to_anchor=(0., 1.02, 1., .102), loc=3,
-           ncol=4, mode="expand", borderaxespad=1.)
-    lgd=plt.legend(bbox_to_anchor=(0., 1.02, 1., .102), loc=3,
-           ncol=4, mode="expand", borderaxespad=1.)    
-    sn='estaciones/%sn.txt' % a
-    with open (sn, "r") as myfile:
-        a=myfile.read().replace('\n', '').replace('LA-','GUAJIRA-').replace('SAN-','S_ANDRES-').replace('NORTE-','N_SANT-').replace('HUIGUAJIRA-','HUILA-')
-plt.title("Yearly Mean Rel. Humidity")
-x1,x2,y1,y2 = plt.axis()
-plt.axis([1975,2015,y1,y2])
-plt.savefig("all-yearly-mean-hum", bbox_extra_artists=(lgd,), bbox_inches='tight')
-#plt.show()

### 2017-07-12 yearlydata/plotmonth.py

@@ -1,29 +1,0 @@
-import pandas as pd
-import numpy as np
-import matplotlib as mpl
-import matplotlib.pyplot as plt
-import sys
-a=sys.argv[1]
-print a
-mpl.rcParams['figure.figsize'] = 12, 8
-s='estaciones/%s.txt' % a
-df = pd.read_csv(s, index_col=False,names='xabcdefghijkl')
-df=df.convert_objects(convert_numeric=True)
-w=df.mean(axis=0)
-dev=df.std(axis=0)
-plt.errorbar(np.arange(1,13),w,yerr=dev,fmt='o', ecolor='r')
-plt.xlabel('Month')
-plt.ylabel('Mean Rel. Humidity ($\%$)')
-xlabels=['','Jan','Feb','Mar','Apr','May','Jun','Jul','Aug','Sep','Oct','Nov','Dec']
-plt.xticks(range(0,13),xlabels)
-s1='estaciones/%sf.txt' % a
-df1 = pd.read_csv(s1)
-syr=int(df1.columns[0])
-eyr=syr+len(df1)/12
-sn='estaciones/%sn.txt' % a
-with open (sn, "r") as myfile:
-    a=myfile.read().replace('\n', '').replace('LA-','GUAJIRA-').replace('SAN-','S_ANDRES-').replace('NORTE-','N_SANT-').replace('HUIGUAJIRA-','HUILA-')
-x1,x2,y1,y2 = plt.axis()
-plt.axis((0,13,y1,y2))
-plt.title("Average Rel. Humidity per Month (%s-%s): "% (syr,eyr-1)+a)
-plt.savefig(a+"-yrly_pattern-hum")

### 2017-07-12 yearlydata/plotmonth_all.py

@@ -1,38 +1,0 @@
-import pandas as pd
-import numpy as np
-import matplotlib as mpl
-import matplotlib.pyplot as plt
-mpl.rcParams['figure.figsize'] = 15,8
-df1 = pd.read_csv('headers.txt')
-
-for i in range(0,len(df1.iloc[0])):
-    a=df1.iloc[0][i]
-    s='estaciones/%s.txt' % a
-    df = pd.read_csv(s, index_col=False,names='xabcdefghijkl')
-    df=df.convert_objects(convert_numeric=True)
-    w=df.mean(axis=0)
-    dev=df.std(axis=0)
-    plt.errorbar(np.arange(1,13),w,yerr=dev,fmt='o',label=list(df1.columns.values)[i])
-#    plt.plot(np.arange(1,13),w,label=list(df1.columns.values)[i])
-    plt.xlabel('Month')
-    plt.ylabel('Mean Rel. Humidity ($\%$)')
-    xlabels=['','Jan','Feb','Mar','Apr','May','Jun','Jul','Aug','Sep','Oct','Nov','Dec']
-    plt.xticks(range(0,13),xlabels)
-    plt.legend(bbox_to_anchor=(0., 1.02, 1., .102), loc=3,
-           ncol=4, mode="expand", borderaxespad=1.)
-    lgd=plt.legend(bbox_to_anchor=(0., 1.02, 1., .102), loc=3,
-           ncol=4, mode="expand", borderaxespad=1.)
-    s1='estaciones/%sf.txt' % a
-    df2 = pd.read_csv(s1)
-    syr=int(df2.columns[0])
-    eyr=syr+len(df2)/12
-    sn='estaciones/%sn.txt' % a
-    with open (sn, "r") as myfile:
-        a=myfile.read().replace('\n', '').replace('LA-','GUAJIRA-').replace('SAN-','S_ANDRES-').replace('NORTE-','N_SANT-').replace('HUIGUAJIRA-','HUILA-')
-    print "junk",list(df1.columns.values)[i]
-    print dev
-plt.title("Average Rel. Humidity per Month (%s-%s) "% (1978,2015))
-x1,x2,y1,y2 = plt.axis()
-plt.axis((0,13,y1,y2))
-#plt.show()
-plt.savefig("all-yrly-pat-hum", bbox_extra_artists=(lgd,), bbox_inches='tight')

### 2017-07-12 yearlydata/puttogether.sh

@@ -1,26 +1,0 @@
-# This creates the pasted-together file.
-# Go to working directory (where the station-specific data is located)
-cd estaciones
-# Paste all (stationcode)f.txt files side by side. First line should be the starting year of measurements. Write to ltemp.txt
-cat ../tags.txt | cut -c1-8 | sed 's/$/f.txt/g' | xargs -n34 | awk '{print "paste -d \" \"  " $0}' | /bin/sh > ../ltemp.txt
-cd ..
-# Remove tabs from paste. Is this necessary?
-sed 's/\t/ /g' ltemp.txt > utemp.txt ; mv utemp.txt ltemp.txt
-# In this block we get the headers (starting year and station code) ready and formatted.
-# Get and separate starting years by comma. Write to ytemp.txt
-sed '1q;d' ltemp.txt | sed 's/ /,/g' > ytemp.txt
-# Add a leading comma to ytemp.txt
-sed 's/^/,/g' ytemp.txt > utemp.txt; mv utemp.txt ytemp.txt 
-# Remove first line from ltemp (line containing st. years, not comma separated).
-sed '1d' ltemp.txt > utemp.txt ; mv utemp.txt ltemp.txt 
-# Add first line from ytemp (line containing st. years, comma separated) to ltemp.txt
-cat ytemp.txt ltemp.txt > utemp.txt ; mv utemp.txt ltemp.txt 
-rm ytemp.txt
-# Get and separate station codes by comma, adding a leading comma and adding to ltemp.txt
-cat tags.txt | xargs -n34 | sed 's/ /,/g' | sed 's/^/,/' > utemp.txt; cat utemp.txt ltemp.txt > otemp.txt ; mv otemp.txt ltemp.txt ; rm utemp.txt
-# Get station location info (DPT+MUN), replace nasty characters. This is NOT efficient, as it has to re-grep the original file.
-awk '{print "grep -m 2 -A3 "$1" IDEAM.txt | grep -A2 DEPTO | awk '\''{print $8}'\'' | xargs -n2 "}' tags.txt | /bin/sh | sed 's/#/N/g' > utemp.txt
-# Separate station location info by commas, add leading comma.
-sed 's/ /-/g' utemp.txt | xargs -n34 | sed 's/ /,/g' | sed 's/^/,/' > otemp.txt ; mv otemp.txt utemp.txt
-# Add first line (location info) to ltemp.txt
-cat utemp.txt ltemp.txt > otemp.txt ; mv otemp.txt ltemp.txt ; rm utemp.txt

### 2017-07-12 yearlydata/tvsalt.py

@@ -1,22 +1,0 @@
-import subprocess
-import sys
-a=sys.argv[1]
-cmd = "grep -A 55 'MEDIOS  DIARIOS DE TEMPERATURA' IDEAM.txt | grep -A55 'ANO  {0}' | grep -A2 'VALORES  ANUALES' | grep MEDIA | awk '{{print $2}}' > {0}mean.txt".format(a)
-p = subprocess.Popen(cmd , shell=True, stdout=subprocess.PIPE, stderr=subprocess.PIPE)
-p.wait()
-cmd = "grep -A 60 'MEDIOS  DIARIOS DE TEMPERATURA' IDEAM.txt | grep -A4 'ANO  {0}' | grep ELEVACION | awk '{{print $2}}' > {0}alt.txt".format(a)
-p = subprocess.Popen(cmd , shell=True, stdout=subprocess.PIPE, stderr=subprocess.PIPE)
-p.wait()
-cmd = "paste -d',' {0}alt.txt {0}mean.txt > {0}tvsa.txt ; rm {0}alt.txt ;rm {0}mean.txt".format(a)
-p = subprocess.Popen(cmd , shell=True, stdout=subprocess.PIPE, stderr=subprocess.PIPE)
-p.wait()
-import pandas as pd
-import matplotlib as mpl
-import matplotlib.pyplot as plt
-mpl.rcParams['figure.figsize'] = 12,8
-df5 = pd.read_csv('%stvsa.txt' % a,index_col=False,names='ab')
-plt.scatter(df5.a,df5.b)
-plt.xlabel('Altitude')
-plt.ylabel('Temperature ($^\circ$C)')
-plt.title('Mean Temperature for %s - All stations' % a)
-plt.savefig('%stvsalt' % a)

### 2017-09-15 pasp/paperclimate.tex
before lores

@@
 \includegraphics[scale=0.5]{opac.pdf}
-\caption{Atmospheric opacity models for the Colombian Andes for different precipitable water vapor (pwv) values corresponding to three scenarios: \textit{Red line} (20 mm pwv) a 2600 masl site with 90\% relative humidity at 13$^\circ$C, \textit{Blue line} (10 mm pwv) a 3400 masl site with 90\% relative humidity at 0$^\circ$C, \textit{Red line} (2 mm pwv) a 3400 masl site with 30\% relative humidity at 0$^\circ$C.}\label{opac}
+\caption{Atmospheric opacity models for the Colombian Andes for different precipitable water vapor (pwv) values corresponding to three scenarios: \textit{Red line} (20 mm pwv) a 2600 masl site with 90\% relative humidity at 13$^\circ$C, \textit{Blue line} (10 mm pwv) a 3400 masl site with 90\% relative humidity at 0$^\circ$C, \textit{Green line} (2 mm pwv) a 3400 masl site with 30\% relative humidity at 0$^\circ$C.}\label{opac}
 \end{center}
